---
title: "GIS and biodiversity data management"
subtitle: "Workshop materials from the TAIEX Expert Mission, Pristina, 28–30 September 2026"
lang: en-GB
toc: false
# The page has no code worth showing, only the lookup that builds the cards.
code-tools: false
---

```{r}
#| label: setup

# The decks, their slots and their one-line summaries. publish_slides.R copies
# the files from the same list, so every link below has a file behind it.
decks <- read.csv("data/workshop_decks.csv", encoding = "UTF-8")

fmt_bytes <- function(x) {
  ifelse(x >= 1024^2, sprintf("%.1f MB", x / 1024^2),
         sprintf("%.0f kB", x / 1024))
}

# Day and month names are spelled out here rather than taken from format(),
# which follows the locale of whichever machine renders the site.
fmt_day <- function(date) {
  d <- as.POSIXlt(as.Date(date))
  wday <- c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday",
            "Friday", "Saturday")[d$wday + 1]
  sprintf("%s %d %s", wday, d$mday, month.name[d$mon + 1])
}
```

This page gathers the material from the five sessions Jonáš Gaigr of the Nature
Conservation Agency of the Czech Republic gave at the mission. Every
presentation is here as slides to view in the browser and as a PDF handout to
download, next to the GBIF data site that several of the sessions used as their
worked example.

## The mission

The Expert Mission on Geographic Information Systems (GIS) and biodiversity
data management was organised by TAIEX, the European Commission's Technical
Assistance and Information Exchange instrument, in co-operation with the
Ministry of Environment, Spatial Planning and Infrastructure (case ID ETT
IND/EXP 82606). For three days, experts from the Nature Conservation Agency of
the Czech Republic and the Ministry of Environment of the Slovak Republic worked
with the institutions that hold Kosovo's[\*]{.designation-mark} biodiversity
data.

Its aim was practical: to show how GIS supports environmental management,
biodiversity conservation and the implementation of the Birds and Habitats
Directives. That covered spatial analysis, habitat mapping, satellite and remote
sensing data, the monitoring of protected areas, assessing the conservation
status of habitats and species, and identifying Natura 2000 sites. Alongside it,
the mission set out to bring spatial data management in line with EU standards
and to strengthen co-operation between the institutions that collect and use
the data.

::: {.designation-note}
\* This designation is without prejudice to positions on status, and is in line
with UNSCR 1244/1999 and the ICJ Opinion on the Kosovo declaration of
independence.
:::

## The GBIF data site

A reproducible analysis of every GBIF occurrence record for Kosovo, built for
these sessions. It runs in the browser, needs nothing installed, and every
figure in it can be rebuilt from the open code.

::: {.signpost-grid}

::: {.download-card .featured}
#### Biodiversity of Kosovo

::: {.meta}
Report, maps and record explorer
:::

What GBIF holds for Kosovo and how far it can be trusted: records mapped by
density and municipality, the protected-area network and the evidence behind
it, species under the EU Nature Directives and the IUCN Red List, a filterable
map and table of every record of conservation interest, and open data
downloads.

::: {.actions}
[Open the report](report.qmd){.btn .btn-primary}
[Explore the records](report.qmd#explore-the-records){.btn .btn-outline-secondary}
[Download the data](report.qmd#download-the-data){.btn .btn-outline-secondary}
:::
:::

::: {.download-card}
#### Towards Natura 2000

::: {.meta}
The Ministry's questions, answered
:::

Evidence tiers, habitat mapping, sufficiency, the database, bird data and SPAs,
and a two-site pilot, computed from the same data as the report.

::: {.actions}
[Read the answers](natura2000.qmd){.btn .btn-primary}
:::
:::

::: {.download-card}
#### Source code

::: {.meta}
R and Quarto, openly licensed
:::

The workflow behind both pages and the per-site reports, to rerun for a new
extract or adapt for another territory.

::: {.actions}
[View on GitHub](https://github.com/jonasgaigr/kosovo-biodiversity-data-workshop-2026){.btn .btn-outline-secondary}
:::
:::

:::

## Presentations

Each deck is a single self-contained file: it opens in any modern browser and
keeps working offline once saved. The arrow keys move through it and Esc shows
every slide at once. The handout has one slide per page.

```{r}
#| label: decks
#| output: asis

card <- function(i) {
  stem <- file.path(decks$folder[i], decks$file[i])
  href <- paste0("slides/", decks$slug[i], "/", decks$file[i])
  sprintf(paste0(
    "<div class='download-card'><h4>%s</h4>",
    "<div class='meta'>%s&ndash;%s</div><p>%s</p><div class='actions'>",
    "<a class='btn btn-primary' href='%s.html'>View slides (%s)</a>",
    "<a class='btn btn-outline-secondary' href='%s.pdf' download>",
    "PDF handout (%s)</a></div></div>"),
    decks$title[i], decks$start[i], decks$end[i], decks$summary[i],
    href, fmt_bytes(file.size(paste0(stem, ".html"))),
    href, fmt_bytes(file.size(paste0(stem, ".pdf"))))
}

for (d in unique(decks$day)) {
  rows <- which(decks$day == d)
  cat(sprintf("\n### Day %d &middot; %s\n\n", d, fmt_day(decks$date[rows[1]])))
  cat("<div class='signpost-grid'>",
      vapply(rows, card, character(1)),
      "</div>\n", sep = "\n")
}
```

```{=html}
<script>
// The report was the site's front page until this one took its place, so
// links shared before then point at its sections here (/#explore-the-records).
// A fragment this page does not have is sent on to the report.
(function () {
  var id = decodeURIComponent(location.hash.slice(1));
  if (id && !document.getElementById(id)) {
    location.replace("report.html" + location.hash);
  }
})();
</script>
```
