---
title: "Biodiversity of Kosovo"
subtitle: "An analysis of GBIF occurrence data in support of conservation planning"
date: today
date-format: "D MMMM YYYY"
lang: en-GB
---

```{r}
#| label: setup

suppressPackageStartupMessages({
  library(dplyr)
  library(sf)
  library(DT)
  library(leaflet)
  library(leaflet.extras)
  library(leafem)
  library(crosstalk)
  library(htmltools)
})

source("R/functions.R")

meta_path <- "data/run_metadata.rds"
pipeline_has_run <- file.exists(meta_path)

if (pipeline_has_run) meta <- readRDS(meta_path)

# Human-readable file sizes for the download panel.
fmt_bytes <- function(x) {
  ifelse(is.na(x), "&mdash;",
    ifelse(x >= 1024^2, sprintf("%.1f MB", x / 1024^2),
      ifelse(x >= 1024, sprintf("%.0f kB", x / 1024), sprintf("%.0f B", x))))
}

# A consistently styled data table for the whole report.
report_table <- function(x, caption = NULL, page_length = 10,
                         buttons = TRUE, ...) {
  DT::datatable(
    x,
    rownames   = FALSE,
    caption    = caption,
    extensions = if (buttons) "Buttons" else character(),
    options = list(
      dom        = if (buttons) "Bfrtip" else "ftip",
      buttons    = if (buttons) list(
                     list(extend = "copy",  text = "Copy"),
                     list(extend = "csv",   text = "Download CSV"),
                     list(extend = "excel", text = "Download Excel")
                   ) else list(),
      pageLength = page_length,
      scrollX    = TRUE,
      autoWidth  = FALSE,
      language   = list(search = "Filter:", emptyTable = "No records available")
    ),
    class = "stripe hover compact order-column",
    ...
  )
}

# A coloured square used to carry a category into running text and tables,
# always beside the category's own label rather than instead of it.
swatch <- function(colour, label) {
  sprintf("<span class='swatch' style='background:%s'></span>%s", colour, label)
}
```

```{r}
#| label: guard
#| output: asis

if (!pipeline_has_run) {
  cat(
    "::: {.callout-important}\n",
    "## The data pipeline has not yet been run\n\n",
    "This report is generated from files produced by `pipeline.R`. ",
    "Set your GBIF credentials in `.Renviron`, then run:\n\n",
    "```bash\nRscript pipeline.R\n```\n\n",
    "and render the site again.\n",
    ":::\n",
    sep = ""
  )
  knitr::knit_exit()
}
```

```{r}
#| label: load-subsets

subset_files <- file.path("data_exports", paste0(meta$manifest$name, ".gpkg"))
names(subset_files) <- meta$manifest$name

read_subset <- function(path) {
  if (file.exists(path)) sf::st_read(path, quiet = TRUE) else NULL
}

subsets  <- lapply(subset_files, read_subset)
boundary <- if (file.exists("data/kosovo_boundary.gpkg")) {
  sf::st_read("data/kosovo_boundary.gpkg", quiet = TRUE)
} else NULL

municipalities <- if (file.exists("data/kosovo_municipalities.gpkg")) {
  sf::st_read("data/kosovo_municipalities.gpkg", quiet = TRUE)
} else NULL

overall <- subsets[["kosovo_overall_biodiversity"]]
overall_flat <- sf::st_drop_geometry(overall)
```

## Why this matters

Kosovo sits within the Balkan biodiversity hotspot, one of the most
species-rich regions in Europe. Evidence-based conservation — designating
protected areas, screening development proposals, and reporting against the EU
Nature Directives — depends on knowing **which species occur where**.

The Global Biodiversity Information Facility (GBIF) aggregates species
occurrence records published by museums, research institutions, monitoring
programmes and citizen-science platforms worldwide. This report brings the
GBIF holdings for the territory of Kosovo into a single, quality-controlled,
openly downloadable evidence base, and highlights the records that relate to
species protected under the EU Birds and Habitats Directives or assessed as
threatened on the IUCN Red List.

```{r}
#| label: key-figures
#| output: asis

m <- meta$manifest

kf <- function(value, label) {
  sprintf("<div class='keyfigure'><span class='value'>%s</span><span class='label'>%s</span></div>",
          value, label)
}

years <- suppressWarnings(as.integer(overall_flat$year))
years <- years[!is.na(years) & years > 1500]

n_threatened_species <- sum(
  meta$iucn_summary$species[meta$iucn_summary$category %in% iucn_threatened]
)

n_municipalities_with_records <- sum(meta$municipal_summary$records > 0)

cat("<div class='keyfigures'>",
    kf(fmt_int(meta$n_clean), "Quality-checked records"),
    kf(fmt_int(meta$n_species), "Species recorded"),
    kf(fmt_int(sum(m$records[m$name %in% c("kosovo_birds_annex_I",
                                           "kosovo_habitats_directive")])),
       "Directive-listed records"),
    kf(fmt_int(n_threatened_species), "IUCN-threatened species"),
    kf(sprintf("%d / %d", n_municipalities_with_records,
               nrow(meta$municipal_summary)), "Municipalities with records"),
    kf(if (length(years)) sprintf("%d&ndash;%d", min(years), max(years)) else "&mdash;",
       "Temporal coverage"),
    "</div>", sep = "\n")
```

## Data provenance and citation

The occurrence data were obtained through the GBIF asynchronous download
service, which mints a permanent Digital Object Identifier (DOI) for the exact
extract used here. Citing that DOI allows any reader to retrieve the identical
dataset, and passes credit back to the institutions that published the records.

```{r}
#| label: citation
#| output: asis

doi <- meta$download$doi
cat(
  "<div class='citation-box'>",
  sprintf("<strong>Dataset DOI:</strong> <a href='https://doi.org/%s'>https://doi.org/%s</a><br>", doi, doi),
  sprintf("<strong>GBIF download key:</strong> <code>%s</code><br>", meta$download$key),
  sprintf("<strong>Retrieved:</strong> %s<br>", fmt_date_en(as.Date(meta$run_date))),
  sprintf("<strong>Spatial filter:</strong> GADM boundary <code>%s</code> (Kosovo)<br>",
          meta$config$gadm_gid),
  sprintf("<strong>Contributing datasets:</strong> %d, from %d publishing institutions<br><br>",
          nrow(meta$datasets), dplyr::n_distinct(meta$datasets$publisher)),
  "<strong>Suggested citation</strong><br>",
  sprintf("<code>%s</code>", meta$download$citation %||% paste0("GBIF Occurrence Download https://doi.org/", doi)),
  "</div>", sep = "\n"
)
```

::: {.callout-note collapse="true"}
## A note on the spatial filter

Records were selected using the GADM administrative boundary for Kosovo
(GID `XKO`) rather than the ISO country code `XK`. The GADM boundary is a true
spatial filter, so it also captures records whose publisher left the country
field blank or assigned it to a neighbouring state — a common situation for
cross-border survey data. Queried against the GBIF occurrence API without any
further filters at the time of extraction, `gadmGid=XKO` returned 52,290
records against 48,947 for `country=XK`, about 7 per cent more.

Note that the GADM identifier for Kosovo is `XKO`. The similar-looking code
`XKX` is the World Bank / ISO alpha-3 style code and is **not** recognised by
GADM — `pred("gadm", "XKX")` returns zero records rather than an error, so a
pipeline using it would silently produce an empty dataset.
:::

## Summary statistics

Each row below is a distinct thematic extract of the cleaned dataset. All six
are available for download at the end of this report.

```{r}
#| label: summary-table

summary_tbl <- bind_rows(lapply(seq_len(nrow(meta$manifest)), function(i) {
  nm <- meta$manifest$name[i]
  summarise_subset(subsets[[nm]], meta$manifest$label[i])
}))

report_table(
  summary_tbl,
  caption = htmltools::tags$caption(
    style = "caption-side: bottom; text-align: left; padding-top: 8px;
             font-size: 0.85rem; color: #6b7880;",
    "Table 1. Cleaned GBIF occurrence records for Kosovo, by thematic subset. ",
    "‘Source datasets’ counts the distinct GBIF datasets contributing records."
  ),
  page_length = 6
) |>
  DT::formatCurrency(c("Records", "Species", "Families", "Source datasets"),
                     currency = "", interval = 3, mark = ",", digits = 0)
```

## Data quality control

Aggregated occurrence data carries well-known geographic artefacts: records
placed at the centre of a country, at a museum's own address, or at coordinates
of exactly zero. Every record was screened with `CoordinateCleaner` and only
records passing all tests were retained.

The screening finds nothing to remove, and that is the expected result rather
than a broken test. GBIF's own filters excluded records without coordinates and
records it had already flagged with a geospatial issue before the extract was
built, and the extract was selected by a true spatial filter, so the failure
modes these tests look for have no way into the dataset. The screening is kept
because it is the guarantee: it would catch them if they ever appeared.

```{r}
#| label: cleaning-table

report_table(
  meta$cleaning_report |> rename(`Records flagged` = Flagged),
  caption = htmltools::tags$caption(
    style = "caption-side: bottom; text-align: left; padding-top: 8px;
             font-size: 0.85rem; color: #6b7880;",
    sprintf(paste0("Table 2. Outcome of the coordinate screening. %s of %s ",
                   "records (%.1f%%) passed every test and form the basis of ",
                   "this report. A record may be flagged by more than one test."),
            fmt_int(meta$n_clean), fmt_int(meta$n_raw),
            100 * meta$n_clean / meta$n_raw)
  ),
  page_length = 10, buttons = FALSE
)
```

::: {.callout-note collapse="true"}
## Which tests were applied, which were left out, and the polygon they all depend on

Applied: zero coordinates, country-coordinate mismatch, GBIF headquarters,
known biodiversity institutions, and identical latitude/longitude values.

Deliberately **not** applied: the *capital city* and *country centroid* tests.
Both discard records falling within a fixed radius of a capital or of a
country's geometric centre. Kosovo is small enough that Pristina and the
national centroid both sit well inside the territory, so these tests would
remove a large number of entirely legitimate records — including most urban
biodiversity data. They can be enabled in `pipeline.R` by setting
`use_capitals_test = TRUE`.

The country-coordinate test needed two corrections, and both fail silently.

**It cannot see Kosovo by default.** Natural Earth, the reference dataset
`CoordinateCleaner` ships with, records Kosovo with no assigned ISO alpha-3
code. Run against that default the test flags **100 per cent** of Kosovo
records as mismatches and, with `value = "clean"`, returns an empty data frame
without warning. The pipeline therefore supplies a bespoke reference polygon
stamped with the code used in the occurrence data.

**That polygon then has to be the right one.** Built from Natural Earth's
1:50m outline — 72 vertices for the whole country, straying up to 4.7 km from
the true border — it flagged 1,296 records, 2.7 per cent of the dataset and
192 species. Every one was a genuine record near the border, discarded only
because the outline used to *test* them was not the outline used to *select*
them: this extract is chosen with `gadmGid=XKO`, so every record lies inside
the GADM polygon by construction. The reference is now GADM level 0 itself, at
1,210 vertices and exactly the geometry GBIF filtered on, and the test flags
nothing. It is kept in the battery because it becomes meaningful again the
moment the extract is defined differently — a `country=XK` download leans on
publisher-supplied country codes and genuinely needs checking.
:::

### How precisely are the records placed?

Passing the coordinate tests means a record is not obviously *wrong*. It does
not mean the record is precise enough for every purpose. A great many Kosovo
records come from atlas-style surveys that report to a 10 km square, and the
stated coordinate uncertainty is the honest way to tell those apart from a
GPS-located field observation.

```{r}
#| label: precision-table

report_table(
  meta$precision_report,
  caption = htmltools::tags$caption(
    style = "caption-side: bottom; text-align: left; padding-top: 8px;
             font-size: 0.85rem; color: #6b7880;",
    "Table 3. Stated coordinate uncertainty of the cleaned records. ",
    "‘Not stated’ means the publisher supplied no uncertainty value; it is ",
    "not a claim of precision either way."
  ),
  page_length = 6, buttons = FALSE, escape = FALSE
) |>
  DT::formatCurrency("Records", currency = "", interval = 3, mark = ",",
                     digits = 0)
```

::: {.callout-note collapse="true"}
## Why coarse records are kept rather than filtered out

The GBIF Viewer developed for Ukraine discards records whose stated uncertainty
exceeds a threshold, on the reasonable grounds that a record placed to the
nearest 10 km cannot support a decision about a particular field.

Applied here, that rule would remove more than half the dataset. Around 54 per
cent of Kosovo records carry an uncertainty of exactly 7,071 m, which is the
half-diagonal of a 10 km square: these are atlas records, and they are perfectly
good evidence that a species occurs in a landscape even though they say nothing
about where in it. Discarding them would make the country look emptier than it
is, which is the more dangerous error for conservation planning.

The pipeline therefore reports the precision profile, keeps
`coordinateUncertaintyInMeters` in every export, and offers precision as a
filter in the record explorer below. The threshold is available as
`coord_uncertainty_max` in `pipeline.R` for anyone who does want it applied.
:::

## Where the records are

Each map carries the same set of switchable layers, reached through the control
in the top-right corner:

- **Occurrence records** — individual records, clustered until you zoom in,
  each with a pop-up carrying the determination, the place, the stated accuracy and a link through to the
  record on GBIF, where the evidence and any photographs can be checked.
- **Record density** — every record aggregated onto a 2 km grid and classed
  against a legend. This is the layer to read when the question is *how many*.
- **Heat map** — the same information as a smooth surface, better for seeing
  the shape of the sampled and unsampled areas.
- **Municipalities** — administrative boundaries, with each unit's own record
  and species count on hover.

The toolbar on the left adds full-screen, a distance and area measure, a
polygon and rectangle sketching tool, and a place search, so an area of
interest can be outlined and sized directly on the map.

::: {.panel-tabset}

### All biodiversity

```{r}
#| label: map-overall
build_occurrence_map(overall, boundary = boundary,
                     municipalities = municipalities,
                     colour_by = "kingdom", max_points = 2500)
```

### Birds

```{r}
#| label: map-birds
build_occurrence_map(subsets[["kosovo_birds"]], boundary = boundary,
                     municipalities = municipalities,
                     colour_by = "iucn", max_points = 2500)
```

### Birds Directive, Annex I

```{r}
#| label: map-birds-annex-i
build_occurrence_map(subsets[["kosovo_birds_annex_I"]], boundary = boundary,
                     municipalities = municipalities,
                     colour_by = "iucn", max_points = 2500)
```

### Habitats Directive

```{r}
#| label: map-habitats
build_occurrence_map(subsets[["kosovo_habitats_directive"]],
                     boundary = boundary, municipalities = municipalities,
                     colour_by = "kingdom")
```

### Habitats Directive, Annex II

```{r}
#| label: map-habitats-annex-ii
build_occurrence_map(subsets[["kosovo_habitats_annex_II"]],
                     boundary = boundary, municipalities = municipalities,
                     colour_by = "kingdom")
```

### IUCN threatened

```{r}
#| label: map-threatened
build_occurrence_map(subsets[["kosovo_threatened_iucn"]],
                     boundary = boundary, municipalities = municipalities,
                     colour_by = "iucn")
```

:::

Both density layers are the more informative reading of **survey effort**.
Dense cells frequently mark well-visited sites rather than exceptionally rich
ones, and pale areas usually indicate that nobody has yet surveyed there. Both
readings matter when prioritising future fieldwork.

::: {.callout-note collapse="true"}
## How the density layers are built, and why the heat map changed

Occurrence counts are extraordinarily uneven. Rounded to about 110 m, Kosovo
has 5,590 occupied positions; three quarters of them hold a single record, and
the busiest holds more than thirteen thousand. Any display that treats that
range naively collapses.

The heat map previously did exactly that. Leaflet's heat layer draws each point
at an opacity of `intensity / max` and then colours the accumulated opacity
through a gradient, so feeding it raw counts made every occupied cell saturate
identically. Two further defaults compounded it: a minimum-opacity floor that
lifted empty-ish areas up to the same value as real clusters, and an internal
zoom scaling that divides every intensity by `2^(maxZoom - currentZoom)` —
about a thousandfold at the country view — pushing the whole surface onto that
floor. The result was a uniform blue wash.

The layer now: aggregates onto the 110 m grid; maps the counts onto 0–1 with a
logarithmic transform followed by a power below one, which spreads the crowded
bottom of the distribution across usable values; pins the zoom reference so the
surface means the same thing at every scale; and uses a single-hue ramp running
light to dark, so that "more" reads as "darker" rather than as "a different
colour".

The **Record density** grid is the same information without the smoothing:
2 km cells, classed on roughly logarithmic breaks, with a legend and the exact
count on hover. Use the heat map to see the shape of the coverage and the grid
when a number is needed.
:::

::: {.callout-note collapse="true"}
## About the common names shown in the pop-ups

Common names are not part of the GBIF download and are retrieved separately
from the GBIF species API, which returns every name any publisher has supplied.
The pipeline takes the most frequently published English name for each taxon,
which recovers the standard name in the large majority of cases.

Publishers do, however, mis-tag the language of their names fairly often —
roughly a fifth of the names returned as English for this dataset were in fact
Norwegian, German, French, Swedish, Hungarian or Ukrainian. Names written in
non-Latin scripts are filtered out automatically, but a mis-tagged
Latin-script name occasionally survives. **The scientific name is always the
authoritative identifier**; treat the common name as a convenience only.
:::

## Survey coverage by municipality

The map above answers "where are the records". This one answers the question a
planning authority actually asks: **how well is my municipality covered?**
Records are counted per unit of area, because a large municipality with a
moderate number of records may be far less well known than a small one with
fewer.

```{r}
#| label: municipal-map

municipal_sf <- summarise_by_municipality(overall_flat, municipalities)

build_municipal_map(municipal_sf, value = "records_per_km2",
                    title = "Records per km&sup2;", boundary = boundary)
```

```{r}
#| label: municipal-table

municipal_tbl <- meta$municipal_summary |>
  transmute(
    Municipality       = municipality,
    District           = district,
    `Area (km²)`       = round(area_km2),
    Records            = records,
    Species            = species,
    `Records per km²`  = round(records_per_km2, 1),
    `Most recent year` = last_year
  ) |>
  arrange(`Records per km²`)

report_table(
  municipal_tbl,
  caption = htmltools::tags$caption(
    style = "caption-side: bottom; text-align: left; padding-top: 8px;
             font-size: 0.85rem; color: #6b7880;",
    sprintf(paste0("Table 4. Coverage by municipality, least-recorded first. ",
                   "Boundaries are GADM level 2; names are given in both ",
                   "official languages. Median coverage is %.1f records per ",
                   "km², and the best-covered unit carries %.0f times the ",
                   "records per unit area of the worst."),
            stats::median(meta$municipal_summary$records_per_km2),
            max(meta$municipal_summary$records_per_km2) /
              max(min(meta$municipal_summary$records_per_km2), 1e-9))
  ),
  page_length = 10
) |>
  DT::formatCurrency(c("Records", "Species", "Area (km²)"),
                     currency = "", interval = 3, mark = ",", digits = 0)
```

::: {.callout-warning collapse="true"}
## One municipality dominates the totals — read it carefully

Fushë Kosovë / Kosovo Polje holds far more records per square kilometre than
anywhere else in the country. That is not a statement about its biodiversity.
A single coordinate inside it carries more than thirteen thousand records,
almost all of them bird observations submitted to one citizen-science platform
from what is in practice one regularly watched site.

This is the characteristic shape of aggregated occurrence data and the reason
raw record counts should never be read as species richness. The species column
in the table above is the more honest comparison, and the two rank very
differently.
:::

## Species protected under the EU Nature Directives

The annex lists were taken directly from the consolidated legal texts on
EUR-Lex — 193 taxa on Birds Directive Annex I, and 1,333 across Habitats
Directive Annexes II, IV and V — and every name was resolved against the GBIF
backbone taxonomy so that records published under superseded names are still
found. *Larus minutus*, for instance, is listed under that name but recorded as
*Hydrocoloeus minutus*.

The table below shows the species of Community interest that **have** been
recorded in Kosovo. The annexes are EU-wide lists dominated by Iberian,
Alpine and Macaronesian endemics, so a listed species with no records here
usually lies outside its natural range rather than indicating a survey gap;
the meaningful gaps are the Balkan species one would expect to find and
doesn't.

```{r}
#| label: directive-table

# Built from the annotated subsets themselves, so that each row reports the
# species as GBIF currently accepts it, together with the annexes it was
# matched against.
directive_subsets <- bind_rows(
  if (!is.null(subsets[["kosovo_birds_annex_I"]]))
    sf::st_drop_geometry(subsets[["kosovo_birds_annex_I"]]) |>
      mutate(Directive = "Birds"),
  if (!is.null(subsets[["kosovo_habitats_directive"]]))
    sf::st_drop_geometry(subsets[["kosovo_habitats_directive"]]) |>
      mutate(Directive = "Habitats")
)

directive_tbl <- directive_subsets |>
  filter(!is.na(species), nzchar(species)) |>
  mutate(year = suppressWarnings(as.integer(year))) |>
  group_by(`Scientific name` = species, Directive) |>
  summarise(
    `Common name` = {
      v <- vernacularName[!is.na(vernacularName)]
      if (length(v)) v[1] else "&mdash;"
    },
    Annex = paste(sort(unique(unlist(strsplit(na.omit(annex), "; ")))),
                  collapse = ", "),
    IUCN = {
      v <- iucnRedListCategory[!is.na(iucnRedListCategory)]
      if (length(v)) v[1] else "NE"
    },
    Records = n(),
    `Most recent year` = {
      y <- year[!is.na(year) & year > 1500]
      if (length(y)) max(y) else NA_integer_
    },
    .groups = "drop"
  ) |>
  mutate(IUCN = swatch(unname(map_palette$iucn[IUCN]), IUCN)) |>
  select(`Scientific name`, `Common name`, Directive, Annex, IUCN, Records,
         `Most recent year`) |>
  arrange(desc(Records), `Scientific name`)

report_table(
  directive_tbl,
  caption = htmltools::tags$caption(
    style = "caption-side: bottom; text-align: left; padding-top: 8px;
             font-size: 0.85rem; color: #6b7880;",
    sprintf(paste0("Table 5. Species of Community interest recorded in Kosovo: ",
                   "%d taxa across %s quality-checked records. The annex column ",
                   "shows every annex a species was matched against, so a ",
                   "species listed on more than one appears once."),
            nrow(directive_tbl), fmt_int(sum(directive_tbl$Records)))
  ),
  page_length = 15,
  escape = FALSE
) |>
  DT::formatCurrency("Records", currency = "", interval = 3, mark = ",",
                     digits = 0)
```

::: {.callout-warning collapse="true"}
## How annex membership was determined

The species lists are built by `R/build_directive_list.R`, which parses the
consolidated directives on EUR-Lex — the legally binding texts — rather than
relying on a secondary checklist. Every name is then resolved against the GBIF
backbone taxonomy, and occurrence records are selected by numeric key rather
than by name, which is robust to synonymy and authorship variation.

Two details materially affect what appears in these subsets.

**Rank matters when matching.** A taxon listed at species rank is matched on
both its species key and its taxon key, so records identified to subspecies are
still captured. A taxon listed at *subspecies* rank is matched on that
subspecies alone. Promoting subspecies listings to their whole species would be
seriously misleading: the Birds Directive lists island endemics such as
*Columba palumbus azorica* and *Fringilla coelebs ombriosa*, and matching those
at species level would place every Wood Pigeon and Chaffinch in Kosovo on
Annex I — around 1,450 records for subspecies confined to the Azores and the
Canaries.

**Exclusions are honoured.** Where an annex lists a species *except* certain
subspecies, as Annex V does for *Rupicapra rupicapra*, the excluded taxa are
discarded rather than being read as listings in their own right.

Around 1.3 per cent of annex entries could not be resolved to a current GBIF
taxon. These are names superseded since the directives were adopted, mostly
Iberian and Macaronesian plants; none are species expected in Kosovo. The
`directive` and `annex` columns are retained in every export so that any
selection can be checked.
:::

## Extinction risk: the IUCN Red List

The EU annexes say what is legally protected. The IUCN Red List says what is at
risk of extinction. The two overlap only partly, and a screening exercise needs
both: a species can be at serious risk without appearing on any annex, and much
of what is annexed is of least concern globally.

Every species in the dataset was matched to its global Red List assessment
through the GBIF backbone taxonomy.

```{r}
#| label: iucn-summary

iucn_tbl <- meta$iucn_summary |>
  mutate(
    Category = swatch(unname(map_palette$iucn[as.character(category)]),
                      sprintf("%s &mdash; %s", as.character(category),
                              unname(iucn_names[as.character(category)]))),
    `Per cent of species` = round(100 * species / sum(species), 1)
  ) |>
  select(Category, Species = species, `Per cent of species`)

report_table(
  iucn_tbl,
  caption = htmltools::tags$caption(
    style = "caption-side: bottom; text-align: left; padding-top: 8px;
             font-size: 0.85rem; color: #6b7880;",
    sprintf(paste0("Table 6. Global Red List assessment of the %s species ",
                   "recorded in Kosovo. Most invertebrates and fungi have ",
                   "never been assessed, which is why ‘not evaluated’ is the ",
                   "largest class; it means the assessment is missing, not ",
                   "that the species is safe."),
            fmt_int(sum(meta$iucn_summary$species)))
  ),
  page_length = 9, buttons = FALSE, escape = FALSE
)
```

```{r}
#| label: threatened-table

threatened_tbl <- sf::st_drop_geometry(subsets[["kosovo_threatened_iucn"]]) |>
  mutate(year = suppressWarnings(as.integer(year))) |>
  group_by(`Scientific name` = species) |>
  summarise(
    `Common name` = {
      v <- vernacularName[!is.na(vernacularName)]
      if (length(v)) v[1] else "&mdash;"
    },
    Category = first(iucnRedListCategory),
    Class    = coalesce(first(class), "—"),
    Records  = n(),
    Municipalities = n_distinct(municipality[!is.na(municipality)]),
    `Most recent year` = {
      y <- year[!is.na(year) & year > 1500]
      if (length(y)) max(y) else NA_integer_
    },
    .groups = "drop"
  ) |>
  arrange(match(Category, iucn_levels), desc(Records)) |>
  mutate(Category = swatch(unname(map_palette$iucn[Category]), Category)) |>
  select(`Scientific name`, `Common name`, Class, Category, Records,
         Municipalities, `Most recent year`)

report_table(
  threatened_tbl,
  caption = htmltools::tags$caption(
    style = "caption-side: bottom; text-align: left; padding-top: 8px;
             font-size: 0.85rem; color: #6b7880;",
    sprintf(paste0("Table 7. The %d species recorded in Kosovo and assessed as ",
                   "Critically Endangered, Endangered or Vulnerable, ordered ",
                   "by severity. Mapped on the ‘IUCN threatened’ tab above."),
            nrow(threatened_tbl))
  ),
  page_length = 12, escape = FALSE
)
```

::: {.callout-note collapse="true"}
## Reading these assessments with care

Two cautions apply.

**The assessment is global, not national.** A species of least concern across
Europe may be on the edge of its range and genuinely rare in Kosovo, and
nothing in this table would show it. A national red list is the instrument for
that question; this is a screen, not a substitute.

**Backbone matching can misfire.** The Red List category is attached through
the GBIF backbone taxonomy, which occasionally links a name to an assessment of
a different taxon. One record in this dataset is attributed to a species
assessed as *Extinct*, which for a beetle collected in the Balkans is far more
likely to be a taxonomic mismatch than a rediscovery. Treat a surprising
category as a question to check rather than as a finding.
:::

## Explore the records

The GBIF Viewer's most useful idea is its filter panel: pick the conservation
criteria that matter to you, and see what is left. That is reproduced here for
the records of conservation interest — everything listed on an EU annex or
assessed as threatened — with the filtering done in your own browser, so it
costs nothing to host and keeps working for as long as this page does.

Filters combine: choosing a class *and* a Red List category shows records that
satisfy both. Selections are shared between the map and the table.

```{r}
#| label: explorer

explorer_src <- bind_rows(
  sf::st_drop_geometry(subsets[["kosovo_birds_annex_I"]]),
  sf::st_drop_geometry(subsets[["kosovo_habitats_directive"]]),
  sf::st_drop_geometry(subsets[["kosovo_threatened_iucn"]])
) |>
  distinct(gbifID, .keep_all = TRUE) |>
  mutate(year = suppressWarnings(as.integer(year))) |>
  filter(!is.na(year), year > 1500, !is.na(species), nzchar(species))

precision_band <- function(x) {
  b <- cut(suppressWarnings(as.numeric(x)),
           breaks = meta$config$precision_breaks, right = TRUE,
           labels = c("100 m or better", "100 m – 1 km", "1 – 10 km",
                      "coarser than 10 km"))
  factor(ifelse(is.na(b), "not stated", as.character(b)),
         levels = c("100 m or better", "100 m – 1 km", "1 – 10 km",
                    "coarser than 10 km", "not stated"))
}

iucn_code <- as.character(iucn_group(explorer_src$iucnRedListCategory))

# The table and the map are given their own `SharedData` over the same keys and
# the same crosstalk group. A single object would work, but every column would
# then be embedded twice — once in the table's payload and once in the map's —
# and the pop-up HTML alone would add close to a megabyte to the page for no
# visible benefit. Filter events are broadcast by group, so the two stay in
# step regardless.
keys <- as.character(seq_len(nrow(explorer_src)))

explorer <- tibble::tibble(
  Species       = explorer_src$species,
  `Common name` = coalesce(explorer_src$vernacularName, "—"),
  Class         = coalesce(explorer_src$class, "—"),
  IUCN          = iucn_code,
  Listing       = coalesce(explorer_src$annex, "not listed"),
  Year          = explorer_src$year,
  Municipality  = coalesce(explorer_src$municipality, "—"),
  Precision     = as.character(
                    precision_band(explorer_src$coordinateUncertaintyInMeters)),
  Basis         = tolower(gsub("_", " ", explorer_src$basisOfRecord))
)

explorer_pts <- tibble::tibble(
  lon   = explorer_src$decimalLongitude,
  lat   = explorer_src$decimalLatitude,
  .fill = unname(map_palette$iucn[iucn_code]),
  popup = build_explorer_popup(explorer_src)
)

sd_tab <- crosstalk::SharedData$new(explorer, key = keys, group = "records")
sd_map <- crosstalk::SharedData$new(explorer_pts, key = keys, group = "records")

filters <- htmltools::div(
  class = "explorer-filters",
  crosstalk::filter_select("f_class", "Taxonomic class", sd_tab, ~Class),
  crosstalk::filter_select("f_iucn",  "IUCN Red List",   sd_tab, ~IUCN),
  crosstalk::filter_select("f_annex", "EU annex",        sd_tab, ~Listing),
  crosstalk::filter_select("f_basis", "Basis of record", sd_tab, ~Basis),
  crosstalk::filter_select("f_mun",   "Municipality",    sd_tab, ~Municipality),
  crosstalk::filter_select("f_prec",  "Coordinate precision", sd_tab, ~Precision),
  # A zero-width space, not an empty string, as the thousands separator: the
  # slider's JavaScript treats an empty `data-prettify-separator` as "unset"
  # and falls back to a space, which turns every year into "2 026".
  crosstalk::filter_slider("f_year",  "Year", sd_tab, ~Year, sep = "​")
)

htmltools::tagList(
  htmltools::div(
    class = "explorer-grid",
    filters,
    htmltools::div(
      class = "explorer-map",
      build_explorer_map(sd_map, boundary = boundary,
                         legend_codes = sort(unique(iucn_code)))
    )
  ),
  DT::datatable(
    sd_tab,
    rownames = FALSE,
    class    = "stripe hover compact order-column",
    options  = list(
      dom        = "Bfrtip",
      buttons    = list(list(extend = "csv",   text = "Download CSV"),
                        list(extend = "excel", text = "Download Excel")),
      pageLength = 10,
      scrollX    = TRUE,
      order      = list(list(5, "desc")),
      language   = list(search = "Filter:",
                        emptyTable = "No records match the current filters")
    ),
    extensions = "Buttons"
  )
)
```

```{r}
#| label: explorer-note
#| output: asis

cat(sprintf(
  paste0("<p class='explorer-note'>%s records of conservation interest, ",
         "covering %s species. Records without a usable year are omitted from ",
         "the explorer so that the year filter behaves predictably; they ",
         "remain in the downloadable files.</p>"),
  fmt_int(nrow(explorer)), fmt_int(dplyr::n_distinct(explorer$Species))))
```

## The species checklist

Every species in the cleaned dataset, with the evidence behind it. Search by
scientific or common name; each name links to its GBIF species page, where the
full record set, images and taxonomic history can be inspected.

```{r}
#| label: checklist

checklist <- overall_flat |>
  filter(!is.na(species), nzchar(species)) |>
  mutate(year = suppressWarnings(as.integer(year))) |>
  group_by(Species = species) |>
  summarise(
    `Common name` = {
      v <- vernacularName[!is.na(vernacularName)]
      if (length(v)) v[1] else "—"
    },
    Kingdom = first(kingdom),
    Class   = coalesce(first(class), "—"),
    Family  = coalesce(first(family), "—"),
    IUCN    = first(iucnRedListCategory),
    Records = n(),
    Municipalities = n_distinct(municipality[!is.na(municipality)]),
    `First year` = {
      y <- year[!is.na(year) & year > 1500]; if (length(y)) min(y) else NA_integer_
    },
    `Most recent year` = {
      y <- year[!is.na(year) & year > 1500]; if (length(y)) max(y) else NA_integer_
    },
    key = first(speciesKey),
    .groups = "drop"
  ) |>
  # No colour swatch on this table, unlike the shorter ones above: it would add
  # sixty bytes to each of four thousand rows for decoration, and the category
  # code in the cell is already the thing that carries the meaning.
  mutate(
    IUCN = coalesce(IUCN, "NE"),
    Species = sprintf("<a href='https://www.gbif.org/species/%s' target='_blank' rel='noopener'><i>%s</i></a>",
                      key, Species)
  ) |>
  select(-key) |>
  arrange(desc(Records))

report_table(
  checklist,
  caption = htmltools::tags$caption(
    style = "caption-side: bottom; text-align: left; padding-top: 8px;
             font-size: 0.85rem; color: #6b7880;",
    sprintf(paste0("Table 8. The %s species recorded in Kosovo, most-recorded ",
                   "first. ‘Municipalities’ counts the administrative units a ",
                   "species has been found in, which is a rough but useful ",
                   "guide to how widely it has been looked for."),
            fmt_int(nrow(checklist)))
  ),
  page_length = 15, escape = FALSE
) |>
  DT::formatCurrency("Records", currency = "", interval = 3, mark = ",",
                     digits = 0)
```

## Who published these records

GBIF-mediated data is other people's work. Every record in this report was
collected, verified and published by one of the institutions below, and
crediting them is a condition of use rather than a courtesy.

```{r}
#| label: datasets-table

datasets_tbl <- meta$datasets |>
  transmute(
    Dataset = ifelse(
      is.na(datasetTitle), datasetKey,
      sprintf("<a href='https://www.gbif.org/dataset/%s' target='_blank' rel='noopener'>%s</a>",
              datasetKey, datasetTitle)),
    Publisher = coalesce(publisher, "—"),
    Licence   = short_licence(datasetLicense),
    Records   = records,
    Species   = species
  )

report_table(
  datasets_tbl,
  caption = htmltools::tags$caption(
    style = "caption-side: bottom; text-align: left; padding-top: 8px;
             font-size: 0.85rem; color: #6b7880;",
    sprintf(paste0("Table 9. The %d datasets contributing records, from %d ",
                   "publishing institutions, largest contribution first. The ",
                   "licence column gives the terms each publisher chose; the ",
                   "same information is in the `license` column of every ",
                   "export, per record."),
            nrow(datasets_tbl), dplyr::n_distinct(meta$datasets$publisher))
  ),
  page_length = 10, escape = FALSE
) |>
  DT::formatCurrency(c("Records", "Species"), currency = "", interval = 3,
                     mark = ",", digits = 0)
```

## Download the data

All six extracts are published as **GeoPackage** (`.gpkg`, ready for QGIS or
ArcGIS, geometry included) and **CSV** (flat table with coordinate columns).
The smaller extracts are also published as **Excel** workbooks. All are open
formats requiring no proprietary software.

```{r}
#| label: downloads
#| output: asis

mf <- meta$manifest

cards <- vapply(seq_len(nrow(mf)), function(i) {

  btn <- function(file, size, label, class) {
    if (is.na(file)) return("")
    sprintf("<a class='btn %s' href='data_exports/%s' download>%s (%s)</a>",
            class, file, label, fmt_bytes(size))
  }

  sprintf(paste0("<div class='download-card'><h4>%s</h4>",
                 "<div class='meta'>%s records &middot; %s species</div>%s%s%s</div>"),
          mf$label[i], fmt_int(mf$records[i]), fmt_int(mf$species[i]),
          btn(mf$gpkg[i], mf$gpkg_size[i], "GeoPackage", "btn-primary"),
          btn(mf$csv[i],  mf$csv_size[i],  "CSV",        "btn-outline-secondary"),
          btn(mf$xlsx[i], mf$xlsx_size[i], "Excel",      "btn-outline-secondary"))
}, character(1))

cat("<div class='download-grid'>", paste(cards, collapse = "\n"), "</div>", sep = "\n")
```

::: {.callout-tip}
## Conditions of use

GBIF-mediated data are made available under the licences chosen by each
publishing institution; the `license` column in every export records the terms
applying to each record. When these data are used in a report or publication,
please cite the DOI given above, which credits all contributing institutions.
:::

## Methods and reproducibility

The entire analysis is scripted and version-controlled. To reproduce it:

1. Install the R packages listed below.
2. Add your GBIF credentials to `.Renviron` (see the repository `README.md`).
3. Run `Rscript pipeline.R` to acquire, clean, enrich and export the data.
4. Run `quarto render` to rebuild this website.

`pipeline.R` records its GBIF download key, so re-running it re-uses the
existing DOI rather than minting a new one. Delete
`data/gbif_download/download_key.txt` to force a fresh extract. The vernacular
name, Red List and dataset-registry lookups are cached under `data/`, so only
the first run makes those requests.

```{r}
#| label: session
#| output: asis

pkgs <- c("rgbif", "sf", "CoordinateCleaner", "leaflet", "leaflet.extras",
          "leafem", "crosstalk", "dplyr", "readr", "DT", "writexl",
          "rnaturalearth")
versions <- vapply(pkgs, function(p) {
  tryCatch(as.character(utils::packageVersion(p)), error = function(e) "not installed")
}, character(1))

cat("<details><summary>Software environment</summary>\n\n",
    sprintf("- %s\n", R.version.string),
    sprintf("- Analysis run: %s\n", fmt_date_en(meta$run_date, time = TRUE)),
    "\n",
    paste(sprintf("- `%s` %s\n", pkgs, versions), collapse = ""),
    "\n</details>\n", sep = "")
```
