---
title: "Biodiversity of Kosovo"
subtitle: "An analysis of GBIF occurrence data in support of conservation planning"
date: today
date-format: "D MMMM YYYY"
lang: en-GB
---

```{r}
#| label: setup

suppressPackageStartupMessages({
  library(dplyr)
  library(sf)
  library(DT)
  library(leaflet)
  library(leaflet.extras)
  library(htmltools)
})

source("R/functions.R")

meta_path <- "data/run_metadata.rds"
pipeline_has_run <- file.exists(meta_path)

if (pipeline_has_run) meta <- readRDS(meta_path)

# Human-readable file sizes for the download panel.
fmt_bytes <- function(x) {
  ifelse(is.na(x), "&mdash;",
    ifelse(x >= 1024^2, sprintf("%.1f MB", x / 1024^2),
      ifelse(x >= 1024, sprintf("%.0f kB", x / 1024), sprintf("%.0f B", x))))
}

# A consistently styled data table for the whole report.
report_table <- function(x, caption = NULL, page_length = 10,
                         buttons = TRUE, ...) {
  DT::datatable(
    x,
    rownames   = FALSE,
    caption    = caption,
    extensions = if (buttons) "Buttons" else character(),
    options = list(
      dom        = if (buttons) "Bfrtip" else "ftip",
      buttons    = if (buttons) list(
                     list(extend = "copy",  text = "Copy"),
                     list(extend = "csv",   text = "Download CSV"),
                     list(extend = "excel", text = "Download Excel")
                   ) else list(),
      pageLength = page_length,
      scrollX    = TRUE,
      autoWidth  = FALSE,
      language   = list(search = "Filter:", emptyTable = "No records available")
    ),
    class = "stripe hover compact order-column",
    ...
  )
}
```

```{r}
#| label: guard
#| output: asis

if (!pipeline_has_run) {
  cat(
    "::: {.callout-important}\n",
    "## The data pipeline has not yet been run\n\n",
    "This report is generated from files produced by `pipeline.R`. ",
    "Set your GBIF credentials in `.Renviron`, then run:\n\n",
    "```bash\nRscript pipeline.R\n```\n\n",
    "and render the site again.\n",
    ":::\n",
    sep = ""
  )
  knitr::knit_exit()
}
```

```{r}
#| label: load-subsets

subset_files <- file.path("data_exports", paste0(meta$manifest$name, ".gpkg"))
names(subset_files) <- meta$manifest$name

read_subset <- function(path) {
  if (file.exists(path)) sf::st_read(path, quiet = TRUE) else NULL
}

subsets  <- lapply(subset_files, read_subset)
boundary <- if (file.exists("data/kosovo_boundary.gpkg")) {
  sf::st_read("data/kosovo_boundary.gpkg", quiet = TRUE)
} else NULL
```

## Why this matters

Kosovo sits within the Balkan biodiversity hotspot, one of the most
species-rich regions in Europe. Evidence-based conservation — designating
protected areas, screening development proposals, and reporting against the EU
Nature Directives — depends on knowing **which species occur where**.

The Global Biodiversity Information Facility (GBIF) aggregates species
occurrence records published by museums, research institutions, monitoring
programmes and citizen-science platforms worldwide. This report brings the
GBIF holdings for the territory of Kosovo into a single, quality-controlled,
openly downloadable evidence base, and highlights the records that relate to
species protected under the EU Birds and Habitats Directives.

```{r}
#| label: key-figures
#| output: asis

overall <- subsets[["kosovo_overall_biodiversity"]]
m       <- meta$manifest

kf <- function(value, label) {
  sprintf("<div class='keyfigure'><span class='value'>%s</span><span class='label'>%s</span></div>",
          value, label)
}

years <- suppressWarnings(as.integer(overall$year))
years <- years[!is.na(years) & years > 1500]

cat("<div class='keyfigures'>",
    kf(fmt_int(meta$n_clean), "Quality-checked records"),
    kf(fmt_int(meta$n_species), "Species recorded"),
    kf(fmt_int(m$records[m$name == "kosovo_birds"]), "Bird records"),
    kf(fmt_int(sum(m$records[m$name %in% c("kosovo_birds_annex_I",
                                           "kosovo_habitats_directive")])),
       "Directive-listed records"),
    kf(if (length(years)) sprintf("%d&ndash;%d", min(years), max(years)) else "&mdash;",
       "Temporal coverage"),
    "</div>", sep = "\n")
```

## Data provenance and citation

The occurrence data were obtained through the GBIF asynchronous download
service, which mints a permanent Digital Object Identifier (DOI) for the exact
extract used here. Citing that DOI allows any reader to retrieve the identical
dataset, and passes credit back to the institutions that published the records.

```{r}
#| label: citation
#| output: asis

doi <- meta$download$doi
cat(
  "<div class='citation-box'>",
  sprintf("<strong>Dataset DOI:</strong> <a href='https://doi.org/%s'>https://doi.org/%s</a><br>", doi, doi),
  sprintf("<strong>GBIF download key:</strong> <code>%s</code><br>", meta$download$key),
  sprintf("<strong>Retrieved:</strong> %s<br>", format(as.Date(meta$run_date), "%d %B %Y")),
  sprintf("<strong>Spatial filter:</strong> GADM boundary <code>%s</code> (Kosovo)<br><br>",
          meta$config$gadm_gid),
  "<strong>Suggested citation</strong><br>",
  sprintf("<code>%s</code>", meta$download$citation %||% paste0("GBIF Occurrence Download https://doi.org/", doi)),
  "</div>", sep = "\n"
)
```

::: {.callout-note collapse="true"}
## A note on the spatial filter

Records were selected using the GADM administrative boundary for Kosovo
(GID `XKO`) rather than the ISO country code `XK`. The GADM boundary is a true
spatial filter, so it also captures records whose publisher left the country
field blank or assigned it to a neighbouring state — a common situation for
cross-border survey data. Queried against the GBIF occurrence API without any
further filters at the time of extraction, `gadmGid=XKO` returned 52,290
records against 48,947 for `country=XK`, about 7 per cent more.

Note that the GADM identifier for Kosovo is `XKO`. The similar-looking code
`XKX` is the World Bank / ISO alpha-3 style code and is **not** recognised by
GADM — `pred("gadm", "XKX")` returns zero records rather than an error, so a
pipeline using it would silently produce an empty dataset.
:::

## Summary statistics

Each row below is a distinct thematic extract of the cleaned dataset. All five
are available for download at the end of this report.

```{r}
#| label: summary-table

summary_tbl <- bind_rows(lapply(seq_len(nrow(meta$manifest)), function(i) {
  nm <- meta$manifest$name[i]
  summarise_subset(subsets[[nm]], meta$manifest$label[i])
}))

report_table(
  summary_tbl,
  caption = htmltools::tags$caption(
    style = "caption-side: bottom; text-align: left; padding-top: 8px;
             font-size: 0.85rem; color: #6b7880;",
    "Table 1. Cleaned GBIF occurrence records for Kosovo, by thematic subset. ",
    "‘Source datasets’ counts the distinct GBIF datasets contributing records."
  ),
  page_length = 5
) |>
  DT::formatCurrency(c("Records", "Species", "Families", "Source datasets"),
                     currency = "", interval = 3, mark = ",", digits = 0)
```

## Data quality control

Aggregated occurrence data carries well-known geographic artefacts: records
placed at the centre of a country, at a museum's own address, or at coordinates
of exactly zero. Every record was screened with `CoordinateCleaner` and only
records passing all tests were retained.

```{r}
#| label: cleaning-table

report_table(
  meta$cleaning_report |> rename(`Records flagged` = Flagged),
  caption = htmltools::tags$caption(
    style = "caption-side: bottom; text-align: left; padding-top: 8px;
             font-size: 0.85rem; color: #6b7880;",
    sprintf(paste0("Table 2. Outcome of the coordinate screening. %s of %s ",
                   "records (%.1f%%) passed every test and form the basis of ",
                   "this report. A record may be flagged by more than one test."),
            fmt_int(meta$n_clean), fmt_int(meta$n_raw),
            100 * meta$n_clean / meta$n_raw)
  ),
  page_length = 10, buttons = FALSE
)
```

::: {.callout-note collapse="true"}
## Which tests were applied, and why two were left out

Applied: zero coordinates, country-coordinate mismatch, GBIF headquarters,
known biodiversity institutions, and identical latitude/longitude values.

Deliberately **not** applied: the *capital city* and *country centroid* tests.
Both discard records falling within a fixed radius of a capital or of a
country's geometric centre. Kosovo is small enough that Pristina and the
national centroid both sit well inside the territory, so these tests would
remove a large number of entirely legitimate records — including most urban
biodiversity data. They can be enabled in `pipeline.R` by setting
`use_capitals_test = TRUE`.

The country-coordinate test also required a correction. Natural Earth, the
reference dataset used by `CoordinateCleaner`, records Kosovo with no assigned
ISO alpha-3 code. Running the test against the default reference flags
**100 per cent** of Kosovo records as mismatches and silently returns an empty
dataset. The pipeline therefore supplies a bespoke Kosovo reference polygon.
:::

## Where the records are

Each map offers a clustered point layer with record details on click, a
toggleable heat map showing sampling density, and a choice of base maps. Use
the layer control in the top-right corner to switch between them.

::: {.panel-tabset}

### All biodiversity

```{r}
#| label: map-overall
build_occurrence_map(subsets[["kosovo_overall_biodiversity"]],
                     boundary = boundary, colour = "#14556b")
```

### Birds

```{r}
#| label: map-birds
build_occurrence_map(subsets[["kosovo_birds"]],
                     boundary = boundary, colour = "#1b6ca8")
```

### Birds Directive, Annex I

```{r}
#| label: map-birds-annex-i
build_occurrence_map(subsets[["kosovo_birds_annex_I"]],
                     boundary = boundary, colour = "#b8560f")
```

### Habitats Directive

```{r}
#| label: map-habitats
build_occurrence_map(subsets[["kosovo_habitats_directive"]],
                     boundary = boundary, colour = "#2e7d54")
```

### Habitats Directive, Annex II

```{r}
#| label: map-habitats-annex-ii
build_occurrence_map(subsets[["kosovo_habitats_annex_II"]],
                     boundary = boundary, colour = "#6a3d9a")
```

:::

The heat map is the more informative layer for interpreting **survey effort**.
Dense clusters frequently mark well-visited sites rather than exceptionally
rich ones, and sparse areas usually indicate that nobody has yet surveyed
there. Both readings matter when prioritising future fieldwork.

::: {.callout-note collapse="true"}
## About the common names shown in the pop-ups

Common names are not part of the GBIF download and are retrieved separately
from the GBIF species API, which returns every name any publisher has supplied.
The pipeline takes the most frequently published English name for each taxon,
which recovers the standard name in the large majority of cases.

Publishers do, however, mis-tag the language of their names fairly often —
roughly a fifth of the names returned as English for this dataset were in fact
Norwegian, German, French, Swedish, Hungarian or Ukrainian. Names written in
non-Latin scripts are filtered out automatically, but a mis-tagged
Latin-script name occasionally survives. **The scientific name is always the
authoritative identifier**; treat the common name as a convenience only.
:::

## Species protected under the EU Nature Directives

The annex lists were taken directly from the consolidated legal texts on
EUR-Lex — 193 taxa on Birds Directive Annex I, and 1,333 across Habitats
Directive Annexes II, IV and V — and every name was resolved against the GBIF
backbone taxonomy so that records published under superseded names are still
found. *Larus minutus*, for instance, is listed under that name but recorded as
*Hydrocoloeus minutus*.

The table below shows the species of Community interest that **have** been
recorded in Kosovo. The annexes are EU-wide lists dominated by Iberian,
Alpine and Macaronesian endemics, so a listed species with no records here
usually lies outside its natural range rather than indicating a survey gap;
the meaningful gaps are the Balkan species one would expect to find and
doesn't.

```{r}
#| label: directive-table

# Built from the annotated subsets themselves, so that each row reports the
# species as GBIF currently accepts it, together with the annexes it was
# matched against.
directive_subsets <- bind_rows(
  if (!is.null(subsets[["kosovo_birds_annex_I"]]))
    sf::st_drop_geometry(subsets[["kosovo_birds_annex_I"]]) |>
      mutate(Directive = "Birds"),
  if (!is.null(subsets[["kosovo_habitats_directive"]]))
    sf::st_drop_geometry(subsets[["kosovo_habitats_directive"]]) |>
      mutate(Directive = "Habitats")
)

directive_tbl <- directive_subsets |>
  filter(!is.na(species), nzchar(species)) |>
  mutate(year = suppressWarnings(as.integer(year))) |>
  group_by(`Scientific name` = species, Directive) |>
  summarise(
    `Common name` = {
      v <- vernacularName[!is.na(vernacularName)]
      if (length(v)) v[1] else "&mdash;"
    },
    Annex = paste(sort(unique(unlist(strsplit(na.omit(annex), "; ")))),
                  collapse = ", "),
    Records = n(),
    `Most recent year` = {
      y <- year[!is.na(year) & year > 1500]
      if (length(y)) max(y) else NA_integer_
    },
    .groups = "drop"
  ) |>
  select(`Scientific name`, `Common name`, Directive, Annex, Records,
         `Most recent year`) |>
  arrange(desc(Records), `Scientific name`)

report_table(
  directive_tbl,
  caption = htmltools::tags$caption(
    style = "caption-side: bottom; text-align: left; padding-top: 8px;
             font-size: 0.85rem; color: #6b7880;",
    sprintf(paste0("Table 3. Species of Community interest recorded in Kosovo: ",
                   "%d taxa across %s quality-checked records. The annex column ",
                   "shows every annex a species was matched against, so a ",
                   "species listed on more than one appears once."),
            nrow(directive_tbl), fmt_int(sum(directive_tbl$Records)))
  ),
  page_length = 15,
  escape = FALSE
) |>
  DT::formatCurrency("Records", currency = "", interval = 3, mark = ",",
                     digits = 0)
```

::: {.callout-warning collapse="true"}
## How annex membership was determined

The species lists are built by `R/build_directive_list.R`, which parses the
consolidated directives on EUR-Lex — the legally binding texts — rather than
relying on a secondary checklist. Every name is then resolved against the GBIF
backbone taxonomy, and occurrence records are selected by numeric key rather
than by name, which is robust to synonymy and authorship variation.

Two details materially affect what appears in these subsets.

**Rank matters when matching.** A taxon listed at species rank is matched on
both its species key and its taxon key, so records identified to subspecies are
still captured. A taxon listed at *subspecies* rank is matched on that
subspecies alone. Promoting subspecies listings to their whole species would be
seriously misleading: the Birds Directive lists island endemics such as
*Columba palumbus azorica* and *Fringilla coelebs ombriosa*, and matching those
at species level would place every Wood Pigeon and Chaffinch in Kosovo on
Annex I — around 1,450 records for subspecies confined to the Azores and the
Canaries.

**Exclusions are honoured.** Where an annex lists a species *except* certain
subspecies, as Annex V does for *Rupicapra rupicapra*, the excluded taxa are
discarded rather than being read as listings in their own right.

Around 1.3 per cent of annex entries could not be resolved to a current GBIF
taxon. These are names superseded since the directives were adopted, mostly
Iberian and Macaronesian plants; none are species expected in Kosovo. The
`directive` and `annex` columns are retained in every export so that any
selection can be checked.
:::

## Download the data

All five extracts are published as **GeoPackage** (`.gpkg`, ready for QGIS or
ArcGIS, geometry included) and **CSV** (flat table with coordinate columns).
Both are open formats requiring no proprietary software.

```{r}
#| label: downloads
#| output: asis

mf <- meta$manifest

cards <- vapply(seq_len(nrow(mf)), function(i) {

  gpkg_btn <- if (!is.na(mf$gpkg[i])) sprintf(
    "<a class='btn btn-primary' href='data_exports/%s' download>GeoPackage (%s)</a>",
    mf$gpkg[i], fmt_bytes(mf$gpkg_size[i])) else ""

  csv_btn <- if (!is.na(mf$csv[i])) sprintf(
    "<a class='btn btn-outline-secondary' href='data_exports/%s' download>CSV (%s)</a>",
    mf$csv[i], fmt_bytes(mf$csv_size[i])) else ""

  sprintf(paste0("<div class='download-card'><h4>%s</h4>",
                 "<div class='meta'>%s records &middot; %s species</div>%s%s</div>"),
          mf$label[i], fmt_int(mf$records[i]), fmt_int(mf$species[i]),
          gpkg_btn, csv_btn)
}, character(1))

cat("<div class='download-grid'>", paste(cards, collapse = "\n"), "</div>", sep = "\n")
```

::: {.callout-tip}
## Conditions of use

GBIF-mediated data are made available under the licences chosen by each
publishing institution; the `license` column in every export records the terms
applying to each record. When these data are used in a report or publication,
please cite the DOI given above, which credits all contributing institutions.
:::

## Methods and reproducibility

The entire analysis is scripted and version-controlled. To reproduce it:

1. Install the R packages listed below.
2. Add your GBIF credentials to `.Renviron` (see the repository `README.md`).
3. Run `Rscript pipeline.R` to acquire, clean and export the data.
4. Run `quarto render` to rebuild this website.

`pipeline.R` records its GBIF download key, so re-running it re-uses the
existing DOI rather than minting a new one. Delete
`data/gbif_download/download_key.txt` to force a fresh extract.

```{r}
#| label: session
#| output: asis

pkgs <- c("rgbif", "sf", "CoordinateCleaner", "leaflet", "leaflet.extras",
          "dplyr", "readr", "DT", "rnaturalearth")
versions <- vapply(pkgs, function(p) {
  tryCatch(as.character(utils::packageVersion(p)), error = function(e) "not installed")
}, character(1))

cat("<details><summary>Software environment</summary>\n\n",
    sprintf("- %s\n", R.version.string),
    sprintf("- Analysis run: %s\n", format(meta$run_date, "%d %B %Y, %H:%M")),
    "\n",
    paste(sprintf("- `%s` %s\n", pkgs, versions), collapse = ""),
    "\n</details>\n", sep = "")
```
