---
title: "Towards Natura 2000"
subtitle: "Answers to the Ministry's questions from the GIS and biodiversity-data sessions of the TAIEX Expert Mission, Pristina, 28–30 September 2026"
date: today
date-format: "D MMMM YYYY"
lang: en-GB
---

```{r}
#| label: setup

suppressPackageStartupMessages({
  library(dplyr)
  library(sf)
})

source("R/functions.R")

meta <- readRDS("data/run_metadata.rds")
run_year <- as.integer(format(as.Date(meta$run_date), "%Y"))

read_export <- function(name) {
  read.csv(file.path("data_exports", paste0(name, ".csv")), encoding = "UTF-8")
}

overall  <- read_export("kosovo_overall_biodiversity")
birds    <- read_export("kosovo_birds")
birds_i  <- read_export("kosovo_birds_annex_I")
hd_ii    <- read_export("kosovo_habitats_annex_II")
annexes  <- read.csv("data/eu_directives_species.csv", encoding = "UTF-8")
ibas     <- read.csv("data/kosovo_ibas.csv", encoding = "UTF-8")
iba_spp  <- read.csv("data/kosovo_iba_species.csv", encoding = "UTF-8")

# The register as the pipeline published it, in a metric CRS for buffering.
register <- st_read("data/kosovo_protected_areas.gpkg", quiet = TRUE) |>
  st_transform(3035)

# The two pilot sites, by their register identifiers.
pilot_ids <- c(sharri = "NP_001", hencit = "SAC_001")
pilot_slug <- c(sharri = "parku-kombetar-sharri",
                hencit = "ligatina-e-hencit-radeves")

pct <- function(x, n, digits = 0) {
  formatC(100 * x / n, format = "f", digits = digits)
}

# eBird Observation Dataset, as GBIF indexes it.
ebird_key <- "4fa7b334-ce0d-4e88-aaae-2e0c138d049e"

as_points <- function(d) {
  st_as_sf(d, coords = c("decimalLongitude", "decimalLatitude"), crs = 4326) |>
    st_transform(3035)
}
```

The Ministry put a set of questions to the expert team about moving from
Kosovo's present biodiversity data towards a Natura 2000 network. This page
answers them as far as they fall within five sessions of the mission: spatial
analysis, condition evaluation, GBIF data, field data collection and automated
reporting. How sites are proposed, how Article 17 reporting works and how the
institutions divide the work are the subject of the sessions by Karel Chobot
and Martin Koška, and are not answered here. The numbering follows the
Ministry's list; question 7 lies outside these sessions.

Each answer gives the recommendation first, then what Kosovo's own data show
about it. Those figures are computed from the same GBIF extract, protected-area
register and pipeline as the [main report](report.qmd), so they move with it.

| Question | Answered here | Discussed in |
|---|---|---|
| [1. Data-quality thresholds](#q1) | Three evidence tiers, computed per record | GBIF data (Day 2) · Field data collection (Day 3) |
| [2. From grid to site boundary](#q2) | A seven-step GIS workflow and a test that catches grid-shaped sites | Spatial analysis (Day 1) · Automated reporting (Day 3) |
| [3. Annex I habitat mapping](#q3) | A minimum mapping standard; what typical species from GBIF can and cannot do | Condition evaluation (Day 2) · Field data collection (Day 3) |
| [4. Sufficiency without national totals](#q4) | Interim denominators, and the EU's own conclusion codes | Spatial analysis (Day 1) |
| [5. The national database](#q5) | Identifiers, provenance, duplicates, versions, sensitive data | Spatial analysis (Day 1) |
| [6. Bird data and SPAs](#q6) | One bird table, the evidence an SPA case needs, Kosovo's IBAs | GBIF data (Day 2) |
| [8. A pilot](#q8) | Sharri National Park and Ligatina e Hencit/Radevës, end to end | Automated reporting (Day 3) · Spatial analysis (Day 1) |

## 1. Which data are good enough for which decision {#q1}

**Treat fitness for use as a property of each record, computed by a written
rule, in three tiers:** national screening, candidate-site delineation and the
Standard Data Form (SDF). A record moves up a tier by gaining precision,
currency, documentation and validation, never because the decision it is
wanted for is more important.

| | National screening | Candidate-site delineation | Standard Data Form |
|---|---|---|---|
| **Spatial precision** | Within the screening cell; 10 km atlas records for 10 km screening only | Points to 100 m or better; habitat polygons mapped at 1:10,000 | As for delineation, and tied to a survey unit: plot, transect, polygon |
| **Currency** | Any date; records before 2000 flagged as historical | Observed within the last 12 years, two six-year reporting cycles | From a standardised survey, ideally within the current reporting cycle |
| **Documentation** | A citable source (dataset DOI or reference); name resolved against a taxonomic backbone | + observer, date and method; a photograph or specimen for difficult taxa | + survey protocol, effort, and counts with their units |
| **Validation** | Automated: coordinates, country, outliers, name matching | Verified by a named specialist for the taxonomic group | + checked at site level, and given an SDF data-quality code: G, M, P or VP |
| **Can support** | Where to look; gap analysis; survey planning | Drawing and justifying a boundary | The population, area and assessment fields of the SDF |

The values are proposals for national agreement, not established standards.
Keep them as rows in a table rather than sentences in a guideline, so that
moving a threshold is a one-line change anyone can review, and the tier of
every record can be recomputed when it moves.

Two rules hold at every tier. A missing record is not evidence of absence; only
surveys that record effort and non-detections can support one. And a name
resolved by a backbone can still be the wrong taxon: synonymy in a global
backbone can fold a Balkan endemic into a widespread species, or a domestic
animal into its wild ancestor. A record whose name changed during matching is
one for a specialist to look at.

### What Kosovo's data show

```{r}
#| label: q1-tiers

tier_counts <- function(d, label) {
  u <- d$coordinateUncertaintyInMeters
  precise <- !is.na(u) & u <= 100
  current <- precise & !is.na(d$year) & d$year >= run_year - 12
  documented <- current & !is.na(d$recordedBy) & trimws(d$recordedBy) != ""
  data.frame(
    Records = label,
    All = fmt_int(nrow(d)),
    `Stated uncertainty ≤ 100 m` = fmt_int(sum(precise)),
    `… and observed since` = fmt_int(sum(current)),
    `… and observer named` = sprintf("%s (%s %%)", fmt_int(sum(documented)),
                                     pct(sum(documented), nrow(d), 1)),
    check.names = FALSE
  )
}

tiers <- rbind(
  tier_counts(overall, "All records"),
  tier_counts(birds_i, "Birds Directive, Annex I"),
  tier_counts(hd_ii,   "Habitats Directive, Annex II")
)
names(tiers)[4] <- paste("… and observed since", run_year - 12)
knitr::kable(tiers, align = "lrrrr")

u <- overall$coordinateUncertaintyInMeters
n_delineation <- sum(!is.na(u) & u <= 100 & !is.na(overall$year) &
                       overall$year >= run_year - 12 &
                       !is.na(overall$recordedBy) & trimws(overall$recordedBy) != "")
```

By these rules, `r pct(n_delineation, nrow(overall), 1)` % of GBIF records
could support a boundary, and even those still need a specialist's
verification. No GBIF record reaches the SDF tier on its own: the extract
carries neither a survey protocol nor a measure of effort, and that tier needs
both.

The Annex I bird row shows why the tiers matter.
**`r pct(sum(birds_i$datasetKey == ebird_key), nrow(birds_i))` % of Kosovo's
Annex I bird records come from eBird**, which publishes the checklist's location
without a stated uncertainty. For a stationary count at a wetland that point is
close to the birds; for a travelling checklist they may be kilometres away. The
rule cannot tell the two apart, so it places both at the screening tier until
the effort fields are read from eBird's own dataset
([question 6](#q6)).

For Natura 2000 purposes GBIF is a screening layer, and a good one: it says
where the evidence is and where it is missing. The
[precision profile](report.qmd#data-quality-control) in the main report gives
the full distribution.

*Discussed in:* GBIF data (Day 2), "How precisely are the records placed?"; Field
data collection (Day 3), "What a form enforces at the point of capture". A
15 m accuracy constraint on the device makes a new record delineation-grade at
the moment it is captured.

## 2. From the screening grid to a site boundary {#q2}

**The grid says where to look; the boundary comes from the ground.** Use the
5 × 5 km cells to rank where evidence concentrates, then discard the cell edges
entirely. Draw the candidate boundary from ecological and physical units at
1:10,000, and record the reason for each stretch of it.

1. **Screen.** Score the cells from screening-tier evidence: listed features,
   records, survey effort. Keep the evidence score and the effort count in
   separate columns, so that a cell that is merely well visited is not mistaken
   for one that is rich.
2. **Form search areas.** Merge adjacent priority cells and buffer them
   generously. This is an analysis envelope, never a draft boundary; from here
   on, cell edges play no part.
3. **Assemble the ecological layers inside it.** Annex I habitat polygons from
   field mapping ([question 3](#q3)); the habitat around delineation-grade
   species records, which means breeding and feeding areas and not just the
   point; catchments and a terrain model; Copernicus land-cover layers; roads
   and settlements as barriers.
4. **Draw along those units.** Use a habitat edge, a watershed, a forest-stand
   edge or a river. Snap the line to features visible on the orthophoto or the
   cadastre at 1:10,000, so that it can be found on the ground and in the land
   register.
5. **Record why.** Give every boundary segment a `boundary_basis` attribute:
   habitat edge, watershed, road, cadastral parcel, expert judgement. A
   boundary whose reasons are written down can be defended; one whose reasons
   are not will be reopened at every consultation.
6. **Test it automatically.** Flag any candidate whose boundary runs along grid
   lines for more than a small share of its length (10 % is a reasonable
   start). Flag any long straight segment that follows no mapped feature. A
   site shaped like its screening cells fails the test by construction.
7. **Check what the line leaves out.** Run the site evidence report on the
   candidate polygon and a 1 km ring around it. Records just outside a boundary
   are the best guide to what it has cut off.

Keep the grid for what it does well: screening, planning surveys and, later,
the distribution maps of Article 17-style reporting. Store every record at its
own precision and derive the grid from it, never the reverse. The one exception
is an atlas record, whose true precision is the cell.

### What Kosovo's data show

```{r}
#| label: q2-ring

hencit <- register[register$national_id == pilot_ids[["hencit"]], ]
bi_pts <- as_points(birds_i)
inside <- lengths(st_intersects(bi_pts, hencit)) > 0
ring   <- st_difference(st_buffer(st_geometry(hencit), 5000), st_geometry(hencit))
in_ring <- lengths(st_intersects(bi_pts, ring)) > 0

sp_inside <- unique(bi_pts$speciesKey[inside])
sp_ring   <- unique(bi_pts$speciesKey[in_ring])
n_ring_only <- length(setdiff(sp_ring, sp_inside))

# The richest Annex I locality in the ring, and how far it lies from the site.
ring_loc <- bi_pts[in_ring, ] |>
  mutate(cell = paste(round(st_coordinates(geometry)[, 1], -2),
                      round(st_coordinates(geometry)[, 2], -2))) |>
  group_by(cell) |>
  summarise(species = n_distinct(speciesKey),
            locality = names(sort(table(locality), decreasing = TRUE))[1],
            .groups = "drop") |>
  arrange(desc(species)) |>
  slice(1)
ring_loc_km <- round(as.numeric(st_distance(ring_loc, hencit)) / 1000, 1)
```

Ligatina e Hencit/Radevës is the country's richest site for Annex I birds, and
its designated polygon covers
`r formatC(as.numeric(st_area(hencit)) / 1e6, format = "f", digits = 1)` km².
Inside it, GBIF holds `r fmt_int(sum(inside))` Annex I bird records of
`r length(sp_inside)` species. Within 5 km outside it lie a further
**`r fmt_int(sum(in_ring))` records of `r length(sp_ring)` species**, about
`r pct(sum(in_ring), sum(inside) + sum(in_ring))` % of all the Annex I bird
records in and around the site. `r n_ring_only` of those species were never
recorded inside. The richest locality in that ring, "`r ring_loc$locality`", lies
`r ring_loc_km` km from the boundary and holds `r ring_loc$species` Annex I
species.

The birds use more of the landscape than the designated wetland. An SPA
boundary copied from the national designation, or from a screening cell, would
leave that out. The ring in step 7 is what shows it.

*Discussed in:* Spatial analysis (Day 1), "Fragmentation: connection, not just
cover"; Automated reporting (Day 3), "A pilot, end to end". Karel Chobot's case
study (Day 1) covers how the Czech sites were proposed.

## 3. Mapping Annex I habitats before assessing them {#q3}

**Several Annex III criteria need both an area and a condition, and only a
polygon map carries both.** These include representativity, relative surface,
and the degree of conservation of structure and function. Points, localities
and centroids say where to map first; they cannot be summed into an area. Map
in the field at 1:10,000, record condition as attributes scored by a written
rule, and validate a random sample independently before any figure goes into
an assessment.

| Element | Recommended minimum | Czech practice, for comparison |
|---|---|---|
| Base and scale | Field mapping on a colour orthophoto, drawn at 1:10,000 (1:5,000 where detail requires) | Orthophoto at 1:10,000 or 1:5,000 |
| Minimum mapping unit | About 0.25 ha for habitats that normally form large stands; smaller for rare or high-value stands | 0.15–0.25 ha recommended for extensive types; smaller only for rarer or high-quality stands |
| Small and linear habitats | Springs, fens, small wetlands and caves as points or lines, with the true area or width recorded | Points and lines for occurrences under about 2,500 m² or narrower than 50 m, with true area or width recorded |
| Mosaics | One segment may hold several habitats, each with its share (%) and its own condition | Mosaic segments; more than five components referred to the methodologist |
| Boundary changes | Recorded when a boundary moves by more than about 30 m | About 30 m |
| Attributes per habitat | Annex I and national code; share of the segment; representativity; degradation; structure and function; typical species present; pressures; surveyor; date; photograph | Share and area per habitat; representativity, degradation, structure and function, regional context; surveyor and date per segment |
| Condition score | Derived from the attributes by a published lookup table, not judged afresh for each site | A quality lookup table, amended when a new combination occurs |
| Validation | Independent re-survey of a random 5–10 % of segments, reported as agreement on the habitat code and on the condition class | — |

Czech practice follows the AOPK ČR methodology for updating the national
habitat-mapping layer (Lustyk & Guth 2011). The recommended column is written
for a first national map; a country mapping for the first time can hold to the
same standard from the start and avoid retrofitting it.

### Typical species from GBIF

**Yes, to decide where to map first, but not for what is reported.** The
*Interpretation Manual of European Union Habitats* lists characteristic species
for each Annex I type. Where several of them are recorded together in one 1 km
cell, that cell is a candidate for early field mapping. The method works where
the typical species are conspicuous and well recorded: aquatic and riparian
types, and dry grasslands with their orchids and butterflies. It works poorly
for forests defined by their dominant trees, and for habitats whose indicators
are bryophytes or small sedges. It cannot give an area, a condition or an SDF
figure. Co-occurrence is not a diagnosis either: most typical species also
occur outside the habitat they characterise.

```{r}
#| label: q3-plants

plants <- overall[overall$kingdom == "Plantae", ]
sharri <- register[register$national_id == pilot_ids[["sharri"]], ]
plants_sharri <- sum(lengths(st_intersects(as_points(plants), sharri)) > 0)
plants_precise <- sum(!is.na(plants$coordinateUncertaintyInMeters) &
                        plants$coordinateUncertaintyInMeters <= 100)
```

In Kosovo the plant record limits it further. Plants make up
`r pct(nrow(plants), nrow(overall))` % of GBIF records: `r fmt_int(nrow(plants))`
records of `r fmt_int(length(unique(plants$speciesKey)))` species, of which
`r fmt_int(plants_precise)` state an uncertainty of 100 m or better.
`r fmt_int(plants_sharri)` lie in Sharri National Park, which covers more than
500 km². Two or three habitat types chosen for well-recorded indicators are a
fair test of the method; a national screen would mostly show where botanists
have been.

*Discussed in:* Condition evaluation (Day 2), "A first habitat map: the minimum
standard"; Field data collection (Day 3), "When the map is the working
document" and "Where to start on Monday". Karel Chobot's habitat session
(Day 1) covers monitoring design.

## 4. Judging sufficiency without national totals {#q4}

**Do not wait for the denominator; make its uncertainty explicit.** Estimate
each national total as a range, compute the network's share at both ends, and
record a conclusion for every feature in the categories the EU's own
sufficiency process uses. Where the honest answer is that more research is
needed, that process has a category for it.

| Interim approach | How | Suits |
|---|---|---|
| **The total as a range** | Lower bound: the mapped or known area or population. Upper bound: modelled potential from land cover, terrain, geology and known records. Compute the network's share against both. | Habitats with partial mapping |
| **Occupied cells, allowing for effort** | Share of occupied grid cells (the EEA 10 km grid) that fall in the network, counting only cells actually surveyed for that group | Species with scattered records |
| **Representation across the range** | Every main part of the range covered: massif, catchment, altitude belt | Every feature, whatever the data |
| **All significant known localities** | Where a feature is known from only a few sites, include every significant one | Rare and priority features |
| **The neighbours' figures** | Article 17 estimates of neighbouring Member States and Emerald Network data of neighbouring Bern Convention Parties, as a check on plausibility | Transboundary features |

The Commission's working reference values are often quoted as a pass mark. A
network holding more than 60 % of a feature's resource is generally taken as
sufficient, and one holding 20 % or less as a priority for scrutiny. They were
designed as a pre-selection filter, and the current criteria say they "can
never be used in isolation". With a denominator given as a range, judge rare and
priority features on the smaller of the two shares.

The conclusion codes of the EU sufficiency process are an honest way to record
an interim result:

| Code | Meaning |
|---|---|
| SUF | Sufficient: no further sites needed |
| IN MIN | Insufficient, minor: the feature should be added to the SDFs of sites already proposed |
| IN MOD | Insufficient, moderate: one or more further sites or extensions needed |
| IN MAJ | Insufficient, major: no site for the feature has been proposed yet |
| SR | Scientific reserve: further research needed to identify the most appropriate sites |
| SR Ref List | Scientific reserve on the reference list: regular occurrence still to be confirmed |
| CD | Correction of data: the SDF information needs correcting or completing |

Keep one row per feature in a *sufficiency ledger*. Each row records the
denominator range and its method, a data-quality code, the network's share, the
conclusion code, its date, and **what evidence would change it**. That last
column turns every SR into a survey commission with a deadline, and makes the
interim state as transparent as the final one.

*Discussed in:* Spatial analysis (Day 1), "A missing baseline is not a reason
for inaction". The biogeographical process itself is part of Karel Chobot's
sessions.

## 5. Records that survive being shared {#q5}

**This page answers the data side of the question.** Every record needs a
permanent identifier, a traceable source, a version history and a validation
status. Duplicates are linked, not deleted. Sensitivity is handled by
generalising the public view, not by hiding the record. Ownership and the
arrangement between the institutions are the subject of the joint
database-architecture session (Day 1) and Martin Koška's governance session
(Day 2).

| Element | Recommendation |
|---|---|
| **Identifiers** | A permanent UUID for every record, survey event and dataset, never re-used, with the provider's own identifier kept alongside (Darwin Core `occurrenceID`). A site keeps its code through every boundary version. |
| **Provenance** | Dataset, source document, observer and method, stored as fields rather than a note. |
| **Duplicates** | Link, do not delete. Records of the same taxon, date and place from different sources share a cluster identifier; analyses count the cluster once and cite every source. |
| **Versions** | Every change to a record logged: who, when, what. Every analysis and every SDF submission made from a dated snapshot with its own DOI. Boundaries and SDFs released as numbered versions. |
| **Validation** | A status on every record, advanced in order: automated checks at import; the taxon specialist for the group; sign-off at site level; approval for SDF use. |
| **Sensitive data** | One store at full precision. The public view generalises listed sensitive species; named users under a data-use agreement see full precision, and access is logged. The sensitive list has an owner, a review cycle, and species can come off it. |
| **Interoperability** | Darwin Core for occurrences; INSPIRE *Protected sites*, *Species distribution* and *Habitats and biotopes* for services, through the national geoportal; one taxonomic backbone and shared code lists. |

```{r}
#| label: q5-duplicates

dups <- overall |>
  filter(!is.na(speciesKey), nchar(eventDate) >= 10) |>
  mutate(date = substr(eventDate, 1, 10),
         lat = round(decimalLatitude, 3), lon = round(decimalLongitude, 3)) |>
  group_by(speciesKey, date, lat, lon) |>
  filter(n_distinct(datasetKey) > 1) |>
  ungroup()
n_clusters <- nrow(distinct(dups, speciesKey, date, lat, lon))
```

Duplication is already present in the published record. **`r n_clusters` groups
of records, `r nrow(dups)` records in all, describe the same species on the
same day at the same place** (to about 100 m) and were published in different
datasets. Counted as they stand, they inflate the evidence; deleted, they lose a
source. Linked, both counts are right. GBIF performs the same kind of linking
across its whole index, and a national system can take its clusters as a
starting point.

*Discussed in:* Spatial analysis (Day 1), "One central Biodiversity Information
System" and "NDOP: open by default, precise by permission".

## 6. Bird data for Special Protection Areas {#q6}

**Keep one table of bird records and derive the layers from it.** Annex I
species (Article 4(1)) and regularly occurring migratory species (Article 4(2))
belong in attributes, not in separate datasets that drift apart. To move from
IBA, KBA and occurrence data to an SPA proposal, each species that is to
qualify the site needs three things. It needs a current population estimate
from standardised counts, and a stated criterion that estimate meets. It also
needs a boundary drawn around the habitat the population actually uses.

| Field | Content | Why |
|---|---|---|
| Annex I | Yes or no, from the consolidated Directive | The Article 4(1) trigger |
| Regular migrant | Yes or no, from a national list the Ministry adopts and versions | The Article 4(2) trigger; the list is a decision, not a lookup |
| Population type | `p` permanent, `r` reproducing, `c` concentration, `w` wintering: the SDF codes | One species can need different sites in different seasons |
| Breeding evidence | Possible, probable or confirmed, as in the European breeding bird atlas | A record in the breeding season is not a breeding record |
| Count and unit | A number, with its unit: pairs, individuals, males, calling males | Only counts with units can be summed or set against a threshold |
| Method and effort | Protocol, duration, distance, whether all species were reported | Turns a record into a survey, and makes absence usable |

The working layers are views over that one table: Annex I breeding, Annex I
passage and wintering, regular migrants, and congregations. When a record is
corrected, every layer follows.

**What an SPA proposal needs, beyond the IBA:**

- **The IBA inventory is the starting point.** The Court of Justice has treated
  it as the best available scientific reference for judging whether enough SPAs
  have been classified (Case C-3/96, *Commission v Netherlands*). An IBA is
  still not an SPA proposal: its boundary is often coarse, and some of its
  qualifying species cannot trigger an SPA at all (see below).
- **A population estimate for each trigger species**, given as minimum and
  maximum with a unit, a year and an SDF data-quality code. It should come from
  standardised surveys over at least two or three seasons:
  - territory mapping or point counts for breeding birds;
  - nest surveys for raptors;
  - International Waterbird Census counts for wintering waterbirds;
  - standardised counts at passage bottlenecks.
- **The criterion it meets.** BirdLife's EU-level criteria are tested
  definitions: C2, the site regularly holds at least 1 % of a flyway or EU
  population of a species threatened in the EU; C6, the site is one of the five
  most important in its region for such a species.
- **A boundary around what the birds use**: nesting, feeding and roosting areas,
  including the fields or river they commute to. Check it against the IBA
  outline; do not copy it.
- **eBird's own dataset for analysis.** Request the eBird Basic Dataset from the
  Cornell Lab of Ornithology. It carries the effort fields and whether each
  checklist reports every species detected, which turns a checklist's missing
  species into non-detections. The copy on GBIF is presence-only.

### Kosovo's IBAs and KBAs

```{r}
#| label: q6-ibas

n_qual <- table(iba_spp$iba_code)
iba_tab <- ibas |>
  mutate(
    Site = sprintf("[%s](%s) (%s)", name, factsheet, code),
    `Qualifying populations` = as.integer(n_qual[code]),
    Area = sprintf("%s km²", fmt_int(area_km2)),
    KBA = ifelse(kbaClassification == "-", "no", tolower(kbaClassification))
  ) |>
  select(Site, `Where it lies` = relation, Area,
         `IBA criteria` = ibaCriteria, KBA, `Qualifying populations`)
knitr::kable(iba_tab, align = "llrlll")
```

BirdLife's inventory lists these four sites for Kosovo. It holds them under
Serbia, where the inventory was compiled; BirdLife has no separate country
record for Kosovo. The digital boundaries are released on request through the
[World Database of Key Biodiversity Areas](https://www.keybiodiversityareas.org/),
free for non-commercial use, so this page uses each site's published centroid
only. How much of Kopaonik and Mokra Gora lies inside Kosovo can be settled only
from those boundaries, and requesting them is a first, cheap step.

```{r}
#| label: q6-species

annex_i_keys <- annexes$gbif_species_key[annexes$directive == "Birds" &
                                         annexes$annex == "I"]
key_of <- function(name) {
  k <- birds$speciesKey[birds$species == name]
  if (length(k)) k[1] else NA
}
qual <- iba_spp |>
  group_by(scientific_name, common_name) |>
  summarise(
    IBAs = paste(sprintf("%s: %s %s", iba_name, population,
                         sub("breeding ", "", units)), collapse = "; "),
    season = paste(sort(unique(season)), collapse = ", "),
    .groups = "drop"
  ) |>
  rowwise() |>
  mutate(
    key = key_of(scientific_name),
    in_annex_i = scientific_name %in% annexes$scientific_name[
                   annexes$directive == "Birds" & annexes$annex == "I"] |
                 (!is.na(key) && key %in% annex_i_keys),
    records = sum(birds$species == scientific_name)
  ) |>
  ungroup() |>
  mutate(Route = case_when(
    in_annex_i ~ "Annex I: Art. 4(1)",
    grepl("breeding|passage|winter|non-breeding", season) ~
      "Art. 4(2), if listed as a regular migrant",
    TRUE ~ "Cannot trigger an SPA"
  )) |>
  arrange(Route, scientific_name)

n_no_trigger <- sum(qual$Route == "Cannot trigger an SPA")
knitr::kable(
  qual |> transmute(Species = sprintf("*%s*, %s", scientific_name, common_name),
                    `IBA populations (pairs)` = IBAs, Season = season, Route,
                    `GBIF records in Kosovo` = fmt_int(records)),
  align = "lllr"
)
```

Every qualifying population in these sites is a mountain or forest bird,
estimated in breeding pairs. **`r n_no_trigger` of the
`r nrow(qual)` qualifying species are neither on Annex I nor migratory**:
biome-restricted residents such as the Wallcreeper. They qualify an IBA but
cannot trigger an SPA. The Annex I species need current counts; several
estimates date from 2008–2013.

### The best bird site in the data is not an IBA

```{r}
#| label: q6-seasons

season_of <- function(m) {
  case_when(m %in% 4:7 ~ "Breeding season (Apr–Jul)",
            m %in% 8:10 ~ "Autumn passage (Aug–Oct)",
            m %in% c(11, 12, 1, 2) ~ "Winter (Nov–Feb)",
            m == 3 ~ "Spring passage (Mar)",
            TRUE ~ "Month not given")
}
birds_i$season <- season_of(birds_i$month)
birds_i$at_hencit <- inside
season_tab <- birds_i |>
  group_by(Season = season) |>
  summarise(`Records, Kosovo` = fmt_int(n()),
            `Species, Kosovo` = n_distinct(speciesKey),
            `Records, Ligatina e Hencit` = fmt_int(sum(at_hencit)),
            `Species, Ligatina e Hencit` = n_distinct(speciesKey[at_hencit]),
            .groups = "drop") |>
  arrange(match(Season, c("Breeding season (Apr–Jul)", "Autumn passage (Aug–Oct)",
                          "Winter (Nov–Feb)", "Spring passage (Mar)",
                          "Month not given")))

hencit_never_breeding <- birds_i |>
  filter(at_hencit) |>
  group_by(speciesKey) |>
  summarise(br = any(season == "Breeding season (Apr–Jul)")) |>
  summarise(n = sum(!br)) |>
  pull(n)
```

Ligatina e Hencit/Radevës holds `r length(sp_inside)` of the
`r length(unique(birds_i$speciesKey))` Annex I bird species recorded in Kosovo.
It is designated nationally as a wetland and appears in no IBA inventory. It is
therefore the case for the occurrence-data route to an SPA, and it is where
standardised counts are most obviously missing. Its birds also differ by
season, which is the population-type field at work.
`r hencit_never_breeding` of its Annex I species were never recorded there in
the breeding season: they are passage or wintering birds, and an SDF would list
them as `c` or `w`.

```{r}
#| label: q6-season-table
knitr::kable(season_tab, align = "lrrrr")
```

The months are a coarse proxy for population type: early migrants arrive in
April, and some wintering birds stay into March. With breeding-evidence codes
on the records, the proxy would not be needed.

*Discussed in:* GBIF data (Day 2), "Birds: from records to an SPA case".

## 8. A pilot, end to end {#q8}

**Yes.** Take two contrasting areas through the whole workflow in one field
season before applying the method nationally: **Sharri National Park** for the
habitat route and **Ligatina e Hencit/Radevës** for the bird route. A pilot
tests the method, the data model and the division of work at a scale where
mistakes are cheap. It also produces the first draft Standard Data Forms along
the way.

```{r}
#| label: q8-pilots

site_stats <- lapply(names(pilot_slug), function(k) {
  gpkg <- file.path("outputs/protected_areas", pilot_slug[[k]],
                    paste0(pilot_slug[[k]], ".gpkg"))
  st_drop_geometry(st_read(gpkg, layer = "site_boundary", quiet = TRUE))
})
names(site_stats) <- names(pilot_slug)

annex_i_species <- function(site) {
  n_distinct(bi_pts$speciesKey[lengths(st_intersects(bi_pts, site)) > 0])
}
sara <- ibas[ibas$code == "RS024", ]

pilot_row <- function(label, sharri, hencit) {
  data.frame(` ` = label, `Sharri National Park` = sharri,
             `Ligatina e Hencit/Radevës` = hencit, check.names = FALSE)
}
s <- site_stats$sharri; h <- site_stats$hencit
pilots <- rbind(
  pilot_row("Register", sprintf("%s · %s, %s", s$national_id, s$designation, s$designation_year),
      sprintf("%s · %s, %s", h$national_id, h$designation, h$designation_year)),
  pilot_row("Area", sprintf("%s km²", formatC(as.numeric(s$computed_area_km2), format = "f", digits = 1)),
      sprintf("%s km²", formatC(as.numeric(h$computed_area_km2), format = "f", digits = 1))),
  pilot_row("GBIF records · species", sprintf("%s · %s", fmt_int(as.integer(s$records)), fmt_int(as.integer(s$species))),
      sprintf("%s · %s", fmt_int(as.integer(h$records)), fmt_int(as.integer(h$species)))),
  pilot_row("Annex I bird species recorded", annex_i_species(sharri), annex_i_species(hencit)),
  pilot_row("Plant records", fmt_int(plants_sharri),
      fmt_int(sum(lengths(st_intersects(as_points(plants), hencit)) > 0))),
  pilot_row("1 km cells with no record", sprintf("%s %% of %s", pct(as.numeric(s$unrecorded_share), 1),
                                            s$grid_cells),
      sprintf("%s %% of %s", pct(as.numeric(h$unrecorded_share), 1), h$grid_cells)),
  pilot_row("IBA / KBA", sprintf("Within %s (%s); BirdLife records the park as contained by it",
                           sara$name, sara$code), "None"),
  pilot_row("Route it tests", "SCI/SAC: Annex I habitats and Annex II species, across a border",
      "SPA: Annex I and migratory birds, from counts"),
  pilot_row("Biggest gap", "A habitat map; most of the park unrecorded",
      "Standardised counts; the land the birds use beyond the boundary")
)
knitr::kable(pilots, align = "lll")
```

| Step | Sharri National Park | Ligatina e Hencit/Radevës | Produces |
|---|---|---|---|
| **1. Database QA** | Evidence tiers applied to every record ([question 1](#q1)) | The same | A tiered record set |
| **2. Gap report** | Site report: unrecorded cells, missing groups | Site report, plus the 5 km ring ([question 2](#q2)) | A survey plan |
| **3. Field verification** | Habitat mapping of a representative sector at 1:10,000; Annex II species surveys | Standardised counts in the breeding season, on passage and in winter | Validated field data |
| **4. Delineation** | From habitat polygons and terrain; IBA boundary requested for comparison | From the area the birds use | A candidate polygon with its `boundary_basis` |
| **5. Assessment** | Annex III stage 1: representativity, relative surface, conservation, global | The SPA criteria, species by species | An assessment table |
| **6. Network review** | The site's share of each feature, given as a range ([question 4](#q4)) | The same, for Annex I species and migrants | Rows in the sufficiency ledger |
| **7. Draft SDF** | With data-quality codes on every value | With population types on every species | Two draft SDFs and a lessons log |

**The lessons log is the main product.** It records every step that took
longer than planned, every attribute nobody could fill in, and every rule that
had to change. The national method is revised from it before roll-out, which is
the point of doing two sites first rather than two hundred.

The evidence packs for step 2 can be produced today, from data already in this
repository:

```bash
Rscript site_reports.R --site NP_001 --site SAC_001
```

*Discussed in:* Automated reporting (Day 3), "A pilot, end to end"; Spatial
analysis (Day 1), "Next steps: governance that attracts funding".

## Sources

- BirdLife International (2026). Important Bird Area factsheets: *Sara
  mountain*, *Prokletije*, *Kopaonik*, *Mokra Gora*. DataZone,
  datazone.birdlife.org, accessed `r fmt_date_en(as.Date(ibas$accessed[1]))`.
  Built into `data/kosovo_ibas.csv` by `R/fetch_iba.R`.
- Court of Justice of the European Union (1998). Case C-3/96, *Commission v
  Kingdom of the Netherlands*.
- European Commission (1997). *Criteria for assessing national lists of pSCI at
  biogeographical level*, Hab. 97/2 rev. 4.
  [eionet.europa.eu](https://www.eionet.europa.eu/etcs/etc-be/activities/hab_97_2_criter_en.pdf)
- European Commission (2011). Implementing Decision 2011/484/EU concerning a
  site information format for Natura 2000 sites.
  [EUR-Lex](https://eur-lex.europa.eu/legal-content/EN/TXT/?uri=CELEX:32011D0484)
- European Commission (2013). *Interpretation Manual of European Union Habitats*,
  EUR28.
- European Topic Centre on Biological Diversity (2016). *Criteria for assessing
  sufficiency of sites designation*, further adapted criteria.
  [eionet.europa.eu](https://www.eionet.europa.eu/etcs/etc-be/activities/further_adapted_criteria.pdf)
- Lustyk, P. & Guth, J. (2011). *Metodika aktualizace vrstvy mapování biotopů*.
  AOPK ČR, Praha.
- GBIF.org. Occurrence download
  [doi:`r meta$download$doi`](https://doi.org/`r meta$download$doi`), the
  extract behind every figure on this page.
