---
title: "NaturAL project, Albania: small terrestrial mammals"
subtitle: "An open data paper from Albania's Natura 2000 preparation project, with fieldwork on the Albanian side of Kosovo's border."
actions:
  - kind: check
    what: "Whether the BioNNA database the paper cites (http://www.bionna.al/) has moved; its host name did not resolve on 29 September 2026"
    contact: "National Agency of Protected Areas, Albania, or the paper's corresponding author"
  - kind: check
    what: "Whether the paper's locality records are on GBIF; a dataset search for the paper found none"
    contact: "https://www.gbif.org/dataset/search?q=Albania"
  - kind: check
    what: "Other NaturAL outputs (habitat mapping, other taxa) and whether they are published as data"
    contact: "Search the NaturAL project's final reports"
todo:
  format: "The paper is open; whether its locality table is machine-readable has not been checked"
  spatial_resolution: "Precision of the trapping localities; read from the paper's locality table"
  latest_version: "2018 publication; check for later NaturAL data releases"
  kosovo_codes_seen: "Not applicable (Albania only); confirm if records near the border are ever merged with Kosovo data"
  tier: "Depends on the locality precision in the paper"
  effort: "Estimate once the locality table's format is known"
  sensitivity: "Small mammals are not usually persecuted; confirm no restricted species are included"
---

```{r}
#| label: setup
source("../R/sources_helpers.R")
row <- source_row("albania-natural")
```

::: {.source-status}
`r status_line(row)`
:::

## At a glance

```{r}
#| label: glance
metadata_table(row)
```

## What it is

NaturAL, "Strengthening capacity in National Nature Protection – preparation
for Natura 2000 network", was Albania's project to prepare its Natura 2000
network. One of its published outputs is a data paper in *ZooKeys* (Bego,
Saçdanaku, Pacifici & Rondinini 2018) on the distribution and diversity of
Albania's small terrestrial mammals. It reports 31 species from field
campaigns in the summers and autumns of 2016 and 2017 in priority protected
areas, and it is licensed CC BY 4.0.

The paper is relevant here because of where the fieldwork took place. It
covers Kukës district (Novosejë-Shishtavec and Korab-Koritnik), Tropojë, Has
(Krumë and Pashtrik) and the Albanian Alps: the Albanian side of the Sharr,
Pashtrik and Bjeshkët e Nemuna/Prokletije massifs that Kosovo shares.

## Access

```{r}
#| label: access
#| echo: true
#| code-fold: show
access_table(list(
  try_access("https://doi.org/10.3897/zookeys.742.22364"),
  try_access("https://pmc.ncbi.nlm.nih.gov/articles/PMC5904422/"),
  try_access("http://www.bionna.al/")
))
```

The paper is open at the publisher and in PubMed Central. The paper says that
"all previous and recent records" of small mammals are held in the BioNNA
database at `http://www.bionna.al/`. That host name did not resolve on
29 September 2026, so the database could not be reached. A GBIF dataset
search found no copy of the paper's records.

## Kosovo coverage check

::: {.callout-warning}
## To be completed
The data describe Albania only, so there is nothing inside Kosovo to count.
The useful check, not yet run, is to take the paper's localities in Kukës,
Tropojë and Has, keep those within 20 km of Kosovo, and compare their species
with the small mammals in the Kosovo GBIF extract. That needs the locality
table read from the paper, or the BioNNA records if the database reappears.
:::

## What it adds for Natura 2000 in Kosovo

Most small mammals on the Habitats Directive annexes are not among the
species here. The value is as a reference list for the same mountains: which
shrews, voles and mice a systematic survey found a few kilometres across the
border, against which the patchy Kosovo record can be judged. It also shows
what a Natura 2000 preparation project in a neighbour chose to survey and
publish, which is a model for Kosovo's own. Tier: to be set once the locality
precision is read ([tiers](../natura2000.qmd#q1)).

## Caveats and sensitivity

- The paper covers one group; other NaturAL outputs were not checked.
- Only the paper was reached; the database behind it was not.
- Records in the paper are Albanian and must not be counted as Kosovo
  records.

## Integration into the pipeline

Low to medium effort. If the locality table is machine-readable, a proposed
`neighbour_reference_records(path, buffer_km = 20)` would read it into a
separate layer beside the GBIF extract, for comparison only. `pipeline.R` is
not changed.

## To be completed

```{r}
#| label: todo
#| output: asis
todo_list(row)
```

## Citation

As the paper gives it: Bego, F., Saçdanaku, E., Pacifici, M. & Rondinini, C.
(2018). Small terrestrial mammals of Albania: distribution and diversity
(Mammalia, Eulipotyphla, Rodentia). *ZooKeys* 742.
[doi:10.3897/zookeys.742.22364](https://doi.org/10.3897/zookeys.742.22364).

::: {.callout-warning}
## To be completed
The page range of the article was not read; take it from the publisher's
page.
:::
