---
title: "Article 12 reporting under the Birds Directive, 2013–2018"
subtitle: "The EU Member States' 10 km bird distribution maps for 2013–2018, read here for the breeding and wintering birds reported near Kosovo."
actions:
  - kind: check
    what: "Whether the EEA has published the 2019–2024 Article 12 spatial dataset; the datahub lists only 2013–2018 and 2008–2012"
    contact: "EEA datahub, https://www.eea.europa.eu/en/datahub/datahubitem-view/96e1b9b1-ee94-4547-ad61-8059df7240bf"
  - kind: check
    what: "Set the border-zone species against the IBA qualifying species and the Annex I birds in the GBIF extract, to see which neighbours' breeding species Kosovo has no record of"
    contact: "data/kosovo_iba_species.csv and data_exports/kosovo_birds_annex_I.csv in this repository"
todo:
  doi: "The catalogue record shows no DOI and DataCite returns none for its identifier; ask the EEA helpdesk"
---

```{r}
#| label: setup
source("../R/sources_helpers.R")
suppressPackageStartupMessages({
  library(sf)
  library(leaflet)
})
row <- source_row("art12-birds")
cache <- file.path(cache_dir("art12-birds"), "art12_border_cells.gpkg")
```

::: {.source-status}
`r status_line(row)`
:::

## At a glance

```{r}
#| label: glance
metadata_table(row)
```

## What it is

Article 12 of the Birds Directive asks every EU Member State to report, every
six years, on the size, trend and distribution of its bird populations. The
EEA publishes the 2013–2018 round as a tabular dataset (population sizes and
short- and long-term trends, breeding and wintering) and a spatial dataset of
10 km grid cells in which each country mapped where each species occurs. The
public spatial version dates from April 2021 and was published on the EEA
datahub on 1 March 2022, under CC-BY 4.0.

The EEA's README describes the spatial file as gridded distribution data "as
reported by EU Member States", aggregated by species and country, and states
that it is the public version without sensitive species.

## Access

Open, with no account. The distribution GeoPackage is read remotely, as on the
[Article 17 page](art17-2013-2018.qmd): GDAL fetches only the part of the file
covering a box around Kosovo, and the 10 km cells within 100 km of the border
are cached here.

```{r}
#| label: access
#| echo: true
#| code-fold: show
landing <- "https://www.eea.europa.eu/en/datahub/datahubitem-view/96e1b9b1-ee94-4547-ad61-8059df7240bf"
share   <- "https://sdi.eea.europa.eu/datashare/s/wcpT9ak6BnyY8kL/download?path=%2FArt12-2013-2018_GPKG&files=ART12_3035_distribution_data_without_sensitive.gpkg"
access_table(list(try_access(landing), try_access(share)))
```

The share token changes when the EEA republishes. The stable address is
`https://sdi.eea.europa.eu/data/e2face16-f352-4aff-9e4f-0ad1306f89b5`.

```{r}
#| label: build
#| echo: true
#| code-fold: true
#| code-summary: "Show how the border-zone extract is built"
build_error <- NULL
if (!file.exists(cache)) {
  build_error <- tryCatch({
    Sys.setenv(GDAL_DISABLE_READDIR_ON_OPEN = "EMPTY_DIR",
               CPL_VSIL_CURL_CHUNK_SIZE = "1048576",
               CPL_VSIL_CURL_CACHE_SIZE = "524288000")
    src  <- paste0("/vsicurl/", share)
    ko   <- st_transform(st_union(kosovo_outline()), 3035)
    zone <- st_buffer(ko, 100000)
    birds <- st_read(src,
      layer = "EU_ART12_birds_distribution_2013_2018_without_sensitive_species",
      wkt_filter = st_as_text(st_as_sfc(st_bbox(zone))), quiet = TRUE)

    # Rebuild the EEA 10 km grid, whose corners sit on round 10 km multiples
    # of EPSG:3035, and test each cell's centre against the dissolved ranges.
    bb <- st_bbox(zone)
    bb[c("xmin", "ymin")] <- floor(bb[c("xmin", "ymin")] / 1e4) * 1e4
    bb[c("xmax", "ymax")] <- ceiling(bb[c("xmax", "ymax")] / 1e4) * 1e4
    grid <- st_sf(geometry = st_make_grid(st_as_sfc(bb), cellsize = 1e4))
    grid <- grid[lengths(st_intersects(grid, zone)) > 0, ]
    xy <- st_coordinates(st_centroid(grid))
    grid$cell_id <- sprintf("10kmE%dN%d", floor(xy[, 1] / 1e4), floor(xy[, 2] / 1e4))
    grid$dist_km <- round(as.numeric(st_distance(grid, ko)) / 1000, 1)

    hit <- st_intersects(st_centroid(st_geometry(grid)), birds)
    a <- st_drop_geometry(birds)
    long <- do.call(rbind, lapply(seq_along(hit), function(i) {
      if (!length(hit[[i]])) return(NULL)
      f <- a[hit[[i]], ]
      data.frame(cell_id = grid$cell_id[i], code = f$speciescodeEU,
                 name = f$speciesnameEU, season = f$season,
                 country = f$country, trend = f$population_trend)
    }))
    grid <- grid[grid$cell_id %in% long$cell_id, ]
    st_write(grid, cache, layer = "cells", delete_dsn = TRUE, quiet = TRUE)
    st_write(long, cache, layer = "features", append = FALSE, quiet = TRUE)
    NULL
  }, error = function(e) conditionMessage(e))
}
have_cache <- file.exists(cache)
```

```{r}
#| label: build-outcome
#| output: asis
if (have_cache) {
  fp <- file_fingerprint(cache)
  cat(sprintf("The extract `%s` (%s, SHA-256 `%s`) was written on %s and is committed with the site.\n",
              fp$file, fp$size, substr(fp$sha256, 1, 16), fp$downloaded))
} else {
  cat("::: {.callout-warning}\n## To be completed\nThe remote read of the EEA GeoPackage failed at render time (",
      build_error %||% "no message", "). Re-render when sdi.eea.europa.eu answers.\n:::\n", sep = "")
}
```

## Kosovo coverage check

Kosovo did not report under the Directive, so no cell lies inside it. What
the file can say is which birds the nearest Member States report within 50 km
and 100 km of the border, by season. For an SPA case that is the regional
context a reviewer would ask for.

```{r}
#| label: check
#| eval: !expr have_cache
cells <- st_read(cache, layer = "cells", quiet = TRUE)
feat  <- st_read(cache, layer = "features", quiet = TRUE)
feat$near <- cells$dist_km[match(feat$cell_id, cells$cell_id)] <= 50
zone_row <- function(f, label) data.frame(
  Zone = label, Cells = length(unique(f$cell_id)),
  Countries = paste(sort(unique(f$country)), collapse = ", "),
  `Breeding species` = length(unique(f$code[f$season == "breeding"])),
  `Wintering species` = length(unique(f$code[f$season == "wintering"])),
  check.names = FALSE)
knitr::kable(rbind(zone_row(feat[feat$near, ], "Within 50 km"),
                   zone_row(feat, "Within 100 km")),
             align = "lrlrr", row.names = FALSE)
```

```{r}
#| label: check-text
#| eval: !expr have_cache
#| output: asis
source("../R/functions.R", local = TRUE)
annex <- read.csv("../data/eu_directives_species.csv", encoding = "UTF-8")
annex_i <- unique(annex$scientific_name[annex$directive == "Birds" & annex$annex == "I"])
codes <- unique(feat[, c("code", "name")])
codes <- codes[!duplicated(codes$code), ]
cat(sprintf(paste(
  "The nearest reported cell lies %s km from Kosovo. Within 100 km the",
  "neighbouring Member States report %d bird species codes, %d of them",
  "matching a name on Annex I of the Birds Directive exactly. The extract",
  "holds %s maps only."),
  format(min(cells$dist_km)), nrow(codes), sum(codes$name %in% annex_i),
  paste(sort(unique(feat$season)), collapse = " and ")))
```

```{r}
#| label: map
#| eval: !expr have_cache
ko <- kosovo_outline()
cells$species <- as.integer(tapply(feat$code, feat$cell_id,
                                   function(x) length(unique(x)))[cells$cell_id])
rings <- st_transform(st_sf(ring = c("50 km", "100 km"),
  geometry = st_buffer(st_transform(st_union(ko), 3035), c(50000, 100000))), 4326)
pal <- colorBin(map_palette$density, cells$species, bins = 5)
leaflet(st_transform(cells, 4326)) |>
  add_basemaps() |>
  addPolygons(fillColor = ~pal(species), fillOpacity = 0.75, weight = 0.5,
              color = "#666666",
              label = ~sprintf("%s: %d species, %s km from Kosovo",
                               cell_id, species, dist_km)) |>
  addPolylines(data = rings, color = "#175CA1", weight = 1.5,
               dashArray = "4 4", label = ~ring) |>
  add_boundary_outline(ko) |>
  addLegend(pal = pal, values = ~species, title = "Bird species<br>per 10 km cell") |>
  addLayersControl(baseGroups = basemap_groups,
                   options = layersControlOptions(collapsed = TRUE))
```

::: {.callout-note collapse="true"}
## Every species reported within 100 km

```{r}
#| label: lists
#| eval: !expr have_cache
lst <- aggregate(cell_id ~ name + season, data = feat,
                 FUN = function(x) length(unique(x)))
lst$annex_i <- ifelse(lst$name %in% annex_i, "yes", "")
lst <- lst[order(lst$season, -lst$cell_id), ]
knitr::kable(data.frame(Species = paste0("*", lst$name, "*"), Season = lst$season,
                        `Annex I` = lst$annex_i, Cells = lst$cell_id,
                        check.names = FALSE),
             align = "llcr", row.names = FALSE)
```

:::

## What it adds for Natura 2000 in Kosovo

The report already has Kosovo's own bird records from GBIF and the IBA
qualifying species on the [Natura 2000 page](../natura2000.qmd#q6). This file
adds the regional frame around them: the Annex I species that the nearest EU
countries hold as breeding populations, and their reported trends. A species
breeding 60 km away that has no Kosovo record is a survey question; one
reported declining there raises the stakes of a Kosovo population. At 10 km
the data serve the [national screening tier](../natura2000.qmd#q1).

## Caveats and sensitivity

- The public version leaves out sensitive species entirely, as the EEA says.
  Raptors, grouse and other persecuted birds may therefore be absent here
  although reported to the Commission.
- The nearest data lie in Bulgaria, because Kosovo's four neighbours are not
  EU members; the Montenegrin, Albanian, Macedonian and Serbian mountains that
  share Kosovo's bird fauna are blank.
- Near Kosovo the file holds breeding distributions only; no wintering map
  reaches this far, and passage is never mapped. The lists therefore say
  nothing about the migrants that give Ligatina e Hencit/Radevës its value.
- Annex I is matched by exact name, so a taxon the EEA names differently
  (for example with "s. str." or "all others") is not flagged even when it is
  listed.

## Integration into the pipeline

Low effort, and the same function as for Article 17 with a different layer:
a proposed `art12_border_species(boundary, radius_km = c(50, 100))` in
`R/functions.R`, run beside the IBA step, whose output a report section on the
SPA route could read. `pipeline.R` is not changed by this page.

## To be completed

```{r}
#| label: todo
#| output: asis
todo_list(row)
```

## Citation

Neither the catalogue record nor the README gives a citation string. Under
the CC-BY 4.0 licence the attribution goes to the copyright holders named
there, DG Environment and the EEA: *Population trend of bird species: datasets
from Article 12, Birds Directive 2009/147/EC reporting (2013–2018), public
version, April 2021.* Catalogue record e2face16-f352-4aff-9e4f-0ad1306f89b5.
