---
title: "Article 17 reporting under the Habitats Directive, 2013–2018"
subtitle: "The EU Member States' own 10 km distribution maps of Annex I habitats and Annex II, IV and V species, read here for the ground just across Kosovo's border."
actions:
  - kind: check
    what: "Whether the EEA has published the 2019–2024 spatial dataset; the Article 17 web tool already offers a 2019–2024 period, the datahub does not"
    contact: "EEA datahub, https://www.eea.europa.eu/en/datahub/datahubitem-view/d8b47719-9213-485a-845b-db1bfe93598d"
  - kind: check
    what: "Rebuild the border-zone lists from the 2019–2024 distribution layer once it is published, and compare the two rounds"
    contact: "Re-run this page's build chunk against the new file"
todo:
  doi: "The catalogue record shows no DOI and DataCite returns none for the record identifier; ask the EEA helpdesk whether one exists"
---

```{r}
#| label: setup
source("../R/sources_helpers.R")
suppressPackageStartupMessages({
  library(sf)
  library(leaflet)
})
row <- source_row("art17-2013-2018")
cache <- file.path(cache_dir("art17-2013-2018"), "art17_border_cells.gpkg")
```

::: {.source-status}
`r status_line(row)`
:::

## At a glance

```{r}
#| label: glance
metadata_table(row)
```

## What it is

Every six years each EU Member State reports, under Article 17 of the Habitats
Directive, on the conservation status of the habitat types and species listed
in the Directive's annexes. The European Environment Agency (EEA) publishes the
reports as two linked products: tabular assessments of status and trend for each
country and biogeographical region, and a spatial layer in which each country
maps where every habitat and species occurs on the 10 km ETRS89 grid. The
public version for the 2013–2018 round was released in August 2020, for the
EU of 28 countries, and is licensed CC-BY 4.0 to the EEA and the Commission's
Directorate-General for Environment.

The EEA describes the spatial part as habitat and species distribution data in
10 km grid cells "as reported by Member States". That wording matters here: the
maps are what each country chose to report, compiled rather than harmonised,
and a country boundary is also the edge of the data.

## Access

The dataset is open and needs no account. The GeoPackage is 151,576,576 bytes
(about 145 MB), so it is not copied into this repository: GDAL reads it over HTTP range requests
(`/vsicurl/`) and fetches only the pages that cover a box around Kosovo, as the
pipeline already does for the protected-area inventory. The result, about a
thousand distribution features, is cut to the 10 km cells within 100 km of
Kosovo and cached as a small GeoPackage.

```{r}
#| label: access
#| echo: true
#| code-fold: show
landing <- "https://www.eea.europa.eu/en/datahub/datahubitem-view/d8b47719-9213-485a-845b-db1bfe93598d"
share   <- "https://sdi.eea.europa.eu/datashare/s/FsDQnwDBiqf2f88/download?path=%2FArt17-2013-2018_GPKG&files=art17_2013_2018_public.gpkg"
access_table(list(try_access(landing), try_access(share)))
```

The share token in that address is not part of the dataset's identifier and
may change when the EEA republishes the file. The stable address is the
catalogue record, `https://sdi.eea.europa.eu/data/9f71b3e3-f8ec-442b-a2d5-c3c190605ac4`,
which redirects to the current share.

```{r}
#| label: build
#| echo: true
#| code-fold: true
#| code-summary: "Show how the border-zone extract is built"
# Runs only when the cache is missing. Any failure is caught and reported below.
build_error <- NULL
if (!file.exists(cache)) {
  build_error <- tryCatch({
    Sys.setenv(GDAL_DISABLE_READDIR_ON_OPEN = "EMPTY_DIR",
               CPL_VSIL_CURL_CHUNK_SIZE = "1048576",
               CPL_VSIL_CURL_CACHE_SIZE = "524288000")
    src <- paste0("/vsicurl/", share)
    ko   <- st_transform(st_union(kosovo_outline()), 3035)
    zone <- st_buffer(ko, 100000)
    box  <- st_as_text(st_as_sfc(st_bbox(zone)))

    sp <- st_read(src, layer = "Art17_species_distribution_2013_2018_MS",
                  wkt_filter = box, quiet = TRUE)
    hb <- st_read(src, layer = "Art17_habitats_distribution_2013_2018_MS",
                  wkt_filter = box, quiet = TRUE)
    sp_names <- unique(st_read(src, layer = "speciesMSassessment",
                               quiet = TRUE)[, c("speciescode", "speciesname")])

    # Each feature is one species (or habitat) in one country and region, its
    # 10 km cells dissolved into one shape. The cells are rebuilt from the EEA
    # grid, whose corners sit on round 10 km multiples of EPSG:3035.
    bb <- st_bbox(zone)
    bb[c("xmin", "ymin")] <- floor(bb[c("xmin", "ymin")] / 1e4) * 1e4
    bb[c("xmax", "ymax")] <- ceiling(bb[c("xmax", "ymax")] / 1e4) * 1e4
    grid <- st_sf(geometry = st_make_grid(st_as_sfc(bb), cellsize = 1e4))
    grid <- grid[lengths(st_intersects(grid, zone)) > 0, ]
    grid$cell_id <- sprintf("10kmE%dN%d",
      floor(st_coordinates(st_centroid(grid))[, 1] / 1e4),
      floor(st_coordinates(st_centroid(grid))[, 2] / 1e4))
    grid$dist_km <- round(as.numeric(st_distance(grid, ko)) / 1000, 1)

    centres <- st_centroid(st_geometry(grid))
    explode <- function(x, code_col, kind) {
      hit <- st_intersects(centres, x)
      do.call(rbind, lapply(seq_along(hit), function(i) {
        if (!length(hit[[i]])) return(NULL)
        f <- st_drop_geometry(x)[hit[[i]], ]
        data.frame(cell_id = grid$cell_id[i], kind = kind,
                   code = as.character(f[[code_col]]), country = f$country,
                   region = f$region,
                   sensitive = if ("sensitive" %in% names(f)) f$sensitive
                               else NA_character_)
      }))
    }
    long <- rbind(explode(sp, "speciescode", "species"),
                  explode(hb, "habitatcode", "habitat"))
    long$name <- sp_names$speciesname[match(long$code, sp_names$speciescode)]
    long$name[long$kind == "habitat"] <- NA

    grid <- grid[grid$cell_id %in% long$cell_id, ]
    st_write(grid, cache, layer = "cells", delete_dsn = TRUE, quiet = TRUE)
    st_write(long, cache, layer = "features", append = FALSE, quiet = TRUE)
    NULL
  }, error = function(e) conditionMessage(e))
}
have_cache <- file.exists(cache)
```

```{r}
#| label: build-outcome
#| output: asis
if (have_cache) {
  fp <- file_fingerprint(cache)
  cat(sprintf("The extract `%s` (%s, SHA-256 `%s`) was written on %s and is committed with the site.\n",
              fp$file, fp$size, substr(fp$sha256, 1, 16), fp$downloaded))
} else {
  cat("::: {.callout-warning}\n## To be completed\nThe remote read of the EEA GeoPackage failed at render time (",
      build_error %||% "no message", "). Re-render when sdi.eea.europa.eu answers; ",
      "the check below is skipped until then.\n:::\n", sep = "")
}
```

## Kosovo coverage check

Kosovo is not an EU Member State and reported nothing, so the dataset has no
cells inside it by design. The useful question is a different one: which habitats and species the neighbouring Member States report
within 50 km and 100 km of the border. Those are the features a future Kosovo
network would share with the EU's own, and the lists a reviewer in the
biogeographical process would expect to see answered.

```{r}
#| label: check
#| eval: !expr have_cache
cells <- st_read(cache, layer = "cells", quiet = TRUE)
feat  <- st_read(cache, layer = "features", quiet = TRUE)
feat$band <- ifelse(cells$dist_km[match(feat$cell_id, cells$cell_id)] <= 50,
                    "within 50 km", "50–100 km")

source("../R/functions.R", local = TRUE)
annex1 <- annex1_habitat_types("../data/eurlex/habitats_directive.html")
feat$name[feat$kind == "habitat"] <-
  annex1$name[match(feat$code[feat$kind == "habitat"], annex1$code)]

n_cells <- table(ifelse(cells$dist_km <= 50, "within 50 km", "50–100 km"))
by_band <- function(b) {
  f <- feat[feat$band == b | (b == "within 100 km"), ]
  c(cells     = length(unique(f$cell_id)),
    countries = paste(sort(unique(f$country)), collapse = ", "),
    species   = length(unique(f$code[f$kind == "species"])),
    habitats  = length(unique(f$code[f$kind == "habitat"])))
}
summary_tab <- rbind(`Within 50 km` = by_band("within 50 km"),
                     `Within 100 km` = by_band("within 100 km"))
knitr::kable(data.frame(Zone = rownames(summary_tab), summary_tab,
                        check.names = FALSE, row.names = NULL),
             col.names = c("Zone", "10 km cells with a record", "Countries",
                           "Species", "Annex I habitat types"),
             align = "lrlrr")
```

```{r}
#| label: check-text
#| eval: !expr have_cache
#| output: asis
near <- feat[feat$band == "within 50 km", ]
cat(sprintf(paste(
  "Within 50 km of Kosovo, %d cells carry at least one reported feature, all",
  "of them in %s. Nothing lies closer than %s km: the nearest reported cell",
  "is that far from the border, because Kosovo is ringed by four countries",
  "that are not in the EU. Features flagged sensitive in the extract: %d of %d."),
  length(unique(near$cell_id)), paste(sort(unique(near$country)), collapse = " and "),
  format(min(cells$dist_km)), sum(feat$sensitive %in% "sensitive"), nrow(feat)))
```

```{r}
#| label: map
#| eval: !expr have_cache
ko <- kosovo_outline()
cells$features <- as.integer(table(feat$cell_id)[cells$cell_id])
rings <- st_transform(st_sf(ring = c("50 km", "100 km"),
                            geometry = st_buffer(st_transform(st_union(ko), 3035),
                                                 c(50000, 100000))), 4326)
pal <- colorBin(map_palette$density, cells$features, bins = 5)
leaflet(st_transform(cells, 4326)) |>
  add_basemaps() |>
  addPolygons(fillColor = ~pal(features), fillOpacity = 0.75, weight = 0.5,
              color = "#666666",
              label = ~sprintf("%s: %d features, %s km from Kosovo",
                               cell_id, features, dist_km)) |>
  addPolylines(data = rings, color = "#175CA1", weight = 1.5, dashArray = "4 4",
               label = ~ring) |>
  add_boundary_outline(ko) |>
  addLegend(pal = pal, values = ~features,
            title = "Reported features<br>per 10 km cell") |>
  addLayersControl(baseGroups = basemap_groups,
                   options = layersControlOptions(collapsed = TRUE))
```

::: {.callout-note collapse="true"}
## Every habitat and species reported within 100 km

```{r}
#| label: lists
#| eval: !expr have_cache
lst <- aggregate(cell_id ~ kind + code + name + band, data = feat,
                 FUN = function(x) length(unique(x)))
lst <- lst[order(lst$band != "within 50 km", lst$kind, -lst$cell_id), ]
knitr::kable(
  data.frame(Zone = lst$band, Kind = lst$kind, Code = lst$code,
             Name = ifelse(lst$kind == "species", paste0("*", lst$name, "*"),
                           lst$name),
             Cells = lst$cell_id),
  align = "lllr", row.names = FALSE)
```

:::

The list is a cross-border reference, not a Kosovo checklist. A species
reported in a Bulgarian cell 60 km away may or may not cross the
border, and the absence of a feature says only that those countries did not
report it there.

## What it adds for Natura 2000 in Kosovo

It answers a question none of the report's current sources can: which Annex I
habitats and Annex II species the EU's own reporting places closest to Kosovo.
The assessment tables in the same file add each feature's status in the
reporting country and region, joined on `CO_MS_RE`. That is a starting list for the
[sufficiency ledger](../natura2000.qmd#q4) and a plausibility check for the
directive-listed species the GBIF extract finds. At 10 km it serves the
[national screening tier](../natura2000.qmd#q1) only.

## Caveats and sensitivity

- The countries that surround Kosovo (Serbia, Montenegro, Albania, North
  Macedonia) are outside the EU and absent, so the nearest data lie about
  50 km away, and most lie beyond it.
- The public version leaves out sensitive species altogether: the EEA's
  README describes it as "the public dataset, without sensitive species". A
  persecuted raptor or a collected orchid can therefore be missing from these
  lists even where a neighbour reported it. The extract keeps the `sensitive`
  column the file carries.
- The data are ten years old by the end of the next round; see the open check
  below on 2019–2024.

## Integration into the pipeline

Low effort. The remote read above is the same pattern as
`kosovo_protected_areas()`. A proposed `art17_border_features(boundary,
radius_km = c(50, 100))` in `R/functions.R`, called from a new section after
7b of `pipeline.R`, would write the long table to `data/` for the report to
read. This page does not change `pipeline.R`.

## To be completed

```{r}
#| label: todo
#| output: asis
todo_list(row)
```

## Citation

Neither the catalogue record nor the README gives a citation string. The
CC-BY 4.0 licence requires attribution to the copyright holders it names, the
Commission's DG Environment and the EEA, so the dataset is cited here by its
own title: *Conservation status of
habitat types and species: datasets from Article 17, Habitats Directive
92/43/EEC reporting (2013–2018), public version, August 2020.* European
Environment Agency and European Commission DG Environment, CC-BY 4.0.
Catalogue record 9f71b3e3-f8ec-442b-a2d5-c3c190605ac4.
