---
title: "Hydropower projects on Balkan rivers"
subtitle: "RiverWatch and EuroNatur's inventory of existing, planned and under-construction hydropower plants, updated for 2024."
actions:
  - kind: request
    what: "Point data of hydropower plants in Kosovo (status, capacity, river), 2024 update, under stated terms of use"
    contact: "RiverWatch: ulrich.eichelmann@riverwatch.eu or cornelia.wieser@riverwatch.eu (as given on the 2024 news page)"
  - kind: check
    what: "Kosovo's figures in the 2024 update (the 2022 update gave 89 planned and 10 under construction), and the country code the dataset uses for Kosovo; the desk review reported 'KV' in older versions"
    contact: "Balkan HPP Update 2024 report, linked from the RiverWatch news page"
todo:
  doi: "No DOI; a report and online map"
  licence: "Terms of use of the point data; ask RiverWatch with the request"
  access_method: "How the point data would be delivered; ask RiverWatch"
  spatial_resolution: "Point locations of plants; precision to be stated with the data"
  temporal_coverage: "Inventory years 2012–2024 across the updates"
  kosovo_codes_seen: "Country code for Kosovo in the dataset ('KV' in older versions, per the desk review)"
  effort: "Low once points arrive: an overlay with NatDA and the river network"
---

```{r}
#| label: setup
source("../R/sources_helpers.R")
row <- source_row("balkan-hydropower")
```

::: {.source-status}
`r status_line(row)`
:::

## At a glance

```{r}
#| label: glance
metadata_table(row)
```

## What it is

Since 2012 RiverWatch and EuroNatur have mapped hydropower plants on Balkan
rivers, with data compiled by Fluvius in Vienna. The 2024 update, announced
on 14 March 2025, counts 3,188 projects planned, 94 under construction and
1,836 in operation across the region, with 110 new installations since 2022.
The 2022 update, prepared by Ulrich Schwarz, lists Kosovo separately: 89
plants planned and 10 under construction at the time, and it names Kosovo
among the countries where most new plants came into operation since 2020. It
overlays the plants with protected areas, including Emerald and proposed
Natura 2000 areas, and was delivered in GIS formats and as the online map at
balkanrivers.net.

## Access

```{r}
#| label: access
#| echo: true
#| code-fold: show
access_table(list(
  try_access("https://riverwatch.eu/en/balkanrivers/news/10-years-hydropower-data-2024-update-reveals-ongoing-threats-and-wins-balkan"),
  try_access("https://balkanrivers.net/uploads/files/3/Balkan_HPP_Update_2022.pdf"),
  try_access("https://balkanrivers.net/")
))
```

The reports and the map are public; the point data are not offered for
download. The 2024 news page gives two contacts for questions. No request was
sent.

## Kosovo coverage check

::: {.callout-warning}
## To be completed
The reports cover Kosovo, but the points were not available, so no overlay
could be run. With the data: clip to `kosovo_outline()`, count plants by
status, intersect with the NatDA sites and with the river network (EU-Hydro
Strahler order), and list the rivers with the most planned plants. Record the
country code the dataset gives Kosovo.
:::

## What it adds for Natura 2000 in Kosovo

Small hydropower is one of the pressures most often cited for Balkan rivers.
A site proposal for freshwater habitats and fish (Annex I 3260, Annex II fish)
needs to know which rivers are dammed or about to be. The overlay would also
show which planned plants fall inside existing protected areas, a question an
appropriate assessment would ask. Point data would serve the
[site tier](../natura2000.qmd#q1) for pressures.

## Caveats and sensitivity

- The inventory is compiled by NGOs from several sources; status and
  location should be checked against official permits.
- No species locations; nothing sensitive.

## Integration into the pipeline

Low effort once points arrive: a proposed `read_hydropower(path, boundary)`
would clip them and add a pressure layer to the report's protected-area map.
`pipeline.R` is not changed.

## To be completed

```{r}
#| label: todo
#| output: asis
todo_list(row)
```

## Citation

For the 2022 report: Schwarz, U. (2022). *Hydropower Projects on Balkan
Rivers – 2022 Update.* Fluvius, Vienna, for RiverWatch and EuroNatur; edited
by A. Huber (EuroNatur).
