---
title: "Biodiversiteti i Kosovës"
subtitle: "A 22-page KEPA overview of Kosovo's biodiversity from 2012, with approximate species counts per group."
actions:
  - kind: check
    what: "Whether KEPA's later State of Nature reports (a 2018–2021 report is listed on its publications page) update the species counts"
    contact: "https://www.ammk-rks.net/al/drejtorite/38/raporte-studime-publikime"
todo:
  doi: "No DOI; an agency document"
  licence: "No terms stated; ask KEPA before reproducing tables"
  spatial_resolution: "Not spatial: a narrative overview with a table of counts"
  temporal_coverage: "The counts' reference period is not stated; the PDF dates from July 2012"
  effort: "Low: a table of approximate counts, if wanted at all"
  sensitivity: "None"
---

```{r}
#| label: setup
source("../R/sources_helpers.R")
row <- source_row("biodiversity-of-kosovo-overview")
url <- "https://www.ammk-rks.net/assets/cms/uploads/files/Biodiversiteti%20IK/Biodiversiteti_i_Kosoves.pdf"
```

::: {.source-status}
`r status_line(row)`
:::

## At a glance

```{r}
#| label: glance
metadata_table(row)
```

## What it is

*Biodiversiteti i Kosovës* ("Kosovo's biodiversity") is a 22-page overview
in Albanian, prepared by Qenan Maxhuni, biodiversity officer at KEPA and the
Kosovo Institute for Nature Protection. The PDF dates from July 2012. It
describes the country's flora, vegetation, fauna and main natural areas and
names the Sharr mountains and Bjeshkët e Nemuna as the richest in plants and
animals. Its Table 7 gives approximate species numbers per group, many marked
as likely underestimates: about 1,800 flowering plants, 180 birds, 45
mammals, 14 amphibians, 14 reptiles and 30 freshwater fish. It states plainly
that Kosovo does not yet have a complete inventory of its biodiversity, and,
at the time, that it had no red list.

## Access

The file is on KEPA's server, whose certificate had expired; the second
request below skips verification.

```{r}
#| label: access
#| echo: true
#| code-fold: show
access_table(list(try_access(url), try_access(url, ssl_verify = FALSE)))
```

## Kosovo coverage check

The document is national and narrative; there are no records or cells to
count and no country codes. What can be compared is its counts with the GBIF
extract.

```{r}
#| label: check
meta <- readRDS(file.path(sources_root(), "data", "run_metadata.rds"))
b <- read.csv(file.path(sources_root(), "data_exports", "kosovo_birds.csv"))
o <- read.csv(file.path(sources_root(), "data_exports", "kosovo_overall_biodiversity.csv"))
cnt <- function(d, cls) length(unique(d$species[d$class %in% cls & nzchar(d$species)]))
knitr::kable(data.frame(
  Group = c("Birds", "Mammals", "Amphibians", "Reptiles"),
  `KEPA 2012 (approximate)` = c("180*", "45*", "14", "14"),
  `Species in the GBIF extract` = c(cnt(o, "Aves"), cnt(o, "Mammalia"),
                                    cnt(o, "Amphibia"),
                                    cnt(o, c("Squamata", "Testudines", "Reptilia"))),
  check.names = FALSE), align = "lrr", row.names = FALSE,
  caption = sprintf("\\* likely more, according to KEPA. GBIF download %s.", meta$download$doi))
```

The comparison is rough: the 2012 counts are approximate, and GBIF species
counts include vagrants and misidentifications. Where GBIF already exceeds the
2012 figure, the old figure was an underestimate, as its authors expected.

## What it adds for Natura 2000 in Kosovo

Little new data, but a useful statement of the baseline problem in the
agency's own words: no complete inventory, no red list in 2012, and the two
mountain massifs singled out as the richest areas. It frames the data gaps
listed on the [hub page](index.qmd#data-gaps) and serves the
[national tier](../natura2000.qmd#q1) only.

## Caveats and sensitivity

- Approximate counts with no stated reference period or method.
- Superseded in part by the two red books (2013, 2018).
- No sensitive content.

## Integration into the pipeline

None needed. The counts could appear in the report as context, with their
date and source. `pipeline.R` is not changed.

## To be completed

```{r}
#| label: todo
#| output: asis
todo_list(row)
```

## Citation

::: {.callout-warning}
## To be completed
The document gives no citation. Cite as Maxhuni, Q. (2012), *Biodiversiteti i
Kosovës*, KEPA/IKMN, and confirm the preferred form with KEPA.
:::
