---
title: "CHELSA climatologies"
subtitle: "High-resolution global climatologies, read here as cloud-hosted GeoTIFFs for 1981–2010 and cut to Kosovo."
actions:
  - kind: check
    what: "Read CHELSA's licence and requested citation on its own site, and record the custodian (the institution behind envicloud)"
    contact: "https://chelsa-climate.org/"
  - kind: check
    what: "Compare CHELSA bio1 with WorldClim's mean temperature for Kosovo, pixel by pixel, to see where the two climatologies disagree"
    contact: "The WorldClim page in this section"
todo:
  custodian: "Institution maintaining CHELSA; read on chelsa-climate.org"
  doi: "The CHELSA data paper and dataset DOIs; read on chelsa-climate.org"
  licence: "CHELSA's licence; read on its site"
---

```{r}
#| label: setup
source("../R/sources_helpers.R")
suppressPackageStartupMessages({
  library(sf)
  library(terra)
  library(leaflet)
})
row <- source_row("chelsa")
dir <- cache_dir("chelsa")
base <- "https://os.zhdk.cloud.switch.ch/chelsav2/GLOBAL/climatologies/1981-2010/bio/"
vars <- c(bio1 = "CHELSA_bio1_1981-2010_V.2.1.tif", bio12 = "CHELSA_bio12_1981-2010_V.2.1.tif")
cache <- file.path(dir, "chelsa_bio1_bio12_1981-2010_kosovo.tif")
```

::: {.source-status}
`r status_line(row)`
:::

## At a glance

```{r}
#| label: glance
metadata_table(row)
```

## What it is

CHELSA ("Climatologies at high resolution for the earth's land surface
areas") provides bioclimatic layers at 30 arc-seconds, about 1 km, for the
whole globe. Version 2.1 is served from a public object store at
`os.zhdk.cloud.switch.ch`; the 1981–2010 climatology is used here. The two
layers used here are `bio1`, mean annual air
temperature, and `bio12`, annual precipitation.

## Access

```{r}
#| label: access
#| echo: true
#| code-fold: show
access_table(lapply(vars, function(f) try_access(paste0(base, f))))
```

The files are global GeoTIFFs; GDAL reads only the part over Kosovo.

```{r}
#| label: build
#| echo: true
#| code-fold: true
#| code-summary: "Show the remote crop"
build_error <- NULL
if (!file.exists(cache)) {
  build_error <- tryCatch({
    Sys.setenv(GDAL_DISABLE_READDIR_ON_OPEN = "EMPTY_DIR")
    ko <- vect(kosovo_outline())
    k <- rast(lapply(vars, function(f)
      mask(crop(rast(paste0("/vsicurl/", base, f)), ko, snap = "out"), ko)))
    names(k) <- names(vars)
    writeRaster(k, cache, overwrite = TRUE, gdal = "COMPRESS=DEFLATE")
    NULL
  }, error = function(e) conditionMessage(e))
}
have_cache <- file.exists(cache)
```

```{r}
#| label: build-failed
#| eval: !expr "!have_cache"
#| output: asis
cat("::: {.callout-warning}\n## To be completed\nThe remote crop failed at render time (",
    build_error %||% "no message", "). Re-render when the CHELSA store answers.\n:::\n", sep = "")
```

## Kosovo coverage check

```{r}
#| label: check
#| eval: !expr have_cache
k <- rast(cache)
inside <- !is.na(values(rasterize(vect(kosovo_outline()), k[[1]]), mat = FALSE))
s <- sapply(names(k), function(n) {
  v <- values(k[[n]], mat = FALSE)[inside]
  c(cells = length(v), na = sum(is.na(v)), min = min(v, na.rm = TRUE),
    mean = mean(v, na.rm = TRUE), max = max(v, na.rm = TRUE))
})
knitr::kable(data.frame(Layer = c("bio1, mean annual temperature (°C)",
                                  "bio12, annual precipitation (mm)"),
                        Cells = fmt_int(s["cells", ]), `No-data` = s["na", ],
                        Min = round(s["min", ], 1), Mean = round(s["mean", ], 1),
                        Max = round(s["max", ], 1), check.names = FALSE),
             align = "lrrrrr", row.names = FALSE)
```

Kosovo is fully covered. The values are those GDAL returns after applying
the scale and offset stored in each file. No country attribute exists, so no
country code applies.

Two things about the read are worth teaching. First, nothing is downloaded
in full: GDAL's `/vsicurl/` prefix turns the file's address into something
`terra` can open, and the crop asks the server only for the parts of the
global grid that cover Kosovo. Second, the check counts no-data cells inside
the outline rather than trusting a map to look complete. A European or
global product can leave a country blank for reasons of its own, and a
blank that falls on the country being studied is easy to miss on a map at
continental scale. Here there is none.

```{r}
#| label: map
#| eval: !expr have_cache
source("../R/functions.R", local = TRUE)
pal <- colorNumeric(grDevices::hcl.colors(9, "RdYlBu", rev = TRUE),
                    values(k[["bio1"]]), na.color = NA)
leaflet() |>
  add_basemaps() |>
  addRasterImage(k[["bio1"]], colors = pal, opacity = 0.75) |>
  add_boundary_outline(kosovo_outline()) |>
  addLegend(pal = pal, values = values(k[["bio1"]]), title = "Mean annual<br>temperature (°C)") |>
  addLayersControl(baseGroups = basemap_groups,
                   options = layersControlOptions(collapsed = TRUE))
```

## What it adds for Natura 2000 in Kosovo

Climate covariates for species distribution models, and for explaining the
split between Kosovo's Alpine and Continental regions. At about 1 km they suit
national screening ([tier](../natura2000.qmd#q1)). If CHELSA also offers
future scenarios at the same resolution, which this check did not confirm,
they would bear on the "future prospects" parameter of habitat assessments.

## Caveats and sensitivity

- Interpolated climate: in mountains the 1 km cells average steep gradients.
- Licence and citation not yet read on CHELSA's site.
- No sensitive content.

## Integration into the pipeline

Low effort: a proposed `kosovo_climate(boundary, source = "chelsa")` would
cache the layers above for the SDM work. `pipeline.R` is not changed.

## To be completed

```{r}
#| label: todo
#| output: asis
todo_list(row)
```

## Citation

::: {.callout-warning}
## To be completed
Use the citation CHELSA requests on its site; it was not read in this check.
:::
