---
title: "European Breeding Bird Atlas 2"
subtitle: "The European Bird Census Council's atlas of breeding birds from fieldwork mostly in 2013–2017, at 50 km observed and 10 km modelled."
actions:
  - kind: request
    what: "EBBA2 data for the 50 km squares covering Kosovo (breeding evidence and abundance class per species) and the 10 km modelled probability of occurrence for the same area"
    contact: "EBBA2 data request form, https://www.ebba2.info/data-req-form/"
  - kind: check
    what: "Read the 'EBBA2 Data Access policy' document linked from the data-request page, and record which data are open (and under which licence) and which need EBCC approval"
    contact: "https://ebba2.info/data-request/"
  - kind: check
    what: "Whether and by whom Kosovo was covered in EBBA2 fieldwork (national coordinator)"
    contact: "EBBA2 map viewer https://mapviewer.ebba2.info/ or EBCC"
todo:
  doi: "No DOI found for the EBBA2 dataset; ask EBCC"
  licence: "The desk review's claim that 50 km data are free under a Creative Commons licence is unconfirmed; the terms are in the 'EBBA2 Data Access policy' document"
  format: "Delivery format of requested data; ask EBCC with the request"
  access_method: "Request form; confirm whether any data are downloadable without approval"
  kosovo_coverage: "Whether Kosovo's squares were surveyed in EBBA2; not stated on the pages opened"
  kosovo_codes_seen: "Which country code or name EBBA2 uses for Kosovo's squares; check in the map viewer"
  effort: "Low once a table of squares × species arrives; estimate after the request"
  sensitivity: "EBBA2's own rules for sensitive species at 10 km; read in the data access policy"
---

```{r}
#| label: setup
source("../R/sources_helpers.R")
row <- source_row("ebba2")
cache <- file.path(cache_dir("ebba2"), "gbif_ebcc_atlas_check.csv")
```

::: {.source-status}
`r status_line(row)`
:::

## At a glance

```{r}
#| label: glance
metadata_table(row)
```

## What it is

The second European Breeding Bird Atlas, published in 2020 by the European
Bird Census Council (EBCC) with Lynx Edicions, maps the distribution,
abundance and change of Europe's breeding birds from fieldwork carried out
mostly between 2013 and 2017 by some 120,000 fieldworkers in 48 countries.
It offers two kinds of map: observed abundance and breeding likelihood at 50
× 50 km, and modelled probability of occurrence at 10 × 10 km. An online
version followed in 2021.

## Access

```{r}
#| label: access
#| echo: true
#| code-fold: show
access_table(list(
  try_access("https://www.ebcc.info/what-we-do/ebba2/"),
  try_access("https://ebba2.info/data-request/"),
  try_access("https://www.ebba2.info/data-req-form/"),
  try_access("https://mapviewer.ebba2.info/")
))
```

The data-request page says that "some data are open access while some
requests require the acceptance of the EBCC", and routes every request
through one form. The detailed terms are in a Word document linked from that
page, which was not opened. No request was submitted; it is listed as an open
action below.

## Kosovo coverage check

The EBBA2 data could not be checked without a request. What can be checked is
whether its predecessor is already in this project's data. The first EBCC
atlas (1997, fieldwork 1972–1995) is published on GBIF.

```{r}
#| label: build
#| echo: true
#| code-fold: true
#| code-summary: "Show the GBIF check"
if (!file.exists(cache)) {
  key <- "c779b049-28f3-4daf-bbf4-0a40830819b6"   # EBCC Atlas of European Breeding Birds
  count <- function(q) tryCatch(
    jsonlite::fromJSON(sprintf("https://api.gbif.org/v1/occurrence/search?datasetKey=%s&limit=0&%s",
                               key, q))$count, error = function(e) NA)
  reg <- read.csv(file.path(sources_root(), "data", "dataset_registry.csv"))
  write.csv(data.frame(
    check = c("Records coded XK", "Records inside GADM XKO", "In this project's GBIF download"),
    value = c(fmt_int(count("country=XK")), fmt_int(count("gadmGid=XKO")),
              ifelse(key %in% reg$datasetKey, "yes", "no"))),
    cache, row.names = FALSE)
}
knitr::kable(read.csv(cache), col.names = c("Check (EBCC atlas 1997 on GBIF)", "Result"))
```

The 1997 atlas is already in the download behind the report, and GBIF codes
its Kosovo squares XK. EBBA2 is not on GBIF, so the newer atlas adds records
the report does not yet hold.

::: {.callout-warning}
## To be completed
Once EBBA2 data arrive: list the 50 km squares that intersect
`kosovo_outline()`, the species recorded in each with breeding evidence, and
compare with the GBIF bird records. Show atlas data only at its native 50 km
or 10 km grid.
:::

## What it adds for Natura 2000 in Kosovo

Nine in ten of the 28,785 bird records in the report's GBIF extract come from
eBird checklists, and they cluster where observers go. An atlas is a systematic survey with breeding evidence in
every square, which is what an SPA case needs to separate breeding from
passage birds (see [question 6](../natura2000.qmd#q6)). The 10 km modelled
maps would also give a first, consistent distribution for Annex I species. At
50 km and 10 km the data serve the [national screening
tier](../natura2000.qmd#q1).

## Caveats and sensitivity

- Access terms are unconfirmed; the claim of freely licensed 50 km data from
  the desk review could not be verified.
- Modelled 10 km maps are predictions, not observations.
- Atlas data must be shown at their native grid; finer locations for
  sensitive species must not be derived from them.

## Integration into the pipeline

Low effort once a squares-by-species table arrives: a proposed
`read_atlas_squares(path, grid_km = 50)` would join it to a grid clipped to
Kosovo and add an atlas layer to the report's bird section. `pipeline.R` is
not changed.

## To be completed

```{r}
#| label: todo
#| output: asis
todo_list(row)
```

## Citation

As EBCC requests it: Keller, V., Herrando, S., Voříšek, P., et al. (2020).
*European Breeding Bird Atlas 2: Distribution, Abundance and Change.*
European Bird Census Council & Lynx Edicions, Barcelona.
