---
title: "eBird and Observation.org records outside GBIF"
subtitle: "Both platforms already publish to GBIF and are in the report's download; the open question is what their own databases hold beyond those copies."
actions:
  - kind: request
    what: "The eBird Basic Dataset (EBD) and its Sampling Event Data for Kosovo, which carry protocol, duration, distance and whether each checklist is complete"
    contact: "Cornell Lab of Ornithology, eBird data request at https://science.ebird.org (the site's bot protection blocked automated reading)"
  - kind: check
    what: "Read the EBD terms of use in a browser and record the licence and whether non-commercial use by a public body is covered"
    contact: "https://science.ebird.org/en/use-ebird-data/download-ebird-data-products"
  - kind: check
    what: "Ask Observation.org whether records hidden from GBIF (sensitive species, unvalidated records) exist for Kosovo, and on what terms they could be shared"
    contact: "Observation International, through https://observation.org"
todo:
  doi: "The GBIF copies have DOIs (eBird 10.15468/aomfnb; Observation.org 10.15468/5nilie); the full eBird database has none"
  format: "File format of the EBD; read on the eBird site, which blocked automated access"
  spatial_resolution: "Checklist location; precision depends on the protocol, and eBird states no coordinate uncertainty"
  temporal_coverage: "Years of Kosovo records in each platform; read from the check below and from the EBD once obtained"
  latest_version: "Release cycle and latest release of the EBD; read on the eBird site"
  kosovo_codes_seen: "Read from the check below"
  tier: "National for travelling checklists; delineation-grade only for stationary counts with precise locations"
  effort: "Medium: the EBD is large and needs the auk R package or equivalent"
  sensitivity: "Both platforms hide or generalise sensitive species in public views; ask what applies to Kosovo"
---

```{r}
#| label: setup
source("../R/sources_helpers.R")
row <- source_row("ebird-observation-org")
cache <- file.path(cache_dir("ebird-observation-org"), "gbif_platform_check.csv")
keys <- c(eBird = "4fa7b334-ce0d-4e88-aaae-2e0c138d049e",
          `Observation.org` = "8a863029-f435-446a-821e-275f4f641165")
```

::: {.source-status}
`r status_line(row)`
:::

## At a glance

```{r}
#| label: glance
metadata_table(row)
```

## What it is

eBird, run by the Cornell Lab of Ornithology, and Observation.org are the two
citizen-science platforms that contribute most to Kosovo's GBIF record. Both
publish to GBIF: eBird as the "EOD – eBird Observation Dataset" under
CC BY 4.0, Observation.org as "Observation.org, Nature data from around the
World" under CC BY-NC 4.0. Both are already among the datasets in the report's
download, and eBird alone supplies nine in ten of its bird records.

The desk review asked about records *not* in GBIF. For eBird that means the
eBird Basic Dataset (EBD). The site's [Natura 2000 page](../natura2000.qmd#q6)
describes it as carrying what the GBIF copy leaves out: the effort fields and
the flag that says a checklist reports every species detected, which turn a
list of sightings into a survey with non-detections. eBird's own description
could not be read to confirm this (see Access). For Observation.org it means
records the platform does not pass to GBIF.

## Access

```{r}
#| label: access
#| echo: true
#| code-fold: show
access_table(list(
  try_access(paste0("https://api.gbif.org/v1/dataset/", keys[["eBird"]])),
  try_access(paste0("https://api.gbif.org/v1/dataset/", keys[["Observation.org"]])),
  try_access("https://science.ebird.org/en/use-ebird-data/download-ebird-data-products"),
  try_access("https://observation.org/")
))
```

The GBIF copies are open. eBird's data-products page returned an "access
denied" page from bot-protection software to automated reads, so its terms
could not be verified; the request for the EBD is listed as an open action.

## Kosovo coverage check

```{r}
#| label: build
#| echo: true
#| code-fold: true
#| code-summary: "Show the GBIF check"
if (!file.exists(cache)) {
  res <- tryCatch({
    reg <- read.csv(file.path(sources_root(), "data", "dataset_registry.csv"))
    do.call(rbind, lapply(names(keys), function(n) {
      q <- function(extra) jsonlite::fromJSON(sprintf(
        "https://api.gbif.org/v1/occurrence/search?datasetKey=%s&limit=0%s", keys[[n]], extra))
      g <- q("&gadmGid=XKO&facet=country&facet=year&facetLimit=100")
      cc <- g$facets$counts[[which(g$facets$field == "COUNTRY")]]
      yr <- as.integer(g$facets$counts[[which(g$facets$field == "YEAR")]]$name)
      data.frame(platform = n, records_in_gadm_xko = g$count,
                 records_coded_xk = q("&country=XK")$count,
                 codes_inside_gadm = paste(sprintf("%s %s", cc$name, cc$count), collapse = "; "),
                 years = sprintf("%d–%d", min(yr), max(yr)),
                 in_report_download = keys[[n]] %in% reg$datasetKey)
    }))
  }, error = function(e) NULL)
  if (!is.null(res)) write.csv(res, cache, row.names = FALSE)
}
have_cache <- file.exists(cache)
```

```{r}
#| label: build-failed
#| eval: !expr "!have_cache"
#| output: asis
cat("::: {.callout-warning}\n## To be completed\nThe GBIF API did not answer at render time; re-render to run the check.\n:::\n")
```

```{r}
#| label: check
#| eval: !expr have_cache
d <- read.csv(cache)
knitr::kable(data.frame(Platform = d$platform,
                        `Records inside GADM XKO` = fmt_int(d$records_in_gadm_xko),
                        `Records coded XK` = fmt_int(d$records_coded_xk),
                        `Country codes inside GADM XKO` = d$codes_inside_gadm,
                        Years = d$years,
                        `In the report's download` = ifelse(d$in_report_download, "yes", "no"),
                        check.names = FALSE), align = "lrrlll", row.names = FALSE)
```

The live GBIF counts differ from the report's figures because GBIF keeps
indexing new records after the report's dated download. Records inside the
GADM outline that carry a neighbour's code are the border records the
pipeline's country test removes.

::: {.callout-warning}
## To be completed
What the platforms hold beyond GBIF could not be measured. For eBird, request
the EBD and Sampling Event Data for Kosovo and compare checklist and species
counts with the GBIF copy. For Observation.org, ask what is withheld from
GBIF for Kosovo and why.
:::

## What it adds for Natura 2000 in Kosovo

An SPA case needs counts with effort, not scattered presences ([question
6](../natura2000.qmd#q6)). The EBD's effort fields and complete-checklist
flag are what make eBird usable for that: they allow detection rates,
seasonal occupancy and absences to be computed for Ligatina e Hencit/Radevës
and the other bird sites. Stationary counts with precise locations can reach
the [delineation tier](../natura2000.qmd#q1); travelling checklists stay at
screening.

## Caveats and sensitivity

- The EBD terms could not be read; the licence of the full eBird database is
  therefore not recorded.
- Both platforms hide or generalise sensitive species in their public data;
  what applies to Kosovo is unknown.
- Observation.org's GBIF copy is CC BY-NC, which limits commercial reuse of
  anything derived from it.

## Integration into the pipeline

Medium effort for the EBD: a proposed `read_ebd(path, boundary)` using the
`auk` package would filter the file to Kosovo and keep checklist-level effort
beside the GBIF records, linked by checklist identifier rather than merged.
`pipeline.R` is not changed.

## To be completed

```{r}
#| label: todo
#| output: asis
todo_list(row)
```

## Citation

Cite the GBIF copies as GBIF shows them (eBird:
[doi:10.15468/aomfnb](https://doi.org/10.15468/aomfnb); Observation.org:
[doi:10.15468/5nilie](https://doi.org/10.15468/5nilie)).

::: {.callout-warning}
## To be completed
The EBD citation form is on eBird's site, which could not be read.
:::
