---
title: "European Commission Kosovo Report 2025"
subtitle: "The Commission's annual assessment of Kosovo's progress towards the EU, whose Chapter 27 judges nature protection."
actions:
  - kind: check
    what: "Read Chapter 27 in full for the nature-protection recommendations the Commission sets for the coming year, and record them as targets"
    contact: "https://enlargement.ec.europa.eu/kosovo-report-2025_en"
todo:
  doi: "No DOI; cite as a Commission staff working document"
  licence: "Reuse terms of Commission documents; read on the Commission page"
  spatial_resolution: "Not spatial"
  temporal_coverage: "The reporting period of the 2025 report; read in the introduction"
  effort: "Low: a document to cite"
  sensitivity: "None"
---

```{r}
#| label: setup
source("../R/sources_helpers.R")
row <- source_row("ec-kosovo-report-2025")
pdf_url <- "https://enlargement.ec.europa.eu/document/download/127563ea-4c03-44a4-b56c-2d569afd86a5_en?filename=kosovo-report-2025.pdf"
```

::: {.source-status}
`r status_line(row)`
:::

## At a glance

```{r}
#| label: glance
metadata_table(row)
```

## What it is

Each year the European Commission assesses Kosovo's progress towards the EU
across the chapters of the acquis. The 2025 report was published on 4
November 2025. Its Chapter 27, on environment and climate change, finds
Kosovo "at an early stage of preparation" with limited progress.

On nature protection the report is short and specific. Kosovo made limited
progress. The identification of potential Natura 2000 sites "is at a very
early stage". Protected areas lack appropriate management and suffer from
pollution, illegal activities and weak enforcement. There was no progress in
aligning legislation on nature protection. The adoption of a strategy for
environmental protection and sustainable development for 2024–2033 remains
pending. On forests it notes that 40 per cent of public forests are subject
to uncontrolled or illegal harvesting.

## Access

```{r}
#| label: access
#| echo: true
#| code-fold: show
access_table(list(
  try_access("https://enlargement.ec.europa.eu/kosovo-report-2025_en"),
  try_access(pdf_url)
))
```

Open; the report is a 2 MB PDF linked from the Commission's page.

## Kosovo coverage check

The report is about Kosovo alone; there are no data to check. The quotations
above were read from the PDF on 29 September 2026.

What can be checked is consistency with the other pages. The Commission's
finding that identifying potential Natura 2000 sites is "at a very early
stage" matches the absence of any candidate site list in this section (see
[Mustafa 2009](mustafa-2009-potential-n2k.qmd) and the
[data gaps](index.qmd#data-gaps)). Its note that the environment strategy for
2024–2033 is still pending matches the lack of a successor to the
[biodiversity action plan](nbsap-2016-2020.qmd).

## What it adds for Natura 2000 in Kosovo

It is the external benchmark. Whatever this site and the Ministry produce
towards Natura 2000 will be read against the Commission's next assessment,
and the phrases above are the ones it will want to see changed: the
identification of potential sites, protected-area management and legal
alignment. It serves the [national tier](../natura2000.qmd#q1) as context.
The finding on illegal harvesting bears on the [forest-loss](forest-loss-hansen.qmd)
and [NFI](kosovo-nfi-2012.qmd) pages.

## Caveats and sensitivity

- An assessment, not data; its judgements are the Commission's.
- No sensitive content.

## Integration into the pipeline

None; a document to cite. `pipeline.R` is not changed.

## To be completed

```{r}
#| label: todo
#| output: asis
todo_list(row)
```

## Citation

European Commission (2025). *Kosovo 2025 Report.* Brussels, 4 November
2025.

::: {.callout-warning}
## To be completed
The SWD document number was not read; add it from the PDF's cover page.
:::
