---
title: "Ecosystem types of Europe 2012, version 3.1"
subtitle: "The EEA's 100 m map of EUNIS level-2 habitat groups across the EEA39 countries, Kosovo included."
actions:
  - kind: check
    what: "Compare the class areas with the CORINE-based habitat screen in the report, class by class, and note where the two disagree"
    contact: "data/kosovo_clc2018.gpkg and the report's section on potential Natura 2000 habitats"
  - kind: check
    what: "Read the reliability map 2012 v3.1, the secondary product the README names, for Kosovo's pixels"
    contact: "EEA catalogue, search 'Ecosystem types of Europe 2012 reliability'"
todo: {}
---

```{r}
#| label: setup
source("../R/sources_helpers.R")
suppressPackageStartupMessages({
  library(sf)
  library(terra)
  library(leaflet)
})
row <- source_row("eea-ecosystem-types")
dir <- cache_dir("eea-ecosystem-types")
cache <- file.path(dir, "ecosystem_types_2012_v3_1_kosovo.tif")
base <- "https://sdi.eea.europa.eu/datashare/s/fFmHEKnWo4Rt4dP/download?path=%2F&files="
tif <- "eea_r_3035_100_m_etm-full_2012_v3-1_r00.tif"
```

::: {.source-status}
`r status_line(row)`
:::

## At a glance

```{r}
#| label: glance
metadata_table(row)
```

## What it is

The EEA's ecosystem-type map combines land cover with habitat information
from the EUNIS classification to show, for every 100 m cell across the EEA39
countries, the most probable EUNIS habitat at level 2 on land (level 3 at
sea). Version 3.1, published in February 2019, describes 2012 and is licensed
CC-BY 4.0. The EEA's own abstract is candid about its limits: because it is
"partly based on non-spatial explicit mapping", its spatial and thematic
accuracy is not that of a delineated habitat map.

## Access

Open, with no account. The 277 MB GeoTIFF is internally tiled, so GDAL reads
only the tiles over Kosovo; the crop takes seconds and is cached here with
the file's own class table.

```{r}
#| label: access
#| echo: true
#| code-fold: show
access_table(list(
  try_access("https://www.eea.europa.eu/en/datahub/datahubitem-view/573ff9d5-6889-407f-b3fc-cfe3f9e23941"),
  try_access(paste0(base, tif))
))
```

```{r}
#| label: build
#| echo: true
#| code-fold: true
#| code-summary: "Show the remote crop"
build_error <- NULL
if (!file.exists(cache)) {
  build_error <- tryCatch({
    Sys.setenv(GDAL_DISABLE_READDIR_ON_OPEN = "EMPTY_DIR")
    r  <- rast(paste0("/vsicurl/", base, tif))
    ko <- vect(st_transform(kosovo_outline(), 3035))
    k  <- mask(crop(r, ko, snap = "out"), ko)
    writeRaster(k, cache, overwrite = TRUE, gdal = "COMPRESS=DEFLATE")
    write.csv(cats(k)[[1]], file.path(dir, "classes.csv"), row.names = FALSE)
    NULL
  }, error = function(e) conditionMessage(e))
}
have_cache <- file.exists(cache)
```

```{r}
#| label: build-failed
#| eval: !expr "!have_cache"
#| output: asis
cat("::: {.callout-warning}\n## To be completed\nThe remote crop failed at render time (",
    build_error %||% "no message", "). Re-render when the EEA share answers.\n:::\n", sep = "")
```

## Kosovo coverage check

```{r}
#| label: check
#| eval: !expr have_cache
k   <- rast(cache)
cls <- read.csv(file.path(dir, "classes.csv"))
ko  <- vect(st_transform(kosovo_outline(), 3035))
inside <- rasterize(ko, k)
n_in   <- global(inside, "sum", na.rm = TRUE)[[1]]
vals   <- values(k, mat = FALSE)[!is.na(values(inside, mat = FALSE))]
n_na   <- sum(is.na(vals))
f <- as.data.frame(table(value = vals))
f$value <- as.integer(as.character(f$value))
f <- merge(f, cls, by.x = "value", by.y = names(cls)[1])
f$km2 <- f$Freq / 100
f <- f[order(-f$km2), ]
knitr::kable(data.frame(`EUNIS` = f$EUNIS, Class = f$EUNIS_L2, `Level 1` = f$EUNIS_L1,
                        `Area (km²)` = format(round(f$km2, 1), big.mark = ","),
                        `Share` = sprintf("%.1f %%", 100 * f$Freq / sum(f$Freq)),
                        check.names = FALSE),
             align = "lllrr", row.names = FALSE)
```

```{r}
#| label: check-text
#| eval: !expr have_cache
#| output: asis
cat(sprintf(paste(
  "The OpenStreetMap outline covers %s cells of 100 m. %s of them (%.2f per",
  "cent) are no-data. %d EUNIS level-2 classes occur; the largest is %s, at",
  "%.0f per cent of the country. The raster carries no country attribute, so",
  "no country code applies."),
  fmt_int(n_in), fmt_int(n_na), 100 * n_na / n_in, nrow(f),
  f$EUNIS_L2[1], 100 * f$Freq[1] / sum(f$Freq)))
```

```{r}
#| label: map
#| eval: !expr have_cache
source("../R/functions.R", local = TRUE)
k4  <- project(aggregate(k, 3, fun = "modal"), "EPSG:3857", method = "near")
lvl <- sort(unique(values(k4, mat = FALSE)))
lab <- cls$EUNIS[match(lvl, cls[[1]])]
pal <- colorFactor(grDevices::hcl.colors(length(lvl), "Dark 3"), domain = lvl, na.color = NA)
leaflet() |>
  add_basemaps() |>
  addRasterImage(k4, colors = pal, opacity = 0.75, project = FALSE) |>
  add_boundary_outline(kosovo_outline()) |>
  addLegend(colors = pal(lvl), labels = lab, title = "EUNIS level 2", opacity = 0.75) |>
  addLayersControl(baseGroups = basemap_groups,
                   options = layersControlOptions(collapsed = TRUE))
```

The map is aggregated to 300 m by the most frequent class, to keep the page
light; the table is computed on the full 100 m grid.

## What it adds for Natura 2000 in Kosovo

The report's habitat screen reads CORINE Land Cover through a crosswalk to
Annex I. This map is the EEA's own step from land cover towards habitats, at
the same 100 m, done once for all of Europe. It offers an independent second
opinion on where grasslands, heaths, forests and mires lie, and it is
consistent with the maps neighbouring countries would use. At EUNIS level 2
it does not name Annex I types, and it serves the
[national screening tier](../natura2000.qmd#q1) only.

## Caveats and sensitivity

- It describes 2012 and is built on CORINE Land Cover of that period.
- Level-2 classes are broad: "Broadleaved deciduous woodland" (G1) spans
  most of Kosovo's Annex I forest types, and forest that qualifies as none.
- The EEA estimates its reliability in a separate map that this page has not
  yet read.
- No species locations; nothing sensitive.

## Integration into the pipeline

Low effort. A proposed `kosovo_ecosystem_types(boundary)` in `R/functions.R`
would do the crop above and cache `data/kosovo_ecosystem_types_2012.tif`, and
the report could tabulate it beside the CORINE screen. `pipeline.R` is not
changed here.

## To be completed

```{r}
#| label: todo
#| output: asis
todo_list(row)
```

## Citation

The catalogue record gives the identifier
`eea_r_3035_100_m_ecosystem-types-full_p_2012_v03_r01` and the licence
CC-BY 4.0, copyright EEA; it gives no citation string. Cite as: *Ecosystem
types of Europe 2012 – Full map (marine and terrestrial habitats) – version 3
revision 1, February 2019.* European Environment Agency. Catalogue record
faff2281-1fca-4548-89d8-c8ec0c507bc7.
