---
title: "Biogeographical regions of Europe"
subtitle: "The official delineation used by the Habitats Directive and the Emerald Network; it maps Kosovo, and the report already uses it."
actions:
  - kind: check
    what: "Whether the Alpine/Continental split through Kosovo, drawn at 1:1,000,000 to 1:10,000,000, is fit for assigning sites; the NBSAP lists 'definition of clear boundaries between bio-geographic regions' as a preparatory task"
    contact: "Kosovo Institute for Nature Protection; the EEA's biogeographical process team"
todo:
  doi: "No DOI: R/functions.R records that DataCite returns none for the record's identifiers"
  temporal_coverage: "Not applicable: a delineation, edition 2016"
---

```{r}
#| label: setup
source("../R/sources_helpers.R")
suppressPackageStartupMessages({
  library(sf)
  library(leaflet)
})
row <- source_row("eea-biogeographical-regions")
regions_path <- file.path(sources_root(), "data", "kosovo_biogeo_regions.gpkg")
```

::: {.source-status}
`r status_line(row)`
:::

## At a glance

```{r}
#| label: glance
metadata_table(row)
```

## What it is

The biogeographical regions dataset is, in the words of its catalogue record,
"the official delineations used in the Habitats Directive" and for the
Emerald Network under the Bern Convention. The 2016 version (edition 01.00,
26 January 2016) corrected the Pannonian region of Serbia, among other fixes.
It is licensed CC-BY 4.0 by the EEA and drawn at scales between 1:1,000,000
and 1:10,000,000.

The desk review listed it as a lead. It is in fact already in use: the
pipeline reads it from the EEA's map service, clips it to Kosovo in
`kosovo_biogeo_regions()`, caches it as `data/kosovo_biogeo_regions.gpkg`,
and uses it in the habitat screen, so that the crosswalk can differ between
regions.

## Access

```{r}
#| label: access
#| echo: true
#| code-fold: show
access_table(list(
  try_access("https://sdi.eea.europa.eu/catalogue/srv/api/records/c6d27566-e699-4d58-a132-bbe3fe01491b"),
  try_access("https://bio.discomap.eea.europa.eu/arcgis/rest/services/BioRegions/BiogeographicalRegions_LAEA/MapServer/0?f=json")
))
```

## Kosovo coverage check

The pipeline's cached extract answers the question directly.

```{r}
#| label: check
#| eval: !expr file.exists(regions_path)
reg <- st_read(regions_path, quiet = TRUE)
knitr::kable(data.frame(Region = reg$name, Code = reg$region,
                        `Area in Kosovo (km²)` = fmt_int(reg$area_km2),
                        Share = sprintf("%.1f %%", 100 * reg$area_km2 / sum(reg$area_km2)),
                        check.names = FALSE),
             align = "llrr", row.names = FALSE,
             caption = sprintf("Read from the EEA map service on %s", reg$accessed[1]))
```

```{r}
#| label: map
#| eval: !expr file.exists(regions_path)
source("../R/functions.R", local = TRUE)
pal <- colorFactor(c("#009E73", "#E69F00", "#56B4E9"), domain = reg$region)
leaflet(st_transform(reg, 4326)) |>
  add_basemaps() |>
  addPolygons(fillColor = ~pal(region), fillOpacity = 0.4, weight = 1,
              color = ~pal(region), label = ~name) |>
  add_boundary_outline(kosovo_outline()) |>
  addLegend(pal = pal, values = ~region, title = "Biogeographical region") |>
  addLayersControl(baseGroups = basemap_groups,
                   options = layersControlOptions(collapsed = TRUE))
```

Kosovo is mapped; the regions carry no country attribute, so no country code
applies. The line between the regions follows the dataset's coarse scale and
should not be read to the kilometre.

## What it adds for Natura 2000 in Kosovo

Every Natura 2000 site is assessed in its biogeographical region, and
sufficiency is judged region by region. Which region each part of Kosovo lies
in therefore decides which biogeographical seminar would review its sites and
which reference lists apply. The NBSAP names a clear delineation of these
boundaries within Kosovo as a preparatory task; this dataset is the European
starting point for it. It serves the [national tier](../natura2000.qmd#q1).

## Caveats and sensitivity

- Drawn at 1:1,000,000 or coarser: the boundary is good to about a kilometre
  at best, as `R/functions.R` notes.
- No sensitive content.

## Integration into the pipeline

Already integrated (`kosovo_biogeo_regions()` in `R/functions.R`). Nothing
further is needed unless Kosovo defines its own finer delineation.

## To be completed

```{r}
#| label: todo
#| output: asis
todo_list(row)
```

## Citation

As recorded in `biogeo_source` in `R/functions.R`, from the catalogue record:
*Biogeographical regions, Europe 2016, ver. 1.* European Environment Agency,
CC-BY 4.0; administrative boundaries © EuroGeographics, © FAO (UN),
© TurkStat, source: European Commission – Eurostat/GISCO. Catalogue record
c6d27566-e699-4d58-a132-bbe3fe01491b.
