---
title: "EFFIS burnt areas"
subtitle: "The European Forest Fire Information System's mapped burnt areas, served openly as a web feature service that codes Kosovo 'KS'."
actions:
  - kind: check
    what: "Read the EFFIS terms of use shown at download and record the licence"
    contact: "https://forest-fire.emergency.copernicus.eu/applications/data-and-services"
  - kind: request
    what: "Historical burnt-area perimeters before the years the web service exposes, if needed for a longer fire history"
    contact: "EFFIS data request form, https://effis.jrc.ec.europa.eu/apps/data.request.form"
todo:
  doi: "No DOI seen; check the EFFIS documentation"
  licence: "Terms shown at download on the EFFIS data page; not recorded in this check"
  spatial_resolution: "MODIS-derived perimeters; EFFIS states a minimum mapped fire size in its documentation, to be read"
  temporal_coverage: "Years present in the check below"
  latest_version: "The service is updated continuously; record the LASTUPDATE range from the check"
---

```{r}
#| label: setup
source("../R/sources_helpers.R")
suppressPackageStartupMessages({
  library(sf)
  library(leaflet)
})
row <- source_row("effis-burnt-areas")
cache <- file.path(cache_dir("effis-burnt-areas"), "effis_burnt_areas_kosovo.gpkg")
wfs <- "https://maps.effis.emergency.copernicus.eu/effis"
```

::: {.source-status}
`r status_line(row)`
:::

## At a glance

```{r}
#| label: glance
metadata_table(row)
```

## What it is

The European Forest Fire Information System (EFFIS), part of the Copernicus
emergency services, maps burnt areas across Europe, the Middle East and North
Africa from MODIS and VIIRS satellite imagery. Each mapped fire comes with
its dates, its area, and the share of it that was broadleaved, coniferous or
mixed forest, scrub, agricultural or other land, and the share inside
Natura 2000. EFFIS distributes the data through web map and web feature
services, as downloads, and through a request form for historical extracts.

## Access

```{r}
#| label: access
#| echo: true
#| code-fold: show
access_table(list(
  try_access("https://forest-fire.emergency.copernicus.eu/applications/data-and-services"),
  try_access(paste0(wfs, "?service=WMS&request=GetCapabilities"))
))
```

The web map service lists yearly layers of burnt areas (`modis.ba.poly.2016`
to `modis.ba.poly.2025` among them). The web feature service's capabilities
request timed out, but a direct GetFeature request for a box around Kosovo
returned every burnt area in it as GML, which the build step below uses.

```{r}
#| label: build
#| echo: true
#| code-fold: true
#| code-summary: "Show the WFS request"
build_error <- NULL
if (!file.exists(cache)) {
  build_error <- tryCatch({
    ko <- st_union(kosovo_outline())
    bb <- st_bbox(st_transform(ko, 3857))
    # EPSG:3857 keeps x before y; with EPSG:4326, WFS 1.1 swaps the axes.
    url <- sprintf(paste0(wfs, "?service=WFS&version=1.1.0&request=GetFeature",
                          "&typename=ms:modis.ba.poly&srsName=EPSG:3857",
                          "&bbox=%f,%f,%f,%f,EPSG:3857&maxFeatures=50000"),
                   bb[1], bb[2], bb[3], bb[4])
    gml <- tempfile(fileext = ".gml")
    got <- try_access(url, gml, timeout = 300)
    if (!got$success) stop(got$error %||% paste("HTTP", got$http_code))
    x <- st_read(gml, quiet = TRUE) |> st_transform(3035) |> st_make_valid()
    x <- x[, setdiff(names(x), "msGeometry")]
    # Keep only the part of each fire inside Kosovo, as polygons: a fire that
    # merely touches the border leaves lines or points, which carry no area.
    x <- suppressWarnings(st_intersection(x, st_transform(ko, 3035)))
    x <- suppressWarnings(st_collection_extract(x, "POLYGON"))
    # Extraction can split one fire into several rows; put each back together.
    parts <- split(seq_len(nrow(x)), x$id)
    geom  <- do.call(c, lapply(parts, function(i) st_union(st_geometry(x)[i])))
    attrs <- st_drop_geometry(x)[!duplicated(x$id), ]
    attrs <- attrs[match(names(parts), as.character(attrs$id)), ]
    x <- st_sf(attrs, geometry = st_cast(geom, "MULTIPOLYGON"))
    x$in_kosovo_ha <- as.numeric(st_area(x)) / 1e4
    x <- x[x$in_kosovo_ha > 0, ]
    x$year <- as.integer(substr(x$FIREDATE, 1, 4))
    # Areas are measured above, on the full geometry; the stored outlines are
    # generalised to 25 m, well below MODIS's own resolution, to keep the
    # cache small.
    x <- st_simplify(x, dTolerance = 25, preserveTopology = TRUE)
    st_write(x[, c("id", "FIREDATE", "FINALDATE", "LASTUPDATE", "COUNTRY", "AREA_HA",
                   "BROADLEA", "CONIFER", "MIXED", "SCLEROPH", "TRANSIT", "OTHERNATLC",
                   "AGRIAREAS", "ARTIFSURF", "OTHERLC", "year", "in_kosovo_ha")],
             cache, delete_dsn = TRUE, quiet = TRUE)
    NULL
  }, error = function(e) conditionMessage(e))
}
have_cache <- file.exists(cache)
```

## Kosovo coverage check

```{r}
#| label: check
#| eval: !expr have_cache
b <- st_read(cache, quiet = TRUE)
num <- function(v) suppressWarnings(as.numeric(v))
forest_share <- (num(b$BROADLEA) + num(b$CONIFER) + num(b$MIXED)) / 100
by_year <- aggregate(cbind(fires = 1, ha = b$in_kosovo_ha,
                           forest_ha = b$in_kosovo_ha * forest_share) ~ year,
                     data = transform(b, forest_share = forest_share), FUN = sum)
knitr::kable(data.frame(Year = by_year$year, Fires = by_year$fires,
                        `Burnt inside Kosovo (ha)` = fmt_int(by_year$ha),
                        `Of which forest (ha)` = fmt_int(by_year$forest_ha),
                        check.names = FALSE), align = "rrrr", row.names = FALSE)
```

```{r}
#| label: check-text
#| eval: !expr have_cache
#| output: asis
cc <- table(b$COUNTRY)
share_in <- num(b$in_kosovo_ha) / num(b$AREA_HA)
rs <- b$COUNTRY == "RS"
cat(sprintf(paste(
  "%d mapped fires between %d and %d (the last year to the date of the",
  "check) burnt %s ha inside Kosovo. The country codes on those fires are:",
  "%s. EFFIS codes Kosovo **KS**, a code used by none of the other sources",
  "in this section."),
  nrow(b), min(b$year), max(b$year), fmt_int(sum(b$in_kosovo_ha)),
  paste(sprintf("%s %d", names(cc), as.integer(cc)), collapse = ", ")))
if (any(rs)) cat(sprintf(paste(
  " Of the %d fires coded RS, %d have more than 90 per cent of their mapped",
  "area inside Kosovo, so they are Kosovo fires filed under Serbia rather",
  "than fires straddling the border; they burnt %s ha. A query of EFFIS by",
  "country code KS alone misses them."),
  sum(rs), sum(rs & share_in > 0.9, na.rm = TRUE),
  fmt_int(sum(b$in_kosovo_ha[rs & share_in > 0.9], na.rm = TRUE))))
```

```{r}
#| label: map
#| eval: !expr have_cache
source("../R/functions.R", local = TRUE)
pal <- colorNumeric(c("#FFCB00", "#D50A00"), domain = range(b$year))
leaflet(st_transform(b, 4326)) |>
  add_basemaps() |>
  addPolygons(fillColor = ~pal(year), fillOpacity = 0.6, weight = 0.5,
              color = ~pal(year),
              label = ~sprintf("%s: %s ha in Kosovo (%s)", substr(FIREDATE, 1, 10),
                               fmt_int(in_kosovo_ha), COUNTRY)) |>
  add_boundary_outline(kosovo_outline()) |>
  addLegend(pal = pal, values = ~year, title = "Fire year",
            labFormat = labelFormat(big.mark = "")) |>
  addLayersControl(baseGroups = basemap_groups,
                   options = layersControlOptions(collapsed = TRUE))
```

```{r}
#| label: no-cache
#| eval: !expr "!have_cache"
#| output: asis
cat("::: {.callout-warning}\n## To be completed\nThe EFFIS web feature service did not answer at render time (",
    build_error %||% "no message", "). Re-render when it does.\n:::\n", sep = "")
```

## What it adds for Natura 2000 in Kosovo

Fire is a recurrent pressure on Kosovo's forests and grasslands; the NFI 2012
abstract counts over 12,000 ha affected. EFFIS gives a consistent, dated
record of where it happened, and how much of each fire was forest, which is
what an SDF's pressure section and a habitat's "future prospects" need. It
also shows which fires burnt inside existing protected areas. At the scale of
MODIS mapping it serves the [national tier](../natura2000.qmd#q1), with the
larger fires usable at site level.

## Caveats and sensitivity

- Small fires fall below the satellite's mapping threshold and are missing.
- Perimeters are mapped from imagery at a coarse resolution, not surveyed.
- No species locations; nothing sensitive.

## Integration into the pipeline

Low effort. A proposed `effis_burnt_areas(boundary)` in `R/functions.R` would
run the request above and write `data/kosovo_burnt_areas.gpkg`; the report's
protected-area section could then add burnt hectares per site. `pipeline.R`
is not changed.

## To be completed

```{r}
#| label: todo
#| output: asis
todo_list(row)
```

## Citation

::: {.callout-warning}
## To be completed
The citation EFFIS requests is shown with its terms at download and was not
recorded. Until then: European Forest Fire Information System (EFFIS),
burnt areas (`modis.ba.poly`), Copernicus Emergency Management Service,
accessed 29 September 2026.
:::
