---
title: "Emerald Network"
subtitle: "The Bern Convention's network of Areas of Special Conservation Interest; Kosovo has none of its own, but Serbia's candidate sites extend into it."
actions:
  - kind: check
    what: "What status, if any, the Bern Convention gives the Serbian candidate sites that lie in Kosovo (Mirusha, Gërmia, and the Kosovo parts of Prokletije and Šar); coe.int returned HTTP 403 to every request from this network"
    contact: "Bern Convention Secretariat, Council of Europe, https://www.coe.int/en/web/bern-convention/emerald-network-reference-portal"
  - kind: check
    what: "Why the 2025 catalogue record lists Kosovo (UNSCR 1244/99) among the areas covered when no site carries a Kosovo code"
    contact: "EEA catalogue record a2a18206-633a-4166-8850-0552661ef25f; ask the EEA helpdesk or the Bern Secretariat"
  - kind: check
    what: "Read the Emerald Standard Data Form explanatory notes, whose field guidance the desk review flagged as useful for Kosovo's own SDFs"
    contact: "Emerald Network reference portal on coe.int (blocked from this network; try another connection)"
todo:
  sensitivity: "Check whether the Emerald SDF tables withhold sensitive species; the vector file carries none"
---

```{r}
#| label: setup
source("../R/sources_helpers.R")
suppressPackageStartupMessages({
  library(sf)
  library(leaflet)
})
row <- source_row("emerald-network")
cache <- file.path(cache_dir("emerald-network"), "emerald_near_kosovo.gpkg")
service <- "https://bio.discomap.eea.europa.eu/arcgis/rest/services/ProtectedSites/EmeraldSites/MapServer/3"
```

::: {.source-status}
`r status_line(row)`
:::

## At a glance

```{r}
#| label: glance
metadata_table(row)
```

## What it is

The Emerald Network is the Bern Convention's counterpart of Natura 2000: a
network of Areas of Special Conservation Interest (ASCIs) that the
Convention's non-EU Parties propose, have assessed, and adopt. The data are
owned by the Council of Europe and processed and published by the EEA. The
2025 version, published on 19 May 2026 under CC-BY 4.0, reflects the
submissions the Standing Committee adopted in December 2025.

Kosovo is not a Party. Its own biodiversity action plan says so plainly:
"Kosovo is not a Party to the Bern Convention, meaning that the Emerald
Network has not been developed in the country" (MESP, Action Plan for
Biodiversity 2016–2020). The desk review therefore expected this page to be
"Not applicable to Kosovo". The check below shows why that is not quite
right.

## Access

```{r}
#| label: access
#| echo: true
#| code-fold: show
access_table(list(
  try_access("https://www.eea.europa.eu/en/datahub/datahubitem-view/4c4c8086-c940-400b-9064-29063143b2de"),
  try_access("https://doi.org/10.2909/a2a18206-633a-4166-8850-0552661ef25f"),
  try_access(paste0(service, "?f=json")),
  try_access("https://www.coe.int/en/web/bern-convention/emerald-network-reference-portal"),
  try_access("https://www.coe.int/en/web/bern-convention/emerald-viewer")
))
```

The EEA copy is open, with a DOI and a map service. The Council of Europe's
own reference portal and viewer returned HTTP 403 to every request from this
network, so the Convention's documents could not be read. The check uses the
EEA's map service, layer "All sites", in the same way the pipeline reads
CORINE Land Cover.

```{r}
#| label: build
#| echo: true
#| code-fold: true
#| code-summary: "Show how the sites around Kosovo are fetched"
build_error <- NULL
if (!file.exists(cache)) {
  build_error <- tryCatch({
    source("../R/functions.R", local = TRUE)
    ko   <- st_transform(st_union(kosovo_outline()), 3035)
    zone <- st_buffer(ko, 50000)
    sites <- arcgis_query_sf(service, st_bbox(st_transform(zone, 4326)),
                             fields = "SITECODE,SITENAME,COUNTRY_CODE,SITESTATUS,AREAHA,DATE_PROP_ASCI",
                             verbose = FALSE)
    sites <- st_make_valid(sites)
    sites <- sites[lengths(st_intersects(sites, zone)) > 0, ]
    sites$site_ha <- as.numeric(st_area(sites)) / 1e4
    sites$ha_in_kosovo <- as.numeric(vapply(seq_len(nrow(sites)), function(i)
      sum(as.numeric(st_area(st_intersection(st_geometry(sites)[i], ko)))), 0)) / 1e4
    sites$km_to_kosovo <- round(as.numeric(st_distance(sites, ko)) / 1000, 1)
    sites$proposed <- format(as.POSIXct(sites$DATE_PROP_ASCI / 1000,
                                        origin = "1970-01-01", tz = "UTC"), "%Y")
    st_write(st_intersection(sites, zone), cache, delete_dsn = TRUE, quiet = TRUE)
    NULL
  }, error = function(e) conditionMessage(e))
}
have_cache <- file.exists(cache)
```

```{r}
#| label: build-outcome
#| output: asis
if (!have_cache) {
  cat("::: {.callout-warning}\n## To be completed\nThe EEA map service did not answer at render time (",
      build_error %||% "no message", "). Re-render when it does.\n:::\n", sep = "")
} else {
  fp <- file_fingerprint(cache)
  cat(sprintf("The sites within 50 km of Kosovo, cut to that zone, are cached as `%s` (%s, SHA-256 `%s`, %s).\n",
              fp$file, fp$size, substr(fp$sha256, 1, 16), fp$downloaded))
}
```

## Kosovo coverage check

The question is whether any Emerald site lies in Kosovo, under which code, and
what lies within 50 km of the border.

```{r}
#| label: check
#| eval: !expr have_cache
s <- st_read(cache, quiet = TRUE)
inside <- s[s$ha_in_kosovo > 1, ]
knitr::kable(
  data.frame(Site = inside$SITENAME, Code = inside$SITECODE,
             `Country code` = inside$COUNTRY_CODE, Status = inside$SITESTATUS,
             `Proposed` = inside$proposed,
             `Mapped area (ha)` = fmt_int(round(inside$site_ha)),
             `Inside Kosovo (ha)` = fmt_int(round(inside$ha_in_kosovo)),
             `Share inside` = sprintf("%.0f %%", 100 * inside$ha_in_kosovo / inside$site_ha),
             check.names = FALSE)[order(-inside$ha_in_kosovo), ],
  align = "lllllrrr", row.names = FALSE)
```

```{r}
#| label: check-text
#| eval: !expr have_cache
#| output: asis
by_cc <- table(s$COUNTRY_CODE)
rs_in <- inside[inside$COUNTRY_CODE == "RS", ]
other <- inside[inside$COUNTRY_CODE != "RS", ]
kosovo_ha <- as.numeric(st_area(st_transform(st_union(kosovo_outline()), 3035))) / 1e4
cat(sprintf(paste(
  "%d Emerald sites lie within 50 km of Kosovo (%s). %d Serbian candidate",
  "sites, coded RS, reach inside Kosovo and cover %s ha of it, about %.0f per",
  "cent of the country. Another %d sites of Montenegro, Albania and North",
  "Macedonia overlap the border by %s ha in all; at a scale of 1:100,000 much",
  "of that is the difference between the Emerald outlines and the",
  "OpenStreetMap border. No site carries a Kosovo code (XK, XKX, KOS or KV)."),
  nrow(s), paste(sprintf("%s %d", names(by_cc), as.integer(by_cc)), collapse = ", "),
  nrow(rs_in), fmt_int(round(sum(rs_in$ha_in_kosovo))),
  100 * sum(rs_in$ha_in_kosovo) / kosovo_ha,
  nrow(other), fmt_int(round(sum(other$ha_in_kosovo)))))
```

The Serbian sites were proposed in 2006, before Kosovo's declaration of
independence, and the Emerald data still file them under Serbia. Mirusha and
Gërmia are named after places in Kosovo and lie inside it; the Kosovo parts
of Prokletije and Šar Planina take in the western and southern mountains
where both national parks lie. This is the "Kosovo merged into RS" pattern: a
user who filters the Emerald data by country counts Kosovo ground as Serbian
without knowing it, and any statistic computed per country, sufficiency
included, may do the same.

```{r}
#| label: map
#| eval: !expr have_cache
source("../R/functions.R", local = TRUE)
ko <- kosovo_outline()
pal <- colorFactor(c("#175CA1", "#D55E00", "#009E73", "#CC79A7"),
                   domain = sort(unique(s$COUNTRY_CODE)))
leaflet(st_transform(s, 4326)) |>
  add_basemaps() |>
  addPolygons(fillColor = ~pal(COUNTRY_CODE), fillOpacity = 0.35, weight = 1,
              color = ~pal(COUNTRY_CODE),
              label = ~sprintf("%s (%s, %s): %s ha inside Kosovo", SITENAME,
                               SITECODE, SITESTATUS, fmt_int(round(ha_in_kosovo)))) |>
  add_boundary_outline(ko) |>
  addLegend(pal = pal, values = ~COUNTRY_CODE, title = "Country code") |>
  addLayersControl(baseGroups = basemap_groups,
                   options = layersControlOptions(collapsed = TRUE))
```

## What it adds for Natura 2000 in Kosovo

Two things. First, the neighbours' candidate sites along the border, which a
Kosovo proposal should meet rather than stop short of. Second, a warning for
every European dataset used in this section: Kosovo's territory can be
present under another country's code, and a check by code alone would miss
it. Emerald boundaries are drawn at 1:100,000, so they serve the site tier
only as a first outline ([tiers](../natura2000.qmd#q1)). The Emerald
Standard Data Form is structurally close to the Natura 2000 SDF, and its
guidance is worth reading when Kosovo designs its own forms; that reading is
an open action, since coe.int was blocked.

## Caveats and sensitivity

- Candidate status is the first of the Emerald stages; none of these border
  sites is adopted.
- The Serbian sites inside Kosovo date from 2006–2008 and describe data of
  that time.
- The status of those sites under the Convention could not be checked,
  because coe.int was unreachable.

## Integration into the pipeline

Low effort. `arcgis_query_sf()` already exists in `R/functions.R`; a proposed
`emerald_sites_near(boundary, radius_km = 50)` called after the
protected-area step would add the layer, with `ha_in_kosovo` and
`COUNTRY_CODE` kept so the report can say which sites are filed under
another country. `pipeline.R` is not changed here.

## To be completed

```{r}
#| label: todo
#| output: asis
todo_list(row)
```

## Citation

The catalogue record identifies the dataset as
`coe_v_3035_100_k_emerald_p_2024-2025_v01_r00`, copyright Council of Europe,
licence CC-BY 4.0. Cite it as: *Emerald Network data (vector) – the
Pan-European network of protected sites, version 2025.* Council of Europe;
published by the European Environment Agency, 2026.
[doi:10.2909/a2a18206-633a-4166-8850-0552661ef25f](https://doi.org/10.2909/a2a18206-633a-4166-8850-0552661ef25f).
