---
title: "Atlas of European Mammals, second edition"
subtitle: "The EMMA2 atlas and its database of European mammal records, due for publication a week after this check."
actions:
  - kind: check
    what: "After publication (6 October 2026), confirm whether Kosovo is mapped in EMMA2 and at what grid"
    contact: "https://www.european-mammals.org/"
  - kind: check
    what: "Whether the EMMA2 grid database is, or will be, published on GBIF; if so, compare it with the existing download to avoid counting records twice"
    contact: "https://www.gbif.org/dataset/search?q=European%20mammals"
  - kind: check
    what: "Whether the GBIF dataset 'Small mammals of Yugoslavia 1992 atlas' holds records inside Kosovo, under which country code, and whether the report's download includes them"
    contact: "https://www.gbif.org/dataset/search?q=Small%20mammals%20of%20Yugoslavia"
  - kind: request
    what: "EMMA2 grid records for the squares covering Kosovo, if the database is not published"
    contact: "EMMA2 project, https://www.european-mammals.org/"
todo:
  doi: "No DOI for the database; check after publication"
  licence: "Terms of the EMMA2 database; not stated on the project site"
  access: "How the database can be obtained; not stated on the project site"
  format: "Grid records; format unknown until released"
  access_method: "Unknown until released"
  spatial_resolution: "Grid size used by EMMA2; not stated on the pages opened (the desk review said 50 km)"
  temporal_coverage: "Years of records compiled for EMMA2; read in the book or project documentation"
  kosovo_coverage: "The desk review reported Kosovo included for the first time; not stated on the pages opened"
  kosovo_codes_seen: "How EMMA2 labels Kosovo's squares"
  tier: "National at a 50 km grid, if that grid is confirmed"
  effort: "Low if the data reach GBIF; estimate otherwise"
  sensitivity: "Whether EMMA2 generalises bat roosts or large carnivores; read the data policy"
---

```{r}
#| label: setup
source("../R/sources_helpers.R")
row <- source_row("emma2")
cache <- file.path(cache_dir("emma2"), "gbif_dataset_search.csv")
```

::: {.source-status}
`r status_line(row)`
:::

## At a glance

```{r}
#| label: glance
metadata_table(row)
```

## What it is

EMMA2, the second *Atlas of European Mammals*, maps the distribution of
Europe's mammals on a common grid. Its website gives the book's publication
date as 6 October 2026, one week after this check, with CRC Press, edited by
Mitchell-Jones, Amori, Kryštufek and colleagues. It covers the area of the
first atlas plus Moldova, Belarus and Ukraine, with Russia excluded. As of
June 2024 the project database held 142,965 records of 245 species.

The desk review reported that Kosovo is included for the first time. The
pages opened in this check do not say so, and that remains to be confirmed
from the published atlas.

## Access

```{r}
#| label: access
#| echo: true
#| code-fold: show
access_table(list(try_access("https://european-mammals.org/")))
```

The project site describes the atlas and its database but gives no route to
the data: no download, no GBIF link and no request form.

## Kosovo coverage check

The executable part of the check is whether EMMA2 records are already on
GBIF, which would mean the report's GBIF extract might contain them.

```{r}
#| label: build
#| echo: true
#| code-fold: true
#| code-summary: "Show the GBIF dataset search"
#| output: asis
if (!file.exists(cache)) {
  q <- c("EMMA2", "Atlas of European Mammals", "European Mammal Foundation")
  hits <- do.call(rbind, lapply(q, function(x) {
    r <- tryCatch(jsonlite::fromJSON(paste0(
      "https://api.gbif.org/v1/dataset/search?limit=20&q=", utils::URLencode(x))),
      error = function(e) NULL)
    if (is.null(r) || !r$count) return(data.frame(query = x, title = NA, publisher = NA))
    data.frame(query = x, title = r$results$title,
               publisher = r$results$publishingOrganizationTitle)
  }))
  write.csv(hits, cache, row.names = FALSE)
}
hits <- read.csv(cache)
hits <- hits[!is.na(hits$title), ]
mammal <- hits[grepl("mammal", hits$title, ignore.case = TRUE), ]
is_emma2 <- grepl("EMMA2|Atlas of European Mammals", mammal$title)
cat(sprintf(paste("The three searches returned %d dataset titles. %d of them",
                  "concern mammals, and %s.\n\n"),
            nrow(hits), nrow(mammal),
            if (any(is_emma2)) "one of those is the EMMA2 atlas" else "none of those is the EMMA2 atlas"))
if (nrow(mammal)) print(knitr::kable(unique(mammal[, c("title", "publisher")]),
                                     col.names = c("Mammal dataset returned", "Publisher"),
                                     row.names = FALSE))
```

No EMMA2 dataset was found on GBIF on 29 September 2026. Besides the
mammal datasets listed, the searches return unrelated datasets that merely
share the name, such as French observers' personal datasets and a Madrid
herbarium called EMMA.

One of the mammal datasets deserves a second look for a different reason:
"Small mammals of Yugoslavia 1992 atlas", published to GBIF by the National
Museum of Natural History in Sofia. An atlas of Yugoslavia from 1992 would
have covered Kosovo, and its records may carry a Yugoslav or Serbian country
code rather than XK. Whether they fall inside Kosovo's outline, and whether
the report's download already holds them, is listed below as an open check.

::: {.callout-warning}
## To be completed
After 6 October 2026: confirm from the atlas whether Kosovo's squares are
mapped, and if the database is released, list the species per square
intersecting `kosovo_outline()`, set them against the Annex II and IV mammals
in the GBIF extract, and check for records already present through GBIF.
:::

## What it adds for Natura 2000 in Kosovo

Mammals on Annexes II and IV, bats above all, are thinly recorded in Kosovo's
GBIF data. A systematic atlas grid gives a presence baseline against which
gaps can be judged, and a cross-check for species reported from only a
handful of records. At the atlas grid it serves the [national screening
tier](../natura2000.qmd#q1).

## Caveats and sensitivity

- The atlas is not yet published, and its data route is unknown.
- Kosovo's inclusion is reported by the desk review, not confirmed here.
- Bat roosts and large carnivores need generalisation if finer data are ever
  supplied.

## Integration into the pipeline

If EMMA2 appears on GBIF, nothing new is needed beyond deduplicating against
the existing download. If it arrives as a grid table, the proposed
`read_atlas_squares()` on the [EBBA2 page](ebba2.qmd) applies. `pipeline.R`
is not changed.

## To be completed

```{r}
#| label: todo
#| output: asis
todo_list(row)
```

## Citation

As the project site gives it: Mitchell-Jones, A. J., Amori, G., Kryštufek,
B., Marnell, F., Miteva, S., Sallay, A., Schley, L., Thissen, J. &
Vohralík, V. (eds) (2026). *Atlas of European Mammals, 2nd edition.* CRC
Press, Boca Raton, FL.
