---
title: "Conservation assessment of the endemic plants from Kosovo (Hacquetia, 2017)"
subtitle: "An IUCN assessment of 16 endemic plant taxa of Kosovo, most of them confined to serpentine or limestone."
actions:
  - kind: check
    what: "List the 16 taxa from the PDF with their categories, match them to the GBIF backbone, and count their records in the GBIF extract"
    contact: "https://ojs.zrc-sazu.si/hacquetia/article/view/4787/4403"
  - kind: check
    what: "Whether the localities behind the assessments exist as a dataset, and whether the authors would share them at a generalised resolution"
    contact: "The paper's authors (Millaku et al.), through the journal"
todo:
  temporal_coverage: "Years of the field observations behind the assessments; read in the paper's methods"
  spatial_resolution: "Whether the paper gives localities, coordinates or maps; the abstract page does not say"
  effort: "Low for the species list (16 names); higher for any localities"
  sensitivity: "Stenoendemics with tiny populations (Solenanthus krasniqii: 20 mature individuals) must not be mapped finer than 10 km"
---

```{r}
#| label: setup
source("../R/sources_helpers.R")
row <- source_row("endemic-plants-hacquetia")
```

::: {.source-status}
`r status_line(row)`
:::

## At a glance

```{r}
#| label: glance
metadata_table(row)
```

## What it is

Millaku, Krasniqi, Berisha and Rexhepi assessed the conservation status of 16
plant taxa endemic to Kosovo or the Balkans against the IUCN Red List
categories and criteria (guidelines version 8.1). The paper appeared in
*Hacquetia* 16(1) in 2017 under a CC BY-NC-ND 4.0 licence. Of the 16 taxa,
four are Balkan endemics, eight local endemics and four stenoendemics. Six
grow only on serpentine, five on limestone, four on carbonate substrates and
one on silicate. The most threatened, *Solenanthus krasniqii*, was reduced to
about 20 mature individuals.

## Access

```{r}
#| label: access
#| echo: true
#| code-fold: show
access_table(list(
  try_access("https://ojs.zrc-sazu.si/hacquetia/article/view/4787"),
  try_access("https://ojs.zrc-sazu.si/hacquetia/article/view/4787/4403"),
  try_access("https://doi.org/10.1515/hacq-2016-0024")
))
```

The article and its PDF are open. The licence permits sharing with
attribution for non-commercial use, but not adapted versions, so a species
list derived from it should be cited rather than presented as new work.

## Kosovo coverage check

::: {.callout-warning}
## To be completed
The paper is about Kosovo, so coverage is not in doubt. The useful check has
not been run: extract the 16 names and categories from the PDF, match them to
the GBIF backbone with `rgbif::name_backbone_checklist()`, and count how many
records of each the GBIF extract holds. A stenoendemic with no GBIF record is
invisible to the report's species tables, however threatened it is.
:::

## What it adds for Natura 2000 in Kosovo

Endemic plants of serpentine and limestone mark exactly the habitats, such as
serpentine grasslands and rocky slopes, that a Kosovo site list would need to
cover. Several may not be on Annex II, but a national assessment of their
status is evidence for choosing between candidate sites that hold the same
Annex I habitat. It serves the [national tier](../natura2000.qmd#q1); with
localities, it could inform site boundaries.

## Caveats and sensitivity

- Sixteen taxa is a small, focused set, not a flora-wide list; the Red Book
  of Vascular Flora (2013) covers 237.
- National assessments made in 2017 may be out of date.
- Stenoendemics with a few dozen individuals are at risk from collecting and
  must never be mapped precisely in public.

## Integration into the pipeline

Low effort for the names: a proposed `data/kosovo_endemic_plants.csv` with
name, category, source DOI and GBIF key, read by `pipeline.R` like the annex
lists, would flag endemic records in every export. `pipeline.R` is not
changed here.

## To be completed

```{r}
#| label: todo
#| output: asis
todo_list(row)
```

## Citation

Millaku, F., Krasniqi, E., Berisha, N. & Rexhepi, F. (2017). Conservation
assessment of the endemic plants from Kosovo. *Hacquetia* 16(1).
[doi:10.1515/hacq-2016-0024](https://doi.org/10.1515/hacq-2016-0024).
