---
title: "EU-Hydro"
subtitle: "The Copernicus river network of Europe at 1:50,000, whose own metadata say Kosovo has no water-body data, yet whose rivers run through it."
actions:
  - kind: check
    what: "Ask the EEA or CLMS what 'no spatial data for any category of water bodies was available' means for Kosovo, given that the river network and inland-water layers do reach it"
    contact: "Copernicus Land Monitoring Service helpdesk"
  - kind: check
    what: "Compare EU-Hydro's river lines with OpenStreetMap waterways and with HydroRIVERS in Kosovo, to judge positional accuracy"
    contact: "The HydroSHEDS page in this section"
todo: {}
---

```{r}
#| label: setup
source("../R/sources_helpers.R")
suppressPackageStartupMessages({
  library(sf)
  library(leaflet)
})
row <- source_row("eu-hydro")
cache <- file.path(cache_dir("eu-hydro"), "eu_hydro_kosovo.gpkg")
svc <- "https://image.discomap.eea.europa.eu/arcgis/rest/services/EUHydro/EUHydro_RiverNetworkDatabase/MapServer"
strahler_layers <- c(`3` = 7, `4` = 8, `5` = 9, `6` = 10, `7` = 11)
```

::: {.source-status}
`r status_line(row)`
:::

## At a glance

```{r}
#| label: glance
metadata_table(row)
```

## What it is

EU-Hydro is the Copernicus Land Monitoring Service's photo-interpreted river
network for the EEA38 countries and the United Kingdom: river lines ordered by
Strahler number, inland water polygons, and a drainage model derived from the
EU-DEM. Version 1.3 (November 2020) is mapped at 1:50,000 with a 1 ha minimum
mapping unit, has a DOI, and is free and open under the Copernicus data
policy, with attribution.

Its metadata contain a sentence that bears directly on Kosovo: for Türkiye,
Albania, Bosnia-Herzegovina, North Macedonia, Montenegro, Kosovo and Serbia,
"no spatial data for any category of water bodies was available". Taken at
face value that would make the product useless here. The check below shows
it is not.

## Access

```{r}
#| label: access
#| echo: true
#| code-fold: show
access_table(list(
  try_access("https://sdi.eea.europa.eu/catalogue/srv/api/records/393359a7-7ebd-4a52-80ac-1a18d5f3db9c"),
  try_access(paste0(svc, "?f=json"))
))
```

Downloads are per major basin as File Geodatabase or GeoPackage; the check
uses the EEA's map service instead, which needs no download.

```{r}
#| label: build
#| echo: true
#| code-fold: true
#| code-summary: "Show the map-service query"
build_error <- NULL
if (!file.exists(cache)) {
  build_error <- tryCatch({
    source("../R/functions.R", local = TRUE)
    ko <- st_transform(st_union(kosovo_outline()), 3035)
    bb <- st_bbox(st_transform(ko, 4326))
    rivers <- do.call(rbind, lapply(names(strahler_layers), function(o) {
      # A layer with nothing in the box, or nothing left after clipping, is
      # skipped rather than allowed to stop the build.
      x <- tryCatch(arcgis_query_sf(paste0(svc, "/", strahler_layers[[o]]), bb,
                                    fields = "OBJECTID", verbose = FALSE),
                    error = function(e) NULL)
      if (is.null(x)) return(NULL)
      x <- suppressWarnings(st_intersection(st_make_valid(x), ko))
      x <- suppressWarnings(st_collection_extract(x, "LINESTRING"))
      if (!nrow(x)) return(NULL)
      st_sf(strahler = rep(as.integer(o), nrow(x)), geometry = st_geometry(x))
    }))
    water <- arcgis_query_sf(paste0(svc, "/2"), bb, fields = "OBJECTID", verbose = FALSE)
    water <- suppressWarnings(st_intersection(st_make_valid(water), ko))
    st_write(rivers, cache, layer = "rivers", delete_dsn = TRUE, quiet = TRUE)
    st_write(st_sf(geometry = st_geometry(water)), cache, layer = "inland_water",
             append = FALSE, quiet = TRUE)
    NULL
  }, error = function(e) conditionMessage(e))
}
have_cache <- file.exists(cache)
```

```{r}
#| label: build-failed
#| eval: !expr "!have_cache"
#| output: asis
cat("::: {.callout-warning}\n## To be completed\nThe map-service query failed at render time (",
    build_error %||% "no message", "). Re-render when the EEA service answers.\n:::\n", sep = "")
```

## Kosovo coverage check

```{r}
#| label: check
#| eval: !expr have_cache
rivers <- st_read(cache, layer = "rivers", quiet = TRUE)
water  <- st_read(cache, layer = "inland_water", quiet = TRUE)
len <- tapply(as.numeric(st_length(rivers)) / 1000, rivers$strahler, sum)
knitr::kable(data.frame(`Strahler order` = names(len),
                        `River length inside Kosovo (km)` = fmt_int(len),
                        check.names = FALSE), align = "rr", row.names = FALSE)
```

```{r}
#| label: check-text
#| eval: !expr have_cache
#| output: asis
cat(sprintf(paste(
  "Inside Kosovo the network holds %s km of rivers of Strahler order 3 and",
  "above, and %d inland-water polygons covering %s ha. Kosovo is therefore",
  "partly covered: the river network and some water bodies are mapped,",
  "whatever the metadata say. The service carries no country attribute on",
  "these layers, so no country code was checked."),
  fmt_int(sum(len)), nrow(water), fmt_int(sum(as.numeric(st_area(water))) / 1e4)))
```

```{r}
#| label: map
#| eval: !expr have_cache
source("../R/functions.R", local = TRUE)
pal <- colorNumeric(c("#9ECAE1", "#08306B"), domain = rivers$strahler)
leaflet(st_transform(rivers, 4326)) |>
  add_basemaps() |>
  addPolylines(color = ~pal(strahler), weight = ~strahler / 1.5, opacity = 0.9,
               label = ~paste("Strahler order", strahler)) |>
  addPolygons(data = st_transform(water, 4326), color = "#0072B2", weight = 1,
              fillOpacity = 0.5) |>
  add_boundary_outline(kosovo_outline()) |>
  addLegend(pal = pal, values = ~strahler, title = "Strahler order") |>
  addLayersControl(baseGroups = basemap_groups,
                   options = layersControlOptions(collapsed = TRUE))
```

## What it adds for Natura 2000 in Kosovo

Freshwater habitats (3260 water courses, 91E0 alluvial forests, 92A0 willow
and poplar galleries) and Annex II fish follow the river network, and a
Strahler order separates headwater streams from the main rivers. River lines
are also natural site boundaries. EU-Hydro is one network for all the EEA's
member and cooperating countries, which makes it a natural choice for a Kosovo
network meant to join its neighbours'. At 1:50,000
it serves the [national tier](../natura2000.qmd#q1) and outlines for sites.

## Caveats and sensitivity

- The metadata's statement about Kosovo conflicts with what the service
  returns; which parts were derived without national data is unknown.
- Lines of Strahler order 1 and 2 were left out of this check to keep it
  light; they are in the service.
- No sensitive content.

## Integration into the pipeline

Low effort. A proposed `kosovo_rivers(boundary, min_strahler = 3)` in
`R/functions.R`, using the existing `arcgis_query_sf()`, would cache the
network for the report's maps and for the freshwater part of the habitat
screen. `pipeline.R` is not changed.

## To be completed

```{r}
#| label: todo
#| output: asis
todo_list(row)
```

## Citation

*EU-Hydro River Network Database 2006–2012 (vector), Europe – version 1.3,
Nov. 2020.* European Environment Agency / Copernicus Land Monitoring Service.
[doi:10.2909/393359a7-7ebd-4a52-80ac-1a18d5f3db9c](https://doi.org/10.2909/393359a7-7ebd-4a52-80ac-1a18d5f3db9c).
