---
title: "Euro+Med PlantBase"
subtitle: "The reference checklist of the vascular plants of Europe and the Mediterranean, which has treated Kosovo as its own area since 2019."
actions:
  - kind: check
    what: "Read europlusmed.org in a browser (it served a bot-protection page to automated reads) and record the licence, the current version and how area data can be exported"
    contact: "https://europlusmed.org/"
  - kind: check
    what: "For the Annex II and IV plants in the GBIF extract, check whether Euro+Med gives 'Ko' in their distribution, and flag those that are recorded in Kosovo by GBIF but not by Euro+Med (or the reverse)"
    contact: "Euro+Med taxon pages; data/eu_directives_species.csv"
todo:
  doi: "The database has no DOI seen; its Notulae in Willdenowia do"
  licence: "Terms of use of the Euro+Med data; the site could not be read automatically"
  access: "Whether area data can be downloaded or only viewed per taxon"
  format: "Export formats, if any"
  access_method: "Web pages per taxon; check for an API or bulk export"
  spatial_resolution: "Political and geographical areas (e.g. 'Ko'), not coordinates"
  temporal_coverage: "Not applicable to a checklist; record the version date instead"
  latest_version: "Version or date of the current online database; not readable automatically"
  effort: "Medium: area codes per taxon need scraping unless an export exists"
  sensitivity: "None: presence per country"
---

```{r}
#| label: setup
source("../R/sources_helpers.R")
row <- source_row("euro-med-plantbase")
```

::: {.source-status}
`r status_line(row)`
:::

## At a glance

```{r}
#| label: glance
metadata_table(row)
```

## What it is

Euro+Med PlantBase is the checklist of the vascular plants of Europe and the
Mediterranean, recording for each accepted taxon the areas in which it occurs.
The desk review asked whether it treats Kosovo as an area of its own or folds
it into Serbia. The answer is in the Euro+Med-Checklist Notulae published in
*Willdenowia* in 2019 (Raab-Straube & Raus, eds), where the Euro+Med PlantBase
Secretariat announced that it would use "Ko" for the Republic of Kosovo and
"Se" for the Republic of Serbia "instead of Sr (Serbia with Kosovo)", as
already practised in *Atlas Florae Europaeae*.

## Access

```{r}
#| label: access
#| echo: true
#| code-fold: show
access_table(list(
  try_access("https://europlusmed.org/"),
  try_access("https://europlusmed.org/explanations"),
  try_access("https://doi.org/10.3372/wi.49.49312")
))
```

The site answers HTTP 200 but returns a bot-protection page ("access denied")
to automated reads, so its own explanation of area codes, its licence and its
version could not be read. The Willdenowia note, published by the checklist's
own editors, is the primary source used here.

## Kosovo coverage check

::: {.callout-warning}
## To be completed
The area scheme includes Kosovo ("Ko") since 2019, so Kosovo is covered in
principle. How completely each taxon's Kosovo occurrence has been entered is
another matter: records that predate the split may still sit under "Sr" or be
missing from "Ko". Check a sample, starting with the Annex II and IV plants
the GBIF extract finds in Kosovo, and record how many carry "Ko".
:::

## What it adds for Natura 2000 in Kosovo

Whether a plant occurs in a country is a question an SDF, a sufficiency ledger
and a reviewer all ask. Euro+Med is the reference answer for Europe's flora;
a GBIF record of an annex plant in Kosovo that Euro+Med does not list for "Ko"
is either a new national record, worth documenting, or a misidentification,
worth a specialist's look. It serves the [national tier](../natura2000.qmd#q1).

## Caveats and sensitivity

- Area codes were changed in 2019; older content may lag behind.
- Presence per area says nothing about where in Kosovo a plant occurs.
- No sensitive content.

## Integration into the pipeline

Medium effort without an export. A proposed `euromed_areas(taxon_names)`
would look up each directive-listed plant and record whether "Ko" is in its
distribution, as a column in `data/eu_directives_species.csv`. `pipeline.R`
is not changed.

## To be completed

```{r}
#| label: todo
#| output: asis
todo_list(row)
```

## Citation

For the area codes: Raab-Straube, E. von & Raus, Th. (eds) (2019).
Euro+Med-Checklist Notulae, 11. *Willdenowia* 49(3): 421–445.
[doi:10.3372/wi.49.49312](https://doi.org/10.3372/wi.49.49312).

::: {.callout-warning}
## To be completed
The citation Euro+Med requests for the database itself is on its site, which
could not be read automatically.
:::
