---
title: "European atlases of amphibians, reptiles, butterflies, dragonflies, bats and orchids"
actions:
  - kind: check
    what: "Amphibians and reptiles: identify the current European atlas and its data (the desk review named none), and whether Kosovo's squares are filled"
    contact: "Societas Europaea Herpetologica"
  - kind: check
    what: "Butterflies: identify the European butterfly distribution atlas or database and whether Kosovo is covered"
    contact: "Butterfly Conservation Europe"
  - kind: check
    what: "Dragonflies: identify the European atlas of dragonflies and damselflies and its database, and Kosovo's coverage"
    contact: "The atlas's publisher or the European dragonfly society"
  - kind: check
    what: "Bats and orchids: identify European distribution atlases or databases, and whether Kosovo is a mapped unit or folded into Serbia"
    contact: "EUROBATS for bats; a European orchid mapping project, if any, for orchids"
  - kind: check
    what: "For each atlas found, check whether its data are on GBIF (and in this project's download) before requesting them"
    contact: "https://www.gbif.org/dataset/search"
todo:
  custodian: "One custodian per atlas; identify each (see actions)"
  url: "One primary URL per atlas; none was confirmed by the desk review"
  doi: "Record the DOI of each atlas or its dataset once identified"
  licence: "Terms of each atlas database"
  access: "Open, request or paid, per atlas"
  format: "Per atlas"
  access_method: "Per atlas"
  spatial_resolution: "Grid of each atlas (often 50 km UTM or 10 km)"
  temporal_coverage: "Recording period of each atlas"
  latest_version: "Edition or year of each atlas"
  kosovo_coverage: "Whether each atlas maps Kosovo's squares"
  kosovo_codes_seen: "Whether each atlas files Kosovo separately or under Serbia"
  tier: "National at atlas-grid resolution, once confirmed"
  n2k_question: "Distribution of Annex II and IV amphibians, reptiles, butterflies, dragonflies and bats, and of orchid-rich grassland indicators (6210*), once content is confirmed"
  effort: "Per atlas, once formats are known"
  sensitivity: "Orchid and reptile locations attract collectors; atlas grids protect them only if kept at native resolution"
---

```{r}
#| label: setup
source("../R/sources_helpers.R")
row <- source_row("european-taxon-atlases")
```

::: {.source-status}
`r status_line(row)`
:::

## At a glance

```{r}
#| label: glance
metadata_table(row)
```

## What it is

::: {.callout-warning}
## To be completed
The desk review grouped five kinds of European distribution atlas under one
lead, and confirmed no URL for any of them. Nothing on this page has yet been
verified. Each atlas should be identified at its publisher, then either given
its own page in this section or recorded here as absent for Kosovo.
:::

Taxon atlases matter because they are the one systematic record for groups
that casual recording covers badly. For Kosovo, where GBIF data are dominated
by birds and by a few well-visited sites, an atlas grid would show presence
across the country for groups that carry many Annex II and IV species:
amphibians and reptiles, butterflies, dragonflies and bats. Orchids are the
indicator the Habitats Directive itself uses to mark priority dry grassland
(6210\*, "important orchid sites").

## Access

No access was attempted: no URL was available to try, and none was guessed.

## Kosovo coverage check

::: {.callout-warning}
## To be completed
For each atlas, once identified: does it map Kosovo, at which grid, and under
which name or code? Several pan-European atlases were compiled before 2008
and may show Kosovo within Serbia (look for "Serbia and Montenegro", RS/SRB or
YU). Then check GBIF for the atlas's dataset and compare with the report's
download, as the [EBBA2 page](ebba2.qmd) does for the first bird atlas.
:::

## What it adds for Natura 2000 in Kosovo

Potentially a great deal for the Habitats Directive side: presence grids for
the invertebrates and herptiles on Annex II, for which the report's GBIF data
are thinnest. That is one of the data gaps listed on the
[hub page](index.qmd#data-gaps). The [tier](../natura2000.qmd#q1) would be
national screening at atlas-grid resolution.

## Caveats and sensitivity

- Unverified lead: nothing here is established.
- Atlas grids are coarse by design and must be shown at their native
  resolution.

## Integration into the pipeline

Unknown until the atlases are identified. If they are on GBIF, a new download
brings them in; otherwise the grid-table reader proposed on the
[EBBA2 page](ebba2.qmd) applies. `pipeline.R` is not changed.

## To be completed

```{r}
#| label: todo
#| output: asis
todo_list(row)
```

## Citation

::: {.callout-warning}
## To be completed
Cite each atlas as its publisher requests, once identified.
:::
