---
title: "European Red Lists"
subtitle: "The European Commission's European and EU-level red list assessments of over 10,000 species, published as tables on the EEA datahub."
actions:
  - kind: check
    what: "Download the current tables and check whether any assessment records Kosovo as a country of occurrence, or Serbia with Kosovo folded in"
    contact: "https://www.eea.europa.eu/en/datahub/datahubitem-view/202f3c2e-54a9-4ff4-a1da-ed7ca524f634"
  - kind: check
    what: "Read the IUCN pages on the European Red List in a browser; iucnredlist.org served a Cloudflare challenge (HTTP 403) to automated reads"
    contact: "https://www.iucnredlist.org/regions/european-red-list"
  - kind: check
    what: "Whether Kosovo's plant red list (2013) is among the national lists compiled in the 2024 database of European vascular plant red lists, and in what form"
    contact: "Lončarević et al. 2024, Scientific Data, doi:10.1038/s41597-024-03963-0 (supplementary list of sources)"
todo:
  doi: "DOI of the EEA dataset; not shown on the datahub page read"
  licence: "Licence of the EEA tables; not shown on the datahub page read"
  spatial_resolution: "Not spatial: status per species at European and EU level"
  temporal_coverage: "Assessment years per group, 2009–2022"
  kosovo_coverage: "Whether assessments list Kosovo as a country of occurrence; check in the tables"
  kosovo_codes_seen: "Country codes used for Kosovo in the tables, if any"
  effort: "Low: a species table joined on name or GBIF key"
  sensitivity: "None: status per species"
---

```{r}
#| label: setup
source("../R/sources_helpers.R")
row <- source_row("european-red-lists")
```

::: {.source-status}
`r status_line(row)`
:::

## At a glance

```{r}
#| label: glance
metadata_table(row)
```

## What it is

The European Red Lists assess the extinction risk of more than 10,000
European species, from mammals and birds to bees, grasshoppers and ferns,
under the IUCN guidelines for regional assessment, at two levels: the EU and
pan-Europe. The EEA publishes the results as tables on its datahub; the
current series covers assessments from 2009 to 2022, and the dataset's owner
is the European Commission's Directorate-General for Environment.

For Kosovo, the pan-European level is the relevant one, and whether its
assessments count Kosovo as a country of occurrence is the open question.
A related finding bears on it. The 2024 *Scientific Data* database of European
national red lists for vascular plants (Lončarević et al.) treats Kosovo as a
country with a red list of its own. Its authors removed the Kosovo taxa from
Serbia's red book of 1999, and did not add them to the Kosovo list.

## Access

```{r}
#| label: access
#| echo: true
#| code-fold: show
access_table(list(
  try_access("https://www.eea.europa.eu/en/datahub/datahubitem-view/202f3c2e-54a9-4ff4-a1da-ed7ca524f634"),
  try_access("https://www.iucnredlist.org/regions/european-red-list"),
  try_access("https://www.nature.com/articles/s41597-024-03963-0")
))
```

The EEA page answers; its tables were not downloaded in this check. IUCN's own
page returned a Cloudflare challenge to every automated request.

## Kosovo coverage check

::: {.callout-warning}
## To be completed
Download the EEA tables, look for a country-of-occurrence field, and record
whether Kosovo appears, under which code, or whether it is folded into
Serbia. Then join the European categories to the species in the GBIF extract
by GBIF key, as the pipeline does for the global Red List.
:::

## What it adds for Natura 2000 in Kosovo

The report already carries each species' global IUCN category. A European
category is often more telling for Natura 2000: a species of Least Concern
globally can be threatened in Europe, and the Habitats Directive is a
European instrument. Where the assessments list countries of occurrence, they
also give a European view of which species occur in Kosovo. It serves the
[national tier](../natura2000.qmd#q1).

## Caveats and sensitivity

- Assessment years range over more than a decade; categories may be dated.
- European and EU27 categories differ; the report should say which it uses.
- No sensitive content.

## Integration into the pipeline

Low effort. A proposed `fetch_european_red_list(path)` would read the EEA
table and join it by GBIF key, adding `europeanRedListCategory` beside the
existing `iucnRedListCategory`, just as `fetch_iucn_categories()` does for the
global list. `pipeline.R` is not changed.

## To be completed

```{r}
#| label: todo
#| output: asis
todo_list(row)
```

## Citation

::: {.callout-warning}
## To be completed
The citation for the EEA tables was not shown on the page read; take it from
the dataset's catalogue record once the tables are downloaded.
:::
