---
title: "Fauna Europaea"
subtitle: "The European checklist of animal species; its portal did not resolve, and its GBIF copy carries names without distribution areas."
actions:
  - kind: check
    what: "Find the current home of Fauna Europaea; fauna-eu.org did not resolve (DNS) on 29 September 2026"
    contact: "A web search for the Fauna Europaea portal, or the PESI species infrastructure, which draws on it"
  - kind: check
    what: "Whether Fauna Europaea's distribution scheme has a Kosovo area or files Kosovo records under Serbia"
    contact: "Fauna Europaea documentation, once the portal is found"
todo:
  custodian: "Current host of Fauna Europaea; the domain did not resolve and the GBIF record's publisher was not read"
  spatial_resolution: "Presence per country or region, according to the desk review; confirm in the documentation"
  temporal_coverage: "Not applicable to a checklist; record the version date instead"
  kosovo_coverage: "Unknown: the GBIF copy exposes no distribution areas, and the portal is unreachable"
  kosovo_codes_seen: "Whether Kosovo is an area of its own (the desk review expected it folded into Serbia)"
  tier: "National at best: presence per country"
  n2k_question: "Whether an animal is recorded for Kosovo, as a check on range statements for Annex II and IV animals"
  effort: "Unknown until the distribution data can be reached"
  sensitivity: "None: presence per country"
---

```{r}
#| label: setup
source("../R/sources_helpers.R")
row <- source_row("fauna-europaea")
key <- "90d9e8a6-0ce1-472d-b682-3451095dbc5a"
```

::: {.source-status}
`r status_line(row)`
:::

## At a glance

```{r}
#| label: glance
metadata_table(row)
```

## What it is

Fauna Europaea is a checklist of Europe's animals that records which species
occur in which European country. The
desk review asked whether it treats Kosovo as an area of its own or folds it
into Serbia. That question could not be answered in this check. The project's
web domain did not resolve, and the copy of the checklist published to GBIF
(DOI 10.15468/ymk1bx, dated 9 September 2016, licence "unspecified") holds
names without the distribution data that would show Kosovo.

## Access

```{r}
#| label: access
#| echo: true
#| code-fold: show
access_table(list(
  try_access("https://fauna-eu.org/"),
  try_access(paste0("https://api.gbif.org/v1/dataset/", key))
))
```

## Kosovo coverage check

The executable check asks the GBIF copy for the distribution records of a few
well-known species that occur in Kosovo.

```{r}
#| label: check
#| echo: true
#| code-fold: true
probe <- c("Rupicapra rupicapra", "Ursus arctos", "Salamandra salamandra",
           "Lycaena dispar", "Rosalia alpina")
res <- do.call(rbind, lapply(probe, function(nm) tryCatch({
  s <- jsonlite::fromJSON(paste0("https://api.gbif.org/v1/species/search?datasetKey=",
                                 key, "&limit=1&q=", utils::URLencode(nm)))$results
  d <- jsonlite::fromJSON(sprintf("https://api.gbif.org/v1/species/%s/distributions?limit=200",
                                  s$key[1]))$results
  areas <- if (NROW(d) && "locality" %in% names(d)) sum(!is.na(d$locality)) else 0
  data.frame(Species = nm, `Name found` = "yes", `Distribution areas` = areas,
             check.names = FALSE)
}, error = function(e) data.frame(Species = nm, `Name found` = "no",
                                  `Distribution areas` = NA, check.names = FALSE))))
knitr::kable(res, align = "llr")
```

None of the probed species carries a distribution area in the GBIF copy, so
the Kosovo question cannot be answered from it.

The probe is small on purpose: five species, all of them known from Kosovo,
chosen across groups. If any carried distribution areas, the next step would
be to list them and look for Kosovo; since none does, a larger sample would
only confirm the same absence. The check is kept on the page so that it
re-runs, and flags the change, if the GBIF copy is ever republished with
distributions.

::: {.callout-warning}
## To be completed
Find the current Fauna Europaea service, read its area scheme, and check the
probe species above for a Kosovo entry. Record whether Kosovo is separate,
merged into Serbia, or labelled "Serbia and Montenegro" or "Yugoslavia".
:::

## What it adds for Natura 2000 in Kosovo

The animal counterpart of [Euro+Med](euro-med-plantbase.qmd): a reference for
whether an annex species is known from Kosovo at all, and so a check on GBIF
records that may be new national records or errors. Only useful if Kosovo is
an area of its own. It would serve the [national tier](../natura2000.qmd#q1).

## Caveats and sensitivity

- The checklist's own service was unreachable; everything here rests on the
  GBIF copy.
- The GBIF copy's licence is "unspecified".
- No sensitive content.

## Integration into the pipeline

Nothing to integrate until the distribution data are reachable.
`pipeline.R` is not changed.

## To be completed

```{r}
#| label: todo
#| output: asis
todo_list(row)
```

## Citation

Cite the GBIF copy as GBIF shows it:
[doi:10.15468/ymk1bx](https://doi.org/10.15468/ymk1bx).

::: {.callout-warning}
## To be completed
The citation Fauna Europaea itself requests could not be read.
:::
