---
title: "Forest Information System for Europe: Kosovo"
subtitle: "The EEA's forest portal keeps a country profile for Kosovo built from FAO and FOREST EUROPE reporting."
actions:
  - kind: check
    what: "Read the profile's figures (forest share 2020, growing stock, increment) in a browser; the automated fetch returned the indicator labels without their values"
    contact: "https://forest.eea.europa.eu/countries/cooperating-countries/kosovo"
  - kind: check
    what: "Compare the FAO/FOREST EUROPE forest area with the NFI 2012 figure of 481,000 ha"
    contact: "Kosovo NFI 2012 page in this section"
todo:
  doi: "Not applicable to a web profile; cite FAO FRA or FOREST EUROPE datasets by their own DOIs if used"
  licence: "Terms of use of the FISE portal; read on forest.eea.europa.eu"
  access_method: "Whether FISE offers the country data as download or API; check the portal's data pages"
  spatial_resolution: "National figures only; confirm whether FISE maps (e.g. forest type map) cover Kosovo"
  latest_version: "Update date of the profile; not shown in the fetched page"
  kosovo_codes_seen: "Which code FISE uses in its data downloads for Kosovo"
  effort: "Low if a download exists; estimate after checking"
  sensitivity: "None expected for national statistics"
---

```{r}
#| label: setup
source("../R/sources_helpers.R")
row <- source_row("fise-kosovo")
```

::: {.source-status}
`r status_line(row)`
:::

## At a glance

```{r}
#| label: glance
metadata_table(row)
```

## What it is

The Forest Information System for Europe (FISE), run by the European
Environment Agency, keeps country profiles for EEA members and cooperating
countries, and Kosovo has one. Its indicators are the share of land covered
by forest (2020), forest per capita (2020), growing stock and its density
(2020), and gross annual increment (2015). The profile names its sources as
the FAO Global Forest Resources Assessment, the FOREST EUROPE pan-European
dataset, Eurostat and the EEA. It is therefore a secondary compilation of
national reporting, not a dataset of its own.

## Access

```{r}
#| label: access
#| echo: true
#| code-fold: show
access_table(list(
  try_access("https://forest.eea.europa.eu/countries/cooperating-countries/kosovo")
))
```

The page is large (about 2 MB) and answered HTTP 200. Its figures are drawn
in the browser, and the automated read returned the indicator names and
sources without their values, so no figure is quoted here.

## Kosovo coverage check

::: {.callout-warning}
## To be completed
The profile exists for Kosovo, which settles coverage at national level. Its
values were not read. Open the page in a browser, record the four indicator
values with their years, and compare the forest share with the 45 per cent
the NFI 2012 abstract reports and with the 43 per cent broadleaved-woodland
share on the [ecosystem-type map](eea-ecosystem-types.qmd).
:::

## What it adds for Natura 2000 in Kosovo

Little directly for site identification: these are national totals. They are
useful as the internationally reported denominator for forest area, a check
on the forest share in the report's CORINE screen, and as a pointer to which
international datasets already hold Kosovo forest data. That serves the
[national tier](../natura2000.qmd#q1).

## Caveats and sensitivity

- A compilation of other datasets; cite the underlying FAO or FOREST EUROPE
  figures, not the portal.
- The values were not read in this check.
- No sensitive content.

## Integration into the pipeline

None needed. If the report quotes a national forest figure, it should take it
from the FAO or FOREST EUROPE source and state the year. `pipeline.R` is not
changed.

## To be completed

```{r}
#| label: todo
#| output: asis
todo_list(row)
```

## Citation

::: {.callout-warning}
## To be completed
No citation guidance was visible in the fetched page. Cite the underlying
dataset (FAO FRA or FOREST EUROPE) for any figure used.
:::
