---
title: "Global Forest Change (Hansen et al.)"
subtitle: "The University of Maryland's 30 m global map of tree-cover loss, version 1.13, now running from 2001 to 2025."
actions:
  - kind: check
    what: "Separate stand-replacing loss from harvest: set the loss pixels against the NFI and Forest Agency cutting records, and against EFFIS burnt areas"
    contact: "Kosovo Forest Agency; the EFFIS page in this section"
  - kind: check
    what: "Add treecover2000 (406 MB tile, read remotely) to express loss as a share of the forest present in 2000"
    contact: "https://storage.googleapis.com/earthenginepartners-hansen/GFC-2025-v1.13/download.html"
todo: {}
---

```{r}
#| label: setup
source("../R/sources_helpers.R")
suppressPackageStartupMessages({
  library(sf)
  library(terra)
  library(leaflet)
})
row <- source_row("forest-loss-hansen")
cache <- file.path(cache_dir("forest-loss-hansen"), "gfc_v1_13_lossyear_kosovo.tif")
tile <- "https://storage.googleapis.com/earthenginepartners-hansen/GFC-2025-v1.13/Hansen_GFC-2025-v1.13_lossyear_50N_020E.tif"
```

::: {.source-status}
`r status_line(row)`
:::

## At a glance

```{r}
#| label: glance
metadata_table(row)
```

## What it is

Hansen and colleagues at the University of Maryland map, for every 30 m pixel
of the world's land, the tree cover in 2000 and the year in which any
stand-replacing loss of it was detected. The data behind their 2013 paper in
*Science* have been extended every year since. Version 1.13, covering
2000–2025, is the latest on the download page and replaces version 1.12
(2000–2024); both are licensed CC BY 4.0. The desk review named the product
without a version; v1.13 is the one used here.

## Access

Open, as 10 × 10 degree GeoTIFF tiles. All of Kosovo falls in the tile
`50N_020E`; its loss-year layer is 37 MB, and the Kosovo window is read
remotely and cached.

```{r}
#| label: access
#| echo: true
#| code-fold: show
access_table(list(
  try_access("https://storage.googleapis.com/earthenginepartners-hansen/GFC-2025-v1.13/download.html"),
  try_access(tile)
))
```

```{r}
#| label: build
#| echo: true
#| code-fold: true
#| code-summary: "Show the remote crop"
build_error <- NULL
if (!file.exists(cache)) {
  build_error <- tryCatch({
    Sys.setenv(GDAL_DISABLE_READDIR_ON_OPEN = "EMPTY_DIR")
    ko <- vect(kosovo_outline())
    k  <- mask(crop(rast(paste0("/vsicurl/", tile)), ko, snap = "out"), ko)
    writeRaster(k, cache, overwrite = TRUE, datatype = "INT1U",
                gdal = "COMPRESS=DEFLATE")
    NULL
  }, error = function(e) conditionMessage(e))
}
have_cache <- file.exists(cache)
```

```{r}
#| label: build-failed
#| eval: !expr "!have_cache"
#| output: asis
cat("::: {.callout-warning}\n## To be completed\nThe remote crop failed at render time (",
    build_error %||% "no message", "). Re-render when the tile server answers.\n:::\n", sep = "")
```

## Kosovo coverage check

```{r}
#| label: check
#| eval: !expr have_cache
k <- rast(cache)
area_km2 <- cellSize(k, unit = "km")
z <- zonal(area_km2, k, fun = "sum")
names(z) <- c("code", "km2")
n_na <- global(is.na(k) & !is.na(rasterize(vect(kosovo_outline()), k)), "sum")[[1]]
loss <- z[z$code > 0, ]
loss$year <- 2000 + loss$code
knitr::kable(data.frame(Year = loss$year, `Tree-cover loss (km²)` = round(loss$km2, 1),
                        check.names = FALSE), align = "rr", row.names = FALSE)
```

```{r}
#| label: check-text
#| eval: !expr have_cache
#| output: asis
top <- loss[order(-loss$km2), ][1:3, ]
cat(sprintf(paste(
  "Between 2001 and %d, %s km² of tree cover was lost inside Kosovo. The",
  "largest years were %s. %s pixels inside the outline are no-data. The",
  "raster has no country attribute, so no country code applies."),
  max(loss$year), format(round(sum(loss$km2)), big.mark = ","),
  paste(sprintf("%d (%.1f km²)", top$year, top$km2), collapse = ", "),
  fmt_int(n_na)))
```

```{r}
#| label: map
#| eval: !expr have_cache
source("../R/functions.R", local = TRUE)
lost <- k > 0
share <- aggregate(lost, 10, fun = "mean", na.rm = TRUE)   # about 300 m
share[share == 0] <- NA
pal <- colorNumeric(map_palette$density, c(0, 1), na.color = NA)
leaflet() |>
  add_basemaps() |>
  addRasterImage(share, colors = pal, opacity = 0.8) |>
  add_boundary_outline(kosovo_outline()) |>
  addLegend(pal = pal, values = c(0, 1), title = "Share of pixels<br>lost 2001–2025",
            labFormat = labelFormat(transform = function(x) 100 * x, suffix = " %")) |>
  addLayersControl(baseGroups = basemap_groups,
                   options = layersControlOptions(collapsed = TRUE))
```

The map shows the share of 30 m pixels lost in each cell of about 300 m.

## What it adds for Natura 2000 in Kosovo

Forest habitat types will make up much of a Kosovo network, and the European
Commission's 2025 report says that 40 per cent of public forests are subject
to uncontrolled or illegal harvesting. Loss mapped year by year shows where
that pressure lands, whether inside existing protected areas, and whether it
is increasing. That is evidence for the pressures section of an SDF and for
the trend in a forest habitat's area. At 30 m it serves the
[national tier](../natura2000.qmd#q1), and the site tier as a first screen.

## Caveats and sensitivity

- "Loss" is stand-replacing removal of tree cover; it includes legal
  harvesting, fire and windthrow, and cannot tell them apart.
- Read the usage notes on the download page before comparing early years
  with late ones; this check reports the years as the layer gives them.
- No species locations; nothing sensitive.

## Integration into the pipeline

Low effort. A proposed `kosovo_forest_loss(boundary)` in `R/functions.R`
would do the crop above; the report could add loss per protected area and per
municipality beside its habitat screen. `pipeline.R` is not changed.

## To be completed

```{r}
#| label: todo
#| output: asis
todo_list(row)
```

## Citation

As the download page asks: Hansen, M. C., P. V. Potapov, R. Moore, M. Hancher,
S. A. Turubanova, A. Tyukavina, D. Thau, S. V. Stehman, S. J. Goetz, T. R.
Loveland, A. Kommareddy, A. Egorov, L. Chini, C. O. Justice, and J. R. G.
Townshend. 2013. High-Resolution Global Maps of 21st-Century Forest Cover
Change. *Science* 342 (15 November): 850–53. Data available on-line from:
https://glad.earthengine.app/view/global-forest-change.
