---
title: "HydroSHEDS"
subtitle: "A globally consistent river network and catchments; the Europe file of HydroRIVERS is downloaded once and cut to Kosovo."
actions:
  - kind: check
    what: "Add HydroBASINS catchments (level 8 to 12) for Kosovo, to give the boundary workflow watershed lines"
    contact: "HydroSHEDS products, https://www.hydrosheds.org/"
todo:
  doi: "HydroRIVERS itself has no DOI seen; the citation paper does (doi:10.1002/hyp.9740)"
  temporal_coverage: "Not applicable: a static network derived from elevation data"
---

```{r}
#| label: setup
source("../R/sources_helpers.R")
suppressPackageStartupMessages({
  library(sf)
  library(leaflet)
})
row <- source_row("hydrosheds")
dir <- cache_dir("hydrosheds")
zip <- file.path(dir, "HydroRIVERS_v10_eu_shp.zip")     # 68 MB: git-ignored
cache <- file.path(dir, "hydrorivers_kosovo.gpkg")
url <- "https://data.hydrosheds.org/file/HydroRIVERS/HydroRIVERS_v10_eu_shp.zip"
```

::: {.source-status}
`r status_line(row)`
:::

## At a glance

```{r}
#| label: glance
metadata_table(row)
```

## What it is

HydroSHEDS is a suite of global hydrographic data derived from elevation
data: flow directions, catchments (HydroBASINS) and rivers (HydroRIVERS).
HydroRIVERS represents all rivers of the world as a vector network extracted
from the HydroSHEDS grids at 15 arc-seconds, about 500 m, with attributes such
as flow order and discharge. It is "freely available for scientific,
educational and commercial use" under the HydroSHEDS licence, with
attribution. Where [EU-Hydro](eu-hydro.qmd) is a photo-interpreted European
network at 1:50,000, HydroRIVERS is coarser and modelled, but identical in
method everywhere, including across all of Kosovo's borders.

## Access

```{r}
#| label: access
#| echo: true
#| code-fold: show
access_table(list(try_access("https://www.hydrosheds.org/products/hydrorivers"),
                  try_access(url)))
```

The Europe file is a 68 MB zipped shapefile. A shapefile has no spatial
index GDAL can use over HTTP, so it is downloaded once, outside version
control, and only the Kosovo extract is kept. The contrast with the
GeoPackages read remotely on other pages ([Article 17](art17-2013-2018.qmd),
the protected areas in the pipeline) is the teaching point: a format with an
R-tree index can be queried in place, one without must be fetched whole. The
archive's size, checksum and date are recorded below, since it is not
committed.

```{r}
#| label: build
#| echo: true
#| code-fold: true
#| code-summary: "Show the download and cut"
build_error <- NULL
if (!file.exists(cache)) {
  build_error <- tryCatch({
    if (!file.exists(zip)) {
      got <- try_access(url, zip, timeout = 900)
      if (!got$success) stop(got$error %||% paste("HTTP", got$http_code))
    }
    shp <- grep("\\.shp$", unzip(zip, list = TRUE)$Name, value = TRUE)
    ko  <- st_union(kosovo_outline())
    box <- st_as_text(st_as_sfc(st_bbox(ko)))
    r   <- st_read(paste0("/vsizip/", zip, "/", shp), wkt_filter = box, quiet = TRUE)
    r   <- suppressWarnings(st_intersection(st_make_valid(r), ko))
    st_write(r, cache, delete_dsn = TRUE, quiet = TRUE)
    NULL
  }, error = function(e) conditionMessage(e))
}
have_cache <- file.exists(cache)
```

```{r}
#| label: build-failed
#| eval: !expr "!have_cache"
#| output: asis
cat("::: {.callout-warning}\n## To be completed\nThe download or cut failed at render time (",
    build_error %||% "no message", "). Re-render when data.hydrosheds.org answers.\n:::\n", sep = "")
```

```{r}
#| label: fingerprint
#| eval: !expr file.exists(zip)
#| output: asis
fp <- file_fingerprint(zip)
cat(sprintf("Downloaded archive: `%s`, %s, SHA-256 `%s`, %s (not committed).\n",
            fp$file, fp$size, fp$sha256, fp$downloaded))
```

## Kosovo coverage check

```{r}
#| label: check
#| eval: !expr have_cache
r <- st_read(cache, quiet = TRUE)
ord <- intersect(c("ORD_STRA", "ORD_CLAS", "ORD_FLOW"), names(r))[1]
len <- tapply(as.numeric(st_length(st_transform(r, 3035))) / 1000, r[[ord]], sum)
knitr::kable(data.frame(Order = names(len), `Length inside Kosovo (km)` = fmt_int(len),
                        check.names = FALSE), align = "rr", row.names = FALSE,
             caption = sprintf("HydroRIVERS segments inside Kosovo by %s", ord))
```

```{r}
#| label: check-text
#| eval: !expr have_cache
#| output: asis
cat(sprintf(paste(
  "%s river segments, %s km in all, lie inside Kosovo. The network is",
  "continuous across the border by construction. It carries no country",
  "attribute, so no country code applies."),
  fmt_int(nrow(r)), fmt_int(sum(len))))
```

```{r}
#| label: map
#| eval: !expr have_cache
source("../R/functions.R", local = TRUE)
leaflet(r) |>
  add_basemaps() |>
  addPolylines(color = "#0072B2", weight = ~pmax(0.5, as.numeric(r[[ord]]) / 2),
               opacity = 0.8) |>
  add_boundary_outline(kosovo_outline()) |>
  addLayersControl(baseGroups = basemap_groups,
                   options = layersControlOptions(collapsed = TRUE))
```

## What it adds for Natura 2000 in Kosovo

A river network that does not stop at the border, which matters for
transboundary rivers such as the White Drin and the Ibër, and a route to
HydroBASINS catchments: the watershed lines the boundary workflow of
[question 2](../natura2000.qmd#q2) says a site boundary may follow. Coarser
than EU-Hydro, it suits national screening ([tier](../natura2000.qmd#q1)).

## Caveats and sensitivity

- Modelled from a 500 m grid: river positions can be hundreds of metres off
  in flat terrain.
- No sensitive content.

## Integration into the pipeline

Low effort; one download and a cut, as above. A proposed
`kosovo_catchments(boundary, level = 8)` would be the more useful addition,
for boundary work. `pipeline.R` is not changed.

## To be completed

```{r}
#| label: todo
#| output: asis
todo_list(row)
```

## Citation

As HydroSHEDS requests: Lehner, B. & Grill, G. (2013). Global river
hydrography and network routing: baseline data and new approaches to study the
world's large river systems. *Hydrological Processes* 27(15): 2171–2186.
[doi:10.1002/hyp.9740](https://doi.org/10.1002/hyp.9740).
