---
title: "Important Bird and Biodiversity Areas and Key Biodiversity Areas"
subtitle: "BirdLife's inventory lists four IBAs that concern Kosovo, held under Serbia; their boundaries are released only on request."
actions:
  - kind: request
    what: "Digital boundaries of IBAs RS022 Kopaonik, RS023 Prokletije, RS024 Sara mountain and RS062 Mokra Gora, and of any KBA touching Kosovo, for non-commercial use"
    contact: "World Database of Key Biodiversity Areas, request form at https://www.keybiodiversityareas.org/request-gis-data"
  - kind: check
    what: "Once boundaries arrive, measure how much of each site lies inside Kosovo; the report now uses centroids only"
    contact: "R/fetch_iba.R and natura2000.qmd, question 6"
todo:
  doi: "No DOI for the IBA inventory; cite the factsheets as BirdLife gives them"
  licence: "Terms come with the KBA data request (non-commercial use free, per the report's Natura 2000 page); read them on the request form"
  format: "Boundaries are supplied as GIS files on request; the format is set by the request"
  access_method: "Attributes via BirdLife's DataZone API; boundaries by request"
  spatial_resolution: "Scale of the IBA boundaries; stated with the delivered data"
  temporal_coverage: "Year of last assessment per site (2019 for all four in data/kosovo_ibas.csv)"
  latest_version: "Date of the World Database of KBAs release used"
  sensitivity: "IBA factsheets name qualifying species with population sizes; boundaries are not sensitive"
---

```{r}
#| label: setup
source("../R/sources_helpers.R")
suppressPackageStartupMessages({
  library(sf)
  library(leaflet)
})
row <- source_row("iba-kba")
api <- "https://bli-prod-fd-dz-eu-bgf5eqfcf2bmgtdn.a02.azurefd.net"
cache <- file.path(cache_dir("iba-kba"), "birdlife_country_check.csv")
```

::: {.source-status}
`r status_line(row)`
:::

## At a glance

```{r}
#| label: glance
metadata_table(row)
```

## What it is

Important Bird and Biodiversity Areas (IBAs) are sites that BirdLife
International and its partners identify by standard criteria for their
importance to bird populations. Key Biodiversity Areas (KBAs) extend the idea
to all biodiversity, and IBAs are part of the World Database of KBAs. The
report already uses the public IBA factsheets: `R/fetch_iba.R` builds
`data/kosovo_ibas.csv` and `data/kosovo_iba_species.csv`, and the
[Natura 2000 page](../natura2000.qmd#q6) discusses the four sites. The
boundaries, which a site proposal needs, are released only through a request
form with terms of use.

## Access

```{r}
#| label: access
#| echo: true
#| code-fold: show
access_table(list(
  try_access("https://datazone.birdlife.org/site/factsheet/3126-sara-mountain"),
  try_access(paste0(api, "/country/271/sites?pageSize=5")),
  try_access("https://www.keybiodiversityareas.org/request-gis-data")
))
```

The factsheets and BirdLife's DataZone API are open; boundaries were not
requested, and the pipeline deliberately does not take them from the map
server behind DataZone, since that would skip the request and its terms.

## Kosovo coverage check

The check, re-run on each render, asks BirdLife's API for a Kosovo country
record and lists the sites concerning Kosovo from the Serbia inventory.

```{r}
#| label: build
#| echo: true
#| code-fold: true
#| code-summary: "Show the API check"
if (!file.exists(cache)) {
  res <- tryCatch({
    k <- curl::curl_fetch_memory(paste0(api, "/country/slug?country=kosovo"))
    s <- jsonlite::fromJSON(paste0(api, "/site/3126"))
    data.frame(check = c("Kosovo country record (country/slug?country=kosovo)",
                         "Country that holds Sara mountain (site 3126)",
                         "ISO2 of that country"),
               result = c(sprintf("HTTP %d, %d bytes", k$status_code, length(k$content)),
                          s$country$birdLifeName, s$country$iso2))
  }, error = function(e) NULL)
  if (!is.null(res)) write.csv(res, cache, row.names = FALSE)
}
if (file.exists(cache)) knitr::kable(read.csv(cache), col.names = c("Check", "Result"))
```

```{r}
#| label: check
ibas <- read.csv(file.path(sources_root(), "data", "kosovo_ibas.csv"), encoding = "UTF-8")
knitr::kable(data.frame(Code = ibas$code, Site = ibas$name, `Where it lies` = ibas$relation,
                        `Area (km²)` = ibas$area_km2, `IBA criteria` = ibas$ibaCriteria,
                        `KBA` = ibas$kbaClassification, check.names = FALSE),
             align = "lllrll", row.names = FALSE,
             caption = sprintf("From data/kosovo_ibas.csv, built by R/fetch_iba.R on %s",
                               ibas$accessed[1]))
```

BirdLife holds no country record for Kosovo: the request returns an empty
answer, and the four sites sit in the inventory compiled for Serbia, with
codes RS022, RS023, RS024 and RS062. That is where the inventory was compiled,
not a statement about status, but it means that any count of IBAs by country
files these sites under Serbia. Whether a KBA other than these IBAs touches
Kosovo was not checked.

```{r}
#| label: map
source("../R/functions.R", local = TRUE)
leaflet(ibas) |>
  add_basemaps() |>
  addCircleMarkers(~lon, ~lat, radius = ~sqrt(area_km2) / 3, color = "#175CA1",
                   weight = 1.5, fillOpacity = 0.25,
                   label = ~sprintf("%s (%s): %s km², %s", name, code, area_km2, relation)) |>
  add_boundary_outline(kosovo_outline()) |>
  addLayersControl(baseGroups = basemap_groups,
                   options = layersControlOptions(collapsed = TRUE))
```

Circles mark the published centroids, sized by site area; they are not
boundaries.

## What it adds for Natura 2000 in Kosovo

The IBA inventory is the starting reference for SPAs; the Court of Justice
has treated it as the best available scientific evidence for judging whether
enough SPAs are classified (see [question 6](../natura2000.qmd#q6)). With
boundaries, the report could say how much of Šar and Prokletije's IBAs lie in
Kosovo, and draw candidate SPAs against them. Boundaries serve the
[site tier](../natura2000.qmd#q1) as a first outline; the attributes serve
national screening.

## Caveats and sensitivity

- Centroids say where a site is, not how much of it lies on either side of a
  border.
- The qualifying populations date from 1998 to 2017 (see the report's table),
  several of them old.
- Boundaries are released under terms that come with the request; they
  must not be scraped from the map server.

## Integration into the pipeline

Low effort. `R/fetch_iba.R` already builds the attribute tables. A proposed
`read_iba_boundaries(path)` would add the requested polygons once they
arrive, and compute each site's share inside Kosovo. `pipeline.R` is not
changed.

## To be completed

```{r}
#| label: todo
#| output: asis
todo_list(row)
```

## Citation

As the report cites them: BirdLife International (2026). Important Bird Area
factsheets: *Sara mountain*, *Prokletije*, *Kopaonik*, *Mokra Gora*. DataZone,
datazone.birdlife.org.
