---
title: "Independent Commission for Mines and Minerals: licences and GIS"
subtitle: "Kosovo's mining regulator publishes licence applications and a GIS portal; the GIS sits behind a login."
actions:
  - kind: request
    what: "Mining and exploration licence areas as GIS data (polygons with licence type, holder, status and dates), for use in a Natura 2000 overlay"
    contact: "Independent Commission for Mines and Minerals (ICMM/KPMM), https://kosovo-mining.org/?lang=en"
  - kind: check
    what: "Terms of use of the GIS portal and whether an account is available to public bodies or researchers"
    contact: "ICMM, through its website"
todo:
  doi: "No DOI; a regulator's register"
  licence: "Terms of the GIS portal and data shop; not stated on the pages that could be read"
  format: "Format of licence data behind the login; not seen"
  access_method: "Login-only GIS portal; check whether a WMS/WFS is offered to account holders"
  spatial_resolution: "Licence polygons; scale not known"
  temporal_coverage: "Licence dates; read in the register"
  latest_version: "Update date of the licence register"
  kosovo_codes_seen: "Not applicable (national register)"
  effort: "Low once licence polygons are obtained"
  sensitivity: "Commercial information in licences; check what may be published"
---

```{r}
#| label: setup
source("../R/sources_helpers.R")
row <- source_row("icmm-mining")
```

::: {.source-status}
`r status_line(row)`
:::

## At a glance

```{r}
#| label: glance
metadata_table(row)
```

## What it is

The Independent Commission for Mines and Minerals (ICMM, KPMM in Albanian) is
Kosovo's mining regulator. Its website offers application forms for
exploration and mining licences, a "data portal", a GIS link and a "data
shop" selling printed maps of Kosovo at 1:50,000, 1:100,000 and 1:200,000.
Its home page also summarises reserves of the main mineral resources:
lignite, lead and zinc, nickel, bauxite and magnesite.

## Access

```{r}
#| label: access
#| echo: true
#| code-fold: show
access_table(list(
  try_access("https://kosovo-mining.org/?lang=en"),
  try_access("https://www.kosovo-mining.org/gdk/gis?lan=en")
))
```

The home page answers, although the automated fetch tool could not verify
the site's certificate chain. The GIS link redirects to a login page
(`/gdk/Login`). Nothing behind the login was attempted, and no account was
requested. A login is a decision by the data holder, and the right response
is a request, listed below, not a way round it.

## Kosovo coverage check

::: {.callout-warning}
## To be completed
The register is national, so coverage is not in doubt, but no licence data
could be read. With licence polygons: clip to Kosovo, count licences by type
and status inside each NatDA site and within 1 km of it, and list the mineral
types involved.
:::

## What it adds for Natura 2000 in Kosovo

Mining is a pressure that a site proposal must face early, because a licence
granted inside a candidate site constrains its designation and management.
Knowing where exploration and mining licences lie, and how they overlap the
existing protected areas and the habitat screen, is part of the pressure
analysis for an SDF (section 4.3, threats and pressures) and for any
appropriate assessment. Licence polygons would serve the
[site tier](../natura2000.qmd#q1).

## Caveats and sensitivity

- A login is required; the terms of access are unknown.
- Licence areas are not mine footprints; they mark where activity is
  permitted.

## Integration into the pipeline

Low effort with data: a proposed `read_mining_licences(path)` would clip the
polygons and report overlaps with `protected_areas`. `pipeline.R` is not
changed.

## To be completed

```{r}
#| label: todo
#| output: asis
todo_list(row)
```

## Citation

::: {.callout-warning}
## To be completed
No citation guidance was seen. Cite the ICMM licence register with its access
date once the data are obtained.
:::
