---
title: "Important Plant Areas"
subtitle: "Plantlife's framework for the richest sites for wild plants; its global database was offline, and no Kosovo IPA was confirmed."
actions:
  - kind: check
    what: "Whether any Important Plant Area has been identified in Kosovo, and by whom"
    contact: "Global IPA Manager at Plantlife (the offline notice names the contact); University of Prishtina botanists"
  - kind: check
    what: "Whether Serbia's 62 IPAs include sites in Kosovo; Plantlife's Serbia page mentions pine in 'the mountains of Kosovo and Metochia province'"
    contact: "Plantlife (2023), Conserving Plant Areas: Investing in the Green Gold of South-East Europe, pages 63–68"
  - kind: check
    what: "Confirm the IPAs on the Albanian and North Macedonian side of Korab-Koritnik and Šar, reported by the desk review"
    contact: "Plantlife's South-East Europe report; the Global IPA Database when it is back online"
todo:
  doi: "No DOI; check the 2023 report"
  licence: "Terms of the Global IPA Database; unreadable while it is offline"
  access: "The database was offline on 29 September 2026"
  format: "Site records and boundaries in the database; not seen"
  access_method: "Unknown while the database is offline"
  spatial_resolution: "IPA boundaries or points; not seen"
  temporal_coverage: "Year of identification per IPA"
  latest_version: "Date of the database or the 2023 South-East Europe report"
  kosovo_coverage: "No IPA in Kosovo found; not ruled out"
  kosovo_codes_seen: "Whether Serbian IPAs in Kosovo are filed under Serbia"
  tier: "National or site, depending on the boundary precision"
  n2k_question: "Where botanists judged plant importance highest, as a check on candidate SCIs for Annex I habitats and Annex II plants"
  effort: "Unknown until data are available"
  sensitivity: "IPA criteria name threatened plants; site boundaries are not sensitive"
---

```{r}
#| label: setup
source("../R/sources_helpers.R")
row <- source_row("important-plant-areas")
```

::: {.source-status}
`r status_line(row)`
:::

## At a glance

```{r}
#| label: glance
metadata_table(row)
```

## What it is

Important Plant Areas (IPAs) are sites identified, under a programme run by
Plantlife, as the most important places for wild plants and their habitats.
Plantlife's Global IPA Database, which holds them, was offline when checked:
its home page states that the database is "currently offline and unavailable
until further notice".

Plantlife's own page on Serbia reports 62 IPAs, drawing on a 2023 report on
South-East Europe, and mentions Macedonian pine in "the mountains of Kosovo
and Metochia province". That phrase suggests the Serbian inventory may reach
into Kosovo, as the Serbian Emerald candidate sites do (see
[Emerald Network](emerald-network.qmd)); it does not confirm any IPA there.
The desk review reported IPAs on the Albanian side of Korab-Koritnik; that was
not verified.

## Access

```{r}
#| label: access
#| echo: true
#| code-fold: show
access_table(list(
  try_access("https://www.plantlifeipa.org/home"),
  try_access("https://www.plantlife.org.uk/our-work/important-plant-areas-of-serbia/")
))
```

## Kosovo coverage check

::: {.callout-warning}
## To be completed
With the database offline, nothing could be counted. When it is back, or once
the 2023 report is obtained: list IPAs within 20 km of Kosovo, check whether
any Serbian IPA lies inside Kosovo and under which country it is filed, and
set them against the report's habitat screen.
:::

## What it adds for Natura 2000 in Kosovo

The Habitats Directive's plant side, Annex I habitats and Annex II plants, is
where Kosovo's data are thinnest. IPAs are expert judgements of where plant
importance concentrates. A Kosovo IPA list, if one exists, would be an
obvious shortlist to test against candidate SCIs.

The Serbian page raises the same question as the
[Emerald Network](emerald-network.qmd) and [IBA](iba-kba.qmd) pages: whether
a regional inventory compiled with Serbia counts Kosovo's mountains as
Serbian sites. If it does, those IPAs are Kosovo evidence filed under another
country, and they should be found and read, not dismissed as foreign.

## Caveats and sensitivity

- Lead, not verified: no Kosovo IPA was confirmed.
- The Serbian page's wording is evidence of scope, not of any particular site.

## Integration into the pipeline

None until data exist. `pipeline.R` is not changed.

## To be completed

```{r}
#| label: todo
#| output: asis
todo_list(row)
```

## Citation

::: {.callout-warning}
## To be completed
Cite the Global IPA Database or the 2023 South-East Europe report as
Plantlife requests, once either is read.
:::
