---
title: "Data sources"
subtitle: "Candidate datasets for identifying Natura 2000 sites in Kosovo, each checked against its custodian and, where the data could be reached, against Kosovo itself."
toc-depth: 2
code-tools: false
---

```{r}
#| label: setup
source("../R/sources_helpers.R")
reg <- read_registry()
reg$completeness <- vapply(seq_len(nrow(reg)), function(i)
  completeness(as.list(reg[i, ])), numeric(1))
dash <- function(x) ifelse(nzchar(x), md_cell(x), "&mdash;")
actions <- do.call(rbind, lapply(reg$slug, page_actions))
```

The [report](../report.qmd) is built on one GBIF download, cleaned against an
OpenStreetMap outline, and set against the EEA's register of Kosovo's
protected areas, the EU annex lists, the global Red List and the Union list
of invasive species. This section looks beyond them. Each page below takes
one candidate dataset from a desk review of what else could inform Natura 2000
site identification in Kosovo[\*]{.designation-mark}, checks what its
custodian actually publishes, and, where the data could be reached, runs a
check against Kosovo's territory. The pages are written so that each can be
read, re-run and completed on its own.

The rule throughout is that a value is shown only if it was verified. Every
filled field in the registry has a row in `sources/registry_evidence.csv`
naming the page it was checked against and the date; a field without one is
blanked automatically and appears as "— to be completed". A gap is not an
error; a guess would be. The precision tiers named below (national screening,
site-level delineation, Standard Data Form) are those defined on the
[Natura 2000 page](../natura2000.qmd#q1).

::: {.designation-note}
\* This designation is without prejudice to positions on status, and is in line
with UNSCR 1244/1999 and the ICJ Opinion on the Kosovo declaration of
independence.
:::

## Status

| Status | Meaning |
|---|---|
| `r status_badge("Complete")` | Every registry field verified, and the Kosovo check run. |
| `r status_badge("Partial")` | Some fields verified, or the Kosovo check only partly run. |
| `r status_badge("Documented only")` | The dataset is described in a report or paper, but not published as data. |
| `r status_badge("Awaiting request")` | The data exist but are released on request or after registration only. |
| `r status_badge("Lead, not verified")` | Named in the desk review; no primary source confirmed yet. |
| `r status_badge("Not applicable to Kosovo")` | The product exists but leaves Kosovo out by design; the page is kept because the reason is itself evidence. |

```{r}
#| label: status-counts
#| output: asis
st <- table(factor(reg$status, levels = status_levels))
cat(sprintf("The %d datasets stand as follows: %s. Across all of them, %d per cent of the verifiable registry fields are filled.\n",
            nrow(reg),
            paste(sprintf("%s: %d", names(st), as.integer(st)), collapse = "; "),
            round(mean(reg$completeness))))
```

## What the checks found

Four findings cut across the pages and are worth knowing before reading any
one of them.

- **Kosovo hides under other codes.** The same territory appears as XK
  (GBIF), XKO (GADM and `geodata`), KS (EFFIS) and Ko (Euro+Med), and inside
  RS where European data were compiled with Serbia: Serbia's candidate
  [Emerald sites](emerald-network.qmd) cover about a fifth of Kosovo,
  [BirdLife](iba-kba.qmd) holds Kosovo's IBAs in its Serbia inventory,
  [GISCO](r-access-toolkit.qmd) has no Kosovo feature at all, and
  [EFFIS](effis-burnt-areas.qmd) files some Kosovo fires under RS. Any query
  by a single country code undercounts.
- **The EU's own reporting stops about 50 km away.** Kosovo's four neighbours
  are outside the EU, so the nearest [Article 17](art17-2013-2018.qmd) and
  [Article 12](art12-birds.qmd) cells are in Bulgaria.
- **European habitat models do cover Kosovo.** The
  [EUNIS level-3 maps](eunis-habitat-maps.qmd) and the
  [ecosystem-type map](eea-ecosystem-types.qmd) have no gaps inside the
  outline, which makes them the closest thing to the Annex I map Kosovo lacks.
- **Much of what Kosovo knows is in PDFs.** The red books, the protected-area
  table and the park plans are published, but not as data; and the national
  [protected-area table](kepa-protected-areas-2020.qmd) and the EEA register
  use different code prefixes, so they cannot be joined without translation.

## The registry

Completeness is the share of the sixteen verifiable fields that are filled.

```{r}
#| label: registry
#| output: asis
for (cat_id in names(category_labels)) {
  sub <- reg[reg$category == cat_id, ]
  if (!nrow(sub)) next
  cat(sprintf("\n### %s. %s\n\n", cat_id, category_labels[[cat_id]]))
  tab <- data.frame(
    Dataset = sprintf("[%s](%s.qmd)", md_cell(sub$title), sub$slug),
    Custodian = dash(sub$custodian),
    Status = vapply(sub$status, status_badge, ""),
    `Kosovo coverage` = dash(sub$kosovo_coverage),
    Tier = dash(sub$tier),
    Effort = dash(sub$effort),
    Complete = sprintf("%d %%", sub$completeness),
    check.names = FALSE)
  cat("::: {.sources-table}\n\n")
  print(knitr::kable(tab, format = "pipe", escape = FALSE, align = "lllllll",
                     row.names = FALSE))
  cat("\n:::\n")
}
```

## Data gaps {#data-gaps}

Some of what a Natura 2000 proposal needs does not exist in any of the
sources above. Each gap is stated with why it matters for the Standard Data
Form (SDF) and who could close it.

### 1. No Annex I habitat map or inventory

**Missing:** a map of Annex I habitat types for Kosovo. The
[EUNIS model](eunis-habitat-maps.qmd) and the report's CORINE screen say
where to look, not what is there. **For the SDF:** section 3.1 needs each
habitat's cover, representativity and conservation, which only mapped
polygons provide ([question 3](../natura2000.qmd#q3)). **Who:** the Kosovo
Institute for Nature Protection with University of Prishtina botanists, by
field mapping at 1:10,000, starting in the pilot sites.

### 2. No official or draft Natura 2000 or Emerald site list with boundaries

**Missing:** any candidate site list for Kosovo. The 2008 identification
([Mustafa](mustafa-2009-potential-n2k.qmd)) is known only from citations, and
Kosovo is not a Bern Convention Party, so it has no Emerald list of its own.
**For the SDF:** there is nothing yet to fill one for. **Who:** the Ministry
and KINP, which the NBSAP makes responsible for the ecological network.

### 3. No confirmed IBA or KBA boundaries

**Missing:** the digital boundaries of the four [IBAs](iba-kba.qmd) that
concern Kosovo; only centroids are public. **For the SDF:** SPA proposals
start from the IBA inventory, and how much of each IBA lies in Kosovo is
unknown. **Who:** a request to the World Database of KBAs, whose terms come with the
request form.

### 4. No published national species monitoring data

**Missing:** surveillance data with method and effort. The red books of
[fauna](red-book-fauna.qmd) and [flora](red-book-flora.qmd) assess species
but do not release their localities as data. **For the SDF:** section 3.2's
population and data-quality fields need surveyed records, not casual ones.
**Who:** KEPA and KINP, with the red-book authors; a data-use agreement could
release localities at a generalised resolution.

### 5. No population-size data for SDF section 3.2

**Missing:** population estimates with units (pairs, individuals, localities)
for Annex II species and Annex I birds. The IBA factsheets give a few, several
from 2008–2013 or earlier. **For the SDF:** section 3.2 requires minimum and
maximum population with a unit and a data-quality code. **Who:** standardised
counts in the pilot sites ([question 6](../natura2000.qmd#q6)); the
[eBird Basic Dataset](ebird-observation-org.qmd) for effort-based bird data.

### 6. No wetland, peatland, cave or grassland inventories

**Missing:** national inventories of the habitat groups most likely to hold
priority Annex I types: wetlands and mires, caves, dry and mountain grasslands.
**For the SDF:** these are small or scattered habitats that land-cover data
miss at a 25 ha minimum unit. **Who:** KINP with specialists; the natural
monuments in the [national table](kepa-protected-areas-2020.qmd), many of them
springs and caves, are a starting list.

### 7. No Article 17-style range assessments, and no delineation of biogeographical regions within Kosovo

**Missing:** national range and distribution maps on the 10 km grid, as EU
countries report under [Article 17](art17-2013-2018.qmd), and a Kosovo
delineation of the [biogeographical regions](eea-biogeographical-regions.qmd)
finer than the European 1:1,000,000 line. **For the SDF:** a site's
contribution is judged against the national population and range in its
region. **Who:** KINP, as the NBSAP foresees; the European layer is the
starting point.

### 8. Scattered evidence for Annex II invertebrates and freshwater fish

**Missing:** systematic data for the Annex II groups least covered by casual
recording. The [data papers](kosovo-data-papers.qmd) add caddisflies and
plants but no Annex II list; the [taxon atlases](european-taxon-atlases.qmd)
are unverified. **For the SDF:** without them, sites cannot be proposed for
these species at all. **Who:** university specialists, with targeted surveys
in the rivers the [hydropower inventory](balkan-hydropower.qmd) shows under
pressure.

## Data requests

Every request listed on the dataset pages, in one place. Each is also on its
own page, under "To be completed".

```{r}
#| label: requests
#| output: asis
req <- actions[actions$kind == "request", ]
req$title <- reg$title[match(req$slug, reg$slug)]
knitr::kable(data.frame(Dataset = sprintf("[%s](%s.qmd)", md_cell(req$title), req$slug),
                        Request = md_cell(req$what), Route = md_cell(req$contact),
                        check.names = FALSE),
             format = "pipe", escape = FALSE, row.names = FALSE) |> print()
cat(sprintf("\n%d requests on %d pages; the pages list a further %d checks that need no one's permission.\n",
            nrow(req), length(unique(req$slug)), sum(actions$kind == "check")))
```

## How this section is maintained

- `sources/registry.csv` holds one row per dataset;
  `sources/registry_evidence.csv` one row per verified value. Edit both
  together.
- `R/sources_helpers.R` reads them, blanks unverified values, and draws the
  tables, badges and checklists.
- Pages cache what they download under `data/sources/<slug>/`. Small open
  extracts are committed; PDFs and anything over 50 MB are not.
- This folder renders with `freeze: auto`. After editing the registry or the
  helpers, run `quarto render sources/` so every page picks up the change.

```{r}
#| label: dropped
#| output: asis
u <- attr(reg, "unverified")
if (length(u)) cat(sprintf("\n::: {.callout-warning}\n## Values blanked for lack of evidence\n%s\n:::\n",
                           paste(u, collapse = "; ")))
```
