---
title: "Kosovo Cadastral Agency State Geoportal"
subtitle: "The national geoportal of the Kosovo Cadastral Agency; open map services could not be found at the addresses tried."
actions:
  - kind: check
    what: "Find the geoportal's documented WMS/WFS endpoints (orthophoto, cadastral parcels, topography) and whether any are open without an account"
    contact: "Kosovo Cadastral Agency, https://akk.rks-gov.net/en; the geoportal's help or 'web services' pages"
  - kind: request
    what: "Access to orthophotos and cadastral parcels for the two pilot sites (Sharri National Park, Ligatina e Hencit/Radevës), for drawing boundaries at 1:10,000"
    contact: "Kosovo Cadastral Agency"
todo:
  doi: "No DOI; a government geoportal"
  licence: "Terms of use of geoportal data; read in the portal"
  access: "Whether registration or payment is needed; the desk review reported registration"
  format: "Formats offered (WMS, WFS, downloads); not established"
  access_method: "Not established: the WMS paths tried returned HTML, not capabilities"
  spatial_resolution: "Orthophoto resolution and cadastral scale; read in the portal"
  temporal_coverage: "Orthophoto years; read in the portal"
  latest_version: "Latest orthophoto and cadastre update"
  kosovo_codes_seen: "Not applicable (national portal)"
  effort: "Low if an open WMS exists; medium otherwise"
  sensitivity: "Cadastral parcels carry ownership information; use parcel geometry only"
---

```{r}
#| label: setup
source("../R/sources_helpers.R")
row <- source_row("kca-geoportal")
```

::: {.source-status}
`r status_line(row)`
:::

## At a glance

```{r}
#| label: glance
metadata_table(row)
```

## What it is

The Kosovo Cadastral Agency's State Geoportal is, in the Agency's own words,
a portal "which aims to provide complete geographic information, offering a
wide range of geographical and textual tools to users", presenting
geographic information from different sources in a unified form. It is the
natural source of the orthophotos and cadastral parcels that the
[Natura 2000 page](../natura2000.qmd#q2) says a site boundary should be
snapped to, so that it can be found on the ground and in the land register.

## Access

```{r}
#| label: access
#| echo: true
#| code-fold: show
access_table(list(
  try_access("https://geoportal.rks-gov.net/en/per-gjeoportalin/-/asset_publisher/EO0MH4mgPwJp/content"),
  try_access("https://geoportal.rks-gov.net/geoserver/ows?service=WMS&request=GetCapabilities"),
  try_access("https://geoportal.rks-gov.net/management/authentication/login")
))
```

The portal and its login page answer. The two service addresses guessed from
common geoportal software returned an HTML page rather than an OGC
capabilities document, so no open WMS or WFS was found. The portal's own
pages are built in the browser, and an automated read returned their titles
only. Nothing behind the login was attempted.

## Kosovo coverage check

::: {.callout-warning}
## To be completed
The geoportal is national, so coverage is not in doubt; what could not be
checked is what it serves and on what terms. Find the documented service
endpoints, request a GetCapabilities document, and list the layers with their
scale and year. If an open WMS exists, add it as a basemap to the report's
maps for boundary work.
:::

## What it adds for Natura 2000 in Kosovo

Site boundaries drawn along cadastral parcels and features visible on the
orthophoto are the ones that survive consultation and enforcement. The
geoportal is also where a future Natura 2000 layer would be published under
INSPIRE, so its service model is the one the national network should follow
(see [question 5](../natura2000.qmd#q5)). It serves the
[site tier](../natura2000.qmd#q1).

## Caveats and sensitivity

- Access terms are unknown.
- Parcel attributes include ownership; only the geometry is needed for
  boundary drawing.

## Integration into the pipeline

If an open WMS exists, low effort: add it to `add_basemaps()` in
`R/functions.R`. Parcel data would belong to the site-report tool
(`site_reports.R`), not to `pipeline.R`, which is not changed.

## To be completed

```{r}
#| label: todo
#| output: asis
todo_list(row)
```

## Citation

::: {.callout-warning}
## To be completed
Cite the geoportal as the Kosovo Cadastral Agency requests, once its terms are
read.
:::
