---
title: "Table of protected areas in Kosovo, 2020"
subtitle: "KEPA's national table of protected areas, with site codes that match the national identifiers Kosovo reports to the EEA."
actions:
  - kind: check
    what: "Resolve every code that appears in one list but not the other (see the check), with KEPA/KINP"
    contact: "Kosovo Institute for Nature Protection, KEPA"
  - kind: request
    what: "The national register in machine-readable form (table and boundaries), with the date of its last update"
    contact: "KEPA/AMMK, https://www.ammk-rks.net/"
  - kind: check
    what: "Whether a table newer than 2020 exists; the State of Nature report 2018–2021 is listed on the same KEPA page"
    contact: "https://www.ammk-rks.net/al/drejtorite/38/raporte-studime-publikime"
todo:
  doi: "No DOI; a national table"
  licence: "No terms stated in the PDF; ask KEPA before redistributing the table"
  temporal_coverage: "Range of designation years; read from the table's year column"
  sensitivity: "None expected for a register of sites"
---

```{r}
#| label: setup
source("../R/sources_helpers.R")
suppressPackageStartupMessages(library(sf))
row <- source_row("kepa-protected-areas-2020")
url <- "https://www.ammk-rks.net/assets/cms/uploads/files/Biodiversiteti/Zonat_e_Mbrojtura_2020.pdf"
dir <- cache_dir("kepa-protected-areas-2020")
pdf <- file.path(dir, "Zonat_e_Mbrojtura_2020.pdf")   # git-ignored: no licence stated
codes_csv <- file.path(dir, "codes_in_table.csv")
```

::: {.source-status}
`r status_line(row)`
:::

## At a glance

```{r}
#| label: glance
metadata_table(row)
```

## What it is

*Tabela e zonave të mbrojtura në Kosovë* is the Kosovo Environmental
Protection Agency's table of the country's protected areas, dated 2020 (the
PDF was produced in November 2021). It runs to 17 pages in Albanian and lists
each area by category (strict nature reserve, national park, natural
monument and so on) with a code, a name, the municipality or park it lies in,
its area in hectares, its IUCN management category, the year of protection
and a short description of its values. Codes take the form `RN_001` for
strict nature reserves and `NP_001` for national parks.

The report already uses the EEA's Nationally designated areas (NatDA v24),
which is how Kosovo reports its sites to Europe: 237 sites, 48 with a mapped
boundary. The desk review asked for a cross-check between the two.

## Access

```{r}
#| label: access
#| echo: true
#| code-fold: show
access_table(list(try_access(url), try_access(url, ssl_verify = FALSE)))
```

The server's TLS certificate had expired, so the file is fetched without
certificate verification, and its SHA-256 is recorded below. It states no
licence, so it is cached locally but not committed; only the list of codes
extracted from it is.

```{r}
#| label: build
#| echo: true
#| code-fold: true
#| code-summary: "Show how the codes are extracted"
if (!file.exists(codes_csv)) {
  if (!file.exists(pdf)) try_access(url, pdf, ssl_verify = FALSE)
  if (file.exists(pdf)) {
    txt <- paste(pdftools::pdf_text(pdf), collapse = "\n")
    codes <- sort(unique(regmatches(txt, gregexpr("\\b[A-Z]{2,4}_[0-9]{3}\\b", txt))[[1]]))
    write.csv(data.frame(code = codes), codes_csv, row.names = FALSE)
  }
}
have_cache <- file.exists(codes_csv)
```

```{r}
#| label: fingerprint
#| eval: !expr file.exists(pdf)
#| output: asis
fp <- file_fingerprint(pdf)
cat(sprintf("Local copy: `%s`, %s, SHA-256 `%s`, downloaded %s (not committed).\n",
            fp$file, fp$size, fp$sha256, fp$downloaded))
```

## Kosovo coverage check

The table is Kosovo's own, so coverage is complete by definition. The check
is whether it and NatDA describe the same set of sites.

```{r}
#| label: check
#| eval: !expr have_cache
table_codes <- read.csv(codes_csv)$code
natda <- rbind(
  st_drop_geometry(st_read(file.path(sources_root(), "data", "kosovo_protected_areas.gpkg"),
                           layer = "protected_area_polygons", quiet = TRUE))[, c("national_id", "site_name", "designated_area_type")],
  st_drop_geometry(st_read(file.path(sources_root(), "data", "kosovo_protected_areas.gpkg"),
                           layer = "protected_area_points", quiet = TRUE))[, c("national_id", "site_name", "designated_area_type")])
natda_codes <- unique(natda$national_id)
prefix <- function(x) sub("_.*", "", x)
cmp <- data.frame(
  Prefix = sort(unique(c(prefix(table_codes), prefix(natda_codes)))))
cmp$`In KEPA table` <- vapply(cmp$Prefix, function(p) sum(prefix(table_codes) == p), 0L)
cmp$`In NatDA v24` <- vapply(cmp$Prefix, function(p) sum(prefix(natda_codes) == p), 0L)
cmp$`In both` <- vapply(cmp$Prefix, function(p)
  length(intersect(table_codes[prefix(table_codes) == p], natda_codes)), 0L)
knitr::kable(cmp, align = "lrrr", row.names = FALSE)
```

```{r}
#| label: check-text
#| eval: !expr have_cache
#| output: asis
# The KEPA table abbreviates categories in Albanian (Rezervat Natyror, Park
# Kombëtar, Peizazh i Mbrojtur, Monument Natyror); NatDA mostly does not.
# Natural monuments appear in NatDA under both MN and NM, so both are tried.
num <- sub(".*_", "", table_codes)
cand <- lapply(seq_along(table_codes), function(i) switch(prefix(table_codes[i]),
  RN = paste0("NR_", num[i]), PK = paste0("NP_", num[i]),
  PM = paste0("PL_", num[i]), MN = paste0(c("MN_", "NM_"), num[i]),
  table_codes[i]))
hit_literal <- table_codes %in% natda_codes
hit_translated <- vapply(cand, function(x) any(x %in% natda_codes), TRUE)
cat(sprintf(paste(
  "The KEPA table carries %d distinct codes and NatDA %d. Taken literally,",
  "only %d codes appear in both, because the two lists abbreviate the",
  "categories differently: strict reserves are RN in the table and NR in",
  "NatDA, national parks PK and NP, protected landscapes PM and PL. Translating",
  "the table's prefixes to NatDA's (and trying both MN and NM for natural",
  "monuments) matches %d of the table's %d codes. Still unmatched in the",
  "table: %s."),
  length(table_codes), length(natda_codes), sum(hit_literal),
  sum(hit_translated), length(table_codes),
  paste(head(table_codes[!hit_translated], 20), collapse = ", ")))
```

A code with no counterpart is not necessarily an error: text extraction from
a PDF table can split or merge cells, and NatDA may hold sites designated
after 2020. But the prefix mismatch itself is a finding. A join on the site
code between the national table and the EEA reporting fails silently for all
but a few natural monuments, so any tool that links them must translate the
prefixes first, and the translation above is an inference to be confirmed by
KEPA.

## What it adds for Natura 2000 in Kosovo

A Natura 2000 proposal starts from the existing national sites, and the
report's figures on how much is protected rest on NatDA. If the national table
and NatDA disagree, the report's coverage figures inherit the disagreement.
The table also carries the IUCN category and short descriptions of each
site's values, a useful first pointer to what each site was protected for. It
serves the [national tier](../natura2000.qmd#q1); boundaries come from NatDA.

## Caveats and sensitivity

- Codes were extracted automatically from a PDF, and the extraction is only
  as good as the PDF's text layer.
- The table has no boundaries; any area comparison must use NatDA.
- No sensitive content.

## Integration into the pipeline

Low effort. The codes already match NatDA's `national_id`, so a proposed
`check_register_against_natda(table_codes, protected_areas)` could run after
`kosovo_protected_areas()` in `pipeline.R` and report mismatches in the run
log. A machine-readable table from KEPA would make the PDF step unnecessary.
`pipeline.R` is not changed here.

## To be completed

```{r}
#| label: todo
#| output: asis
todo_list(row)
```

## Citation

::: {.callout-warning}
## To be completed
The PDF gives no citation. Cite as KEPA (2020), *Tabela e zonave të mbrojtura
në Kosovë*, Agjencia për Mbrojtjen e Mjedisit të Kosovës, and confirm the
preferred form with KEPA.
:::
