---
title: "Kosovo National Forest Inventory 2012"
subtitle: "The second national forest inventory re-measured permanent plots across all of Kosovo's forest; only a four-page abstract is public."
actions:
  - kind: request
    what: "Plot-level data of the 2012 inventory (and 2003): plot coordinates, tree species, stand structure, dead wood and damage codes, under a data-use agreement"
    contact: "Kosovo Forest Agency (Ministry of Agriculture, Forestry and Rural Development); the Norwegian Forestry Group, which hosts the abstract, can say who holds the data"
  - kind: check
    what: "Whether a full NFI 2012 report exists in print or PDF, and whether a third inventory is planned or under way"
    contact: "Kosovo Forest Agency"
todo:
  custodian: "The abstract names no data holder; confirm with the Kosovo Forest Agency"
  doi: "No DOI expected; confirm once the full report is found"
  licence: "Terms for plot data; to be set in a data-use agreement with the holder"
  access: "Not established: plot data are not published, and no request route is stated"
  effort: "Estimate once the plot data format is known"
  sensitivity: "Plot coordinates of national inventories are often kept confidential to keep plots unbiased; ask the holder"
---

```{r}
#| label: setup
source("../R/sources_helpers.R")
row <- source_row("kosovo-nfi-2012")
```

::: {.source-status}
`r status_line(row)`
:::

## At a glance

```{r}
#| label: glance
metadata_table(row)
```

## What it is

Kosovo's second National Forest Inventory (NFI) was carried out in 2012. The
public abstract, hosted by the Norwegian Forestry Group, says that "for the
first time, Kosovo's entire forest area has been assessed". Of 3,453 permanent
sample plots laid across the country, 1,860 forest plots were visited in six
months of 2012 by twelve foresters, re-measuring trees mapped in the first
inventory of 2003. The abstract reports 481,000 ha of forest, 45 per cent of
the land, with a growing stock of about 40 million m³ that has stayed stable
since 2003. It also records that almost 99 per cent of forests show traces of
human activity, that fire affected over 12,000 ha, and that the two national
parks hold about 90 per cent of the coniferous growing stock.

## Access

```{r}
#| label: access
#| echo: true
#| code-fold: show
access_table(list(
  try_access("https://nfg.no/wp-content/uploads/2024/06/Kosovo-National-Forest-Inventory-2012-abstract.pdf")
))
```

The abstract downloaded as a 1 MB PDF of four pages. It contains summary
tables and a map of plot locations drawn as a figure; no plot data, table of
coordinates or link to a dataset. It names no data holder.

## Kosovo coverage check

::: {.callout-warning}
## To be completed
The inventory is national by design and covers Kosovo entirely, but its data
are not accessible, so no check could be run. With plot data: count plots per
Annex I forest type candidate (from species composition and altitude), set
plot-level tree species against the EUNIS and CORINE screens, and map dead
wood and damage by municipality. Plot positions must stay unpublished if the
holder requires it.
:::

## What it adds for Natura 2000 in Kosovo

Forest is almost half of Kosovo, and forest habitat types will be a large
part of any Natura 2000 network. The NFI is the only systematic,
plot-based sample of that forest, with repeat measurements since 2003. Plot
data would ground-truth the land-cover screens, give a first estimate of the
area of forest types, and provide structure and dead-wood indicators for the
"structure and functions" parameter of habitat assessment. Summary figures
serve the [national tier](../natura2000.qmd#q1); plot data, with species
composition, could inform site selection.

## Caveats and sensitivity

- Only the abstract was seen; the method details and plot data were not.
- Plot coordinates are commonly withheld to prevent plots being treated
  differently from the forest around them.
- The figures are from 2012; forest cover has changed since (see
  [forest loss](forest-loss-hansen.qmd) and [burnt areas](effis-burnt-areas.qmd)).

## Integration into the pipeline

Medium effort if plot data arrive as a table with coordinates: a proposed
`read_nfi_plots(path)` would keep them outside `data_exports/` and use them
only for aggregated checks against the habitat screens. Nothing can be done
before a request succeeds. `pipeline.R` is not changed.

## To be completed

```{r}
#| label: todo
#| output: asis
todo_list(row)
```

## Citation

::: {.callout-warning}
## To be completed
The abstract gives no citation. Cite the full NFI report once found; until
then: *The Second Kosovo National Forest Inventory: Resources Remain Stable*
(abstract, Kosovo National Forest Inventory 2012), hosted by the Norwegian
Forestry Group.
:::
