---
title: "Occurrence data papers from Kosovo (Biodiversity Data Journal)"
subtitle: "Recent papers on new records and new species from Kosovo, and whether their data already reach GBIF."
actions:
  - kind: check
    what: "Extend the search beyond titles and beyond BDJ: papers with Kosovo records in other journals (ZooKeys, Check List, Hacquetia, Acta Zoologica Bulgarica, national journals), and whether each deposited data"
    contact: "Europe PMC and Crossref searches; Pensoft journals publish data to GBIF automatically"
  - kind: check
    what: "For the 2014 paper without a GBIF dataset (doi:10.3897/bdj.2.e4140), check whether its records are in GBIF under another dataset"
    contact: "https://www.gbif.org/occurrence/search?country=XK&taxon_key=Limnephilus"
  - kind: check
    what: "Re-run the check after the next GBIF download, so papers published after the current extract are picked up"
    contact: "This page's build chunk"
todo:
  sensitivity: "New species are known from one or two sites; check whether any paper withheld coordinates"
---

```{r}
#| label: setup
source("../R/sources_helpers.R")
row <- source_row("kosovo-data-papers")
cache <- file.path(cache_dir("kosovo-data-papers"), "bdj_kosovo_gbif_check.csv")
```

::: {.source-status}
`r status_line(row)`
:::

## At a glance

```{r}
#| label: glance
metadata_table(row)
```

## What it is

Kosovo's biodiversity is being described paper by paper. The *Biodiversity
Data Journal* (BDJ), published by Pensoft, is one of the main venues: its
articles are open under CC BY 4.0, and Pensoft publishes each article's
occurrence records to GBIF as a dataset of its own. The desk review's example,
Krasniqi & Berisha (2025), records *Anchusella cretica* for the first time in
Kosovo, a new genus and species for the national flora, from four
populations found in 2024–2025 and assessed as Near Threatened nationally.
Most of the other Kosovo papers in BDJ describe caddisflies, including
species new to science from Kosovo's mountain rivers.

The question for this page is not whether the papers exist but whether their
records are already in the GBIF extract the report is built on.

## Access

```{r}
#| label: access
#| echo: true
#| code-fold: show
access_table(list(
  try_access("https://bdj.pensoft.net/"),
  try_access("https://www.ebi.ac.uk/europepmc/webservices/rest/search?query=JOURNAL:%22Biodiversity%20Data%20Journal%22%20AND%20TITLE:Kosovo&format=json"),
  try_access("https://api.gbif.org/v1/dataset/doi/10.3897/bdj.13.e164900")
))
```

PubMed Central served a reCAPTCHA to automated reads, so the papers were
listed through the Europe PMC API instead, which returns the same records.

## Kosovo coverage check

For every BDJ paper with Kosovo in its title, the check looks up GBIF's
dataset for the article's DOI, counts its records inside Kosovo (GADM XKO),
notes the country codes they carry, and tests whether the dataset is among
those in the report's download.

```{r}
#| label: build
#| echo: true
#| code-fold: true
#| code-summary: "Show the check"
if (!file.exists(cache)) {
  res <- tryCatch({
    papers <- jsonlite::fromJSON(paste0(
      "https://www.ebi.ac.uk/europepmc/webservices/rest/search?query=",
      "JOURNAL:%22Biodiversity%20Data%20Journal%22%20AND%20TITLE:Kosovo",
      "&format=json&pageSize=100"))$resultList$result
    reg <- read.csv(file.path(sources_root(), "data", "dataset_registry.csv"))
    do.call(rbind, lapply(seq_len(nrow(papers)), function(i) {
      doi <- papers$doi[i]
      ds  <- jsonlite::fromJSON(paste0("https://api.gbif.org/v1/dataset/doi/", doi))$results
      key <- if (length(ds) && NROW(ds)) ds$key[1] else NA
      occ <- if (is.na(key)) NULL else jsonlite::fromJSON(sprintf(
        "https://api.gbif.org/v1/occurrence/search?datasetKey=%s&limit=300&facet=country", key))
      codes <- if (is.null(occ) || !length(occ$facets$counts)) "" else
        paste(occ$facets$counts[[1]]$name, collapse = ", ")
      u <- if (is.null(occ)) NULL else occ$results$coordinateUncertaintyInMeters
      unc <- if (is.null(u) || all(is.na(u))) NA_real_ else stats::median(u, na.rm = TRUE)
      xko <- if (is.na(key)) NA else jsonlite::fromJSON(sprintf(
        "https://api.gbif.org/v1/occurrence/search?datasetKey=%s&gadmGid=XKO&limit=0", key))$count
      data.frame(year = papers$pubYear[i], doi = doi,
                 title = trimws(gsub("[[:space:]]+", " ", gsub("<[^>]+>|&lt;[^&]*&gt;", " ",
                                                         papers$title[i]))),
                 gbif_dataset = key, records_total = if (is.null(occ)) NA else occ$count,
                 records_in_kosovo = xko, country_codes = codes,
                 median_uncertainty_m = unc,
                 in_report_download = !is.na(key) & key %in% reg$datasetKey)
    }))
  }, error = function(e) NULL)
  if (!is.null(res)) write.csv(res, cache, row.names = FALSE)
}
have_cache <- file.exists(cache)
```

```{r}
#| label: check
#| eval: !expr have_cache
d <- read.csv(cache)
d <- d[order(-d$year), ]
knitr::kable(data.frame(
  Year = d$year, Paper = sprintf("[%s](https://doi.org/%s)", substr(d$title, 1, 70), d$doi),
  `GBIF dataset` = ifelse(is.na(d$gbif_dataset), "none found",
                          sprintf("[yes](https://www.gbif.org/dataset/%s)", d$gbif_dataset)),
  `Records in Kosovo` = d$records_in_kosovo, `Country codes` = d$country_codes,
  `Median uncertainty (m)` = ifelse(is.na(d$median_uncertainty_m), "not stated",
                                    format(d$median_uncertainty_m, big.mark = ",")),
  `In the report's download` = ifelse(d$in_report_download, "yes", "no"),
  check.names = FALSE), align = "lllllll", row.names = FALSE)
```

```{r}
#| label: check-text
#| eval: !expr have_cache
#| output: asis
n_ds <- sum(!is.na(d$gbif_dataset))
cat(sprintf(paste(
  "%d of %d papers have a GBIF dataset, holding %s records inside Kosovo in",
  "all. %d of those %d datasets are in the report's download%s. Country",
  "codes seen: %s."),
  n_ds, nrow(d), fmt_int(sum(d$records_in_kosovo, na.rm = TRUE)),
  sum(d$in_report_download), n_ds,
  if (sum(d$in_report_download) < n_ds) "; the rest were published after it or fell outside its GADM filter" else "",
  paste(sort(unique(unlist(strsplit(d$country_codes[nzchar(d$country_codes)], ", ")))), collapse = ", ")))
```

```{r}
#| label: no-cache
#| eval: !expr "!have_cache"
#| output: asis
cat("::: {.callout-warning}\n## To be completed\nThe Europe PMC or GBIF API did not answer at render time; re-render to run the check.\n:::\n")
```

## What it adds for Natura 2000 in Kosovo

New national records and newly described species are exactly the features a
Natura 2000 gap analysis must not miss, and they are usually known from a
single site. Knowing which of them are already in GBIF, and so in the report,
tells the reader whether the report's species lists are current. None of the
paper datasets states a coordinate uncertainty, so by the report's rules they
sit at the [national screening tier](../natura2000.qmd#q1) until someone
checks each locality against the paper; a verified locality could then reach
the delineation tier.

## Caveats and sensitivity

- The search covers BDJ titles containing "Kosovo" only; papers that name a
  mountain or river instead are missed.
- A dataset on GBIF is not necessarily in the report's download: the
  download is a dated snapshot.
- Newly described species with one known site are vulnerable to collecting;
  check before mapping them publicly.

## Integration into the pipeline

Low effort: the records already reach GBIF. What the pipeline lacks is a
refresh. Re-running the download (a new DOI) picks them up, and a proposed
`check_recent_datasets(since = download_date)` could list GBIF datasets with
Kosovo records created after the snapshot, so the report can say what it
does not yet include. `pipeline.R` is not changed.

## To be completed

```{r}
#| label: todo
#| output: asis
todo_list(row)
```

## Citation

Cite each paper by its own DOI, and each GBIF dataset as GBIF shows it. The
example paper: Krasniqi, E. & Berisha, N. (2025). *Anchusella cretica*
(Boraginaceae): a new genus and species record for the flora of Kosovo
(Southeast Europe). *Biodiversity Data Journal* 13.
[doi:10.3897/bdj.13.e164900](https://doi.org/10.3897/bdj.13.e164900).
