---
title: "Natura 2000 research in Montenegro: species reports"
subtitle: "Montenegro's Environmental Protection Agency is finalising its first draft Natura 2000 network from eight years of habitat and species fieldwork."
actions:
  - kind: request
    what: "The final species reports of the Natura 2000 research project, and the draft site proposal for areas bordering Kosovo (Prokletije, Hajla, Mokra Gora)"
    contact: "Environmental Protection Agency of Montenegro, https://epa.org.me/"
  - kind: check
    what: "Whether EPA Montenegro publishes the species reports itself; the only copies found were uploads on academia.edu and ResearchGate, both HTTP 403"
    contact: "https://epa.org.me/ publications section"
todo:
  doi: "No DOI found; the reports are known only from third-party uploads"
  licence: "Terms of the species reports; ask EPA Montenegro"
  access: "Not established: the uploads are blocked and EPA has not published the reports"
  format: "Reports are PDFs according to the upload titles; the underlying data format is unknown"
  access_method: "Unknown until EPA Montenegro says how data are shared"
  spatial_resolution: "Resolution of the species survey data; ask EPA Montenegro"
  temporal_coverage: "Years of fieldwork; EPA says eight years of research without giving the years"
  kosovo_codes_seen: "Not applicable unless the Montenegrin data extend across the border; check once obtained"
  tier: "Depends on the precision of the survey records"
  effort: "Estimate once the data format is known"
  sensitivity: "Ask EPA Montenegro which species it treats as sensitive (large carnivores, raptors)"
---

```{r}
#| label: setup
source("../R/sources_helpers.R")
row <- source_row("montenegro-n2k")
```

::: {.source-status}
`r status_line(row)`
:::

## At a glance

```{r}
#| label: glance
metadata_table(row)
```

## What it is

Montenegro is preparing its Natura 2000 network ahead of accession. On
25 March 2026 its Environmental Protection Agency (EPA) reported two days of
expert meetings to finalise the first draft proposal of Natura 2000 areas.
The work brought together the EPA, the EU IPA project "Support for the
establishment of the Natura 2000 network for Montenegro", run by Eptisa, and
experts who had carried out eight years of field research mapping habitats
and species. The EPA describes Natura 2000 as "one of the key requirements"
of Montenegro's accession, and part of the closing benchmark for Chapter 27.

The "species reports" named in the desk review come from that research. The
only copies found are uploads by individuals on academia.edu and
ResearchGate. Neither could be opened, so their content is not described here.

## Access

```{r}
#| label: access
#| echo: true
#| code-fold: show
access_table(list(
  try_access("https://www.academia.edu/39138586/Research_on_Natura_2000_network_Montenegro_Species_Reports"),
  try_access("https://www.researchgate.net/publication/337448196_Natura_2000_network_Montenegro_Species_Reports_2019"),
  try_access("https://epa.org.me/2026/03/25/natura-2000-kao-kljucni-korak-ka-eu-odrzani-ekspertski-sastanci/"),
  try_access("https://www.eptisa.com/en/new-project-support-for-the-establishment-of-the-natura-2000-network-for-montenegro/")
))
```

Both upload sites refused the request with HTTP 403, from the R client and
from a second fetch tool alike. The EPA news item and Eptisa's project page
answered; neither links to reports, data or maps. `natura2000.me`, a domain
that suggests a dedicated portal, answered HTTP 401 (authentication required).

## Kosovo coverage check

::: {.callout-warning}
## To be completed
No data were accessible, so no check was run. By its nature the Montenegrin
work stops at the border, but its fieldwork on Prokletije, Hajla and Mokra
Gora is the nearest recent survey of the same mountains. Once reports or data
are obtained: list the survey localities within 20 km of Kosovo and the
species recorded there, and compare them with the Kosovo GBIF extract. The
EEA's Emerald data already place Montenegrin candidate sites at Hajla and
Plavsko-Gusinjske Prokletije next to Kosovo (see
[Emerald Network](emerald-network.qmd)).
:::

## What it adds for Natura 2000 in Kosovo

Bjeshkët e Nemuna/Prokletije is one massif across Kosovo, Montenegro and
Albania. A Montenegrin site proposal there, with its species lists and
population figures, gives Kosovo both a transboundary partner and an
independent reference for what the mountain holds. It serves the
[national screening tier](../natura2000.qmd#q1) at least, and possibly the
site tier, depending on the precision of the survey data.

## Caveats and sensitivity

- Nothing about the reports' content was verified, only that the programme
  exists and that a first draft network is being finalised.
- Survey data for large carnivores and raptors may be restricted.
- Uploads on academia.edu are not a primary source; the EPA is.

## Integration into the pipeline

Nothing to integrate until data are released. If the EPA supplies point data
with a licence that permits use, a proposed `neighbour_survey_records()` step
would read them beside the GBIF extract, clearly separated from Kosovo's own
records. `pipeline.R` is not changed.

## To be completed

```{r}
#| label: todo
#| output: asis
todo_list(row)
```

## Citation

::: {.callout-warning}
## To be completed
No citation guidance was found. Ask EPA Montenegro how the reports are to be
cited.
:::
