---
title: "Natura 2000 potential areas in Kosovo"
subtitle: "The first attempt to identify Natura 2000 sites in Kosovo, drafted in 2008–2009 under an EU forestry project and known today only from citations."
actions:
  - kind: request
    what: "A copy of the report 'Natura 2000 potential areas in Kosovo' (Mustafa 2009) and any maps or site list it contains"
    contact: "Kosovo Institute for Nature Protection, which the NBSAP names as responsible for the ecological network; and the author, Behxhet Mustafa"
  - kind: check
    what: "If the report contains maps of the potential areas, record their scale and whether digitising them is worthwhile; do not digitise them without the author's agreement"
    contact: "The report, once obtained"
  - kind: check
    what: "Find Administrative Instruction No. 03/156 (2013) on the proclamation of the ecological network, which the NBSAP cites, and whether it maps the network"
    contact: "Official Gazette, https://gzk.rks-gov.net/"
todo:
  custodian: "Holder of the report; the NBSAP says it was drafted within the Sustainable Forest Management project"
  url: "No online copy found; the report is known from citations only"
  doi: "No DOI expected for a project report"
  licence: "Terms of the report; ask the holder"
  access: "Not established: no copy found"
  format: "Report, possibly with maps; not seen"
  access_method: "Unknown until a copy is found"
  spatial_resolution: "Scale of any maps in the report"
  temporal_coverage: "Drafted 2008 (NBSAP) or 2009 (as cited); data years unknown"
  latest_version: "2009 as cited in Veselaj & Mustafa (2015)"
  kosovo_codes_seen: "Not applicable (national report)"
  effort: "Depends on whether maps exist and in what form"
  sensitivity: "Unknown until seen"
---

```{r}
#| label: setup
source("../R/sources_helpers.R")
row <- source_row("mustafa-2009-potential-n2k")
```

::: {.source-status}
`r status_line(row)`
:::

## At a glance

```{r}
#| label: glance
metadata_table(row)
```

## What it is

Two sources, one of them governmental, confirm that a first identification of
potential Natura 2000 sites in Kosovo was made more than fifteen years ago.
Kosovo's Action Plan for Biodiversity 2016–2020 records that "already in 2008,
a report of preliminary sites in Kosovo has been drafted within the 'Project
of Sustainable Forest Management'". Veselaj and Mustafa's overview of nature
protection progress in Kosovo (*Landscape Online*, 2015) cites the work as
Mustafa, B. (2009), *Natura 2000 Potential Areas in Kosovo*. Neither source
says how many areas were identified or where, and no copy of the report was
found online.

The same action plan cites an Administrative Instruction No. 03/156 of 2013
on the proclamation of a national ecological network. It gives the Kosovo
Institute for Nature Protection responsibility for that network, and so for
Natura 2000 preparation.

## Access

```{r}
#| label: access
#| echo: true
#| code-fold: show
access_table(list(
  try_access("https://landscape-online.org/index.php/lo/article/view/LO.201545"),
  try_access("https://kryeministri.rks-gov.net/wp-content/uploads/2022/08/Eng-SAPB-2016-2020.pdf")
))
```

A web search for the title found only citations. The 2015 article is open
(CC BY 4.0) and cites the report in its references without describing it.

## Kosovo coverage check

::: {.callout-warning}
## To be completed
The report is about Kosovo, but its content is unknown: no site list, map or
area figure was found. Once a copy is obtained: list the potential areas,
compare them with today's protected-area register and with the report's
habitat screen, and note which have since been designated. If the areas exist
only as a map figure, digitising it is an open action for the author's
agreement, not something this page does.
:::

## What it adds for Natura 2000 in Kosovo

A starting list. Whatever its method, a set of areas that Kosovo's own
experts judged to be of Natura 2000 interest in 2008–2009 is the obvious
first test for today's data: are those areas still where the evidence is, and
what did they miss? It also matters for the record, since the European
Commission's 2025 report calls the identification of potential sites "at a
very early stage". The [tier](../natura2000.qmd#q1) depends on the map scale.

## Caveats and sensitivity

- Documented only: known from two citations, not seen.
- The two sources give different years (2008 drafted, 2009 cited); probably
  the same work, not confirmed.

## Integration into the pipeline

None until a copy exists. If the report includes digital boundaries, a
proposed `read_candidate_areas(path)` would load them beside the protected
areas for comparison. `pipeline.R` is not changed.

## To be completed

```{r}
#| label: todo
#| output: asis
todo_list(row)
```

## Citation

As cited in Veselaj & Mustafa (2015): Mustafa, B. (2009). *Natura 2000
Potential Areas in Kosovo.*

::: {.callout-warning}
## To be completed
Publisher, place and full author list were not verified; complete from the
report itself.
:::
