---
title: "Biodiversity Strategy and Action Plan 2016–2020"
subtitle: "Kosovo's action plan for biodiversity, in English, with the state's own account of its data gaps and its first steps towards Natura 2000."
actions:
  - kind: check
    what: "Whether a successor biodiversity strategy or action plan after 2020 has been adopted or drafted; the EC Kosovo Report 2025 says the environment strategy for 2024–2033 is still pending"
    contact: "Ministry of Environment, Spatial Planning and Infrastructure"
  - kind: check
    what: "Whether the action plan's Natura 2000 measures (ecological network, habitat mapping, inventories) were reported on at the end of 2020"
    contact: "Ministry of Environment, Spatial Planning and Infrastructure; KEPA"
todo:
  doi: "No DOI; a government document"
  licence: "No terms stated; a public government document"
  spatial_resolution: "Not spatial: a strategy document"
  effort: "Low: no data to integrate; cite as context"
  sensitivity: "None"
---

```{r}
#| label: setup
source("../R/sources_helpers.R")
row <- source_row("nbsap-2016-2020")
url <- "https://kryeministri.rks-gov.net/wp-content/uploads/2022/08/Eng-SAPB-2016-2020.pdf"
```

::: {.source-status}
`r status_line(row)`
:::

## At a glance

```{r}
#| label: glance
metadata_table(row)
```

## What it is

The Ministry of Environment and Spatial Planning's Action Plan for
Biodiversity 2016–2020 revises the implementation part of Kosovo's Strategy
and Action Plan for Biodiversity 2011–2020, which the Assembly approved on 7
October 2011. The English PDF runs to 78 pages. It is interesting less for
its measures than for its candour about the starting point. Among the main
problems it lists are the lack of full inventories for flora, fauna and
habitats (except the Red Book of Vascular Flora), the lack of a red list of
fauna, and the lack of national biodiversity indicators.

On Natura 2000 it records three facts that other pages in this section
depend on. Kosovo "is not a Party to the Bern Convention", so no Emerald
network was developed. A report of preliminary Natura 2000 sites was drafted
in 2008 under the Sustainable Forest Management project. And an
administrative instruction of 2013 on a national ecological network put the
Kosovo Institute for Nature Protection in charge of establishing it.

## Access

```{r}
#| label: access
#| echo: true
#| code-fold: show
access_table(list(try_access(url)))
```

Open PDF on the Office of the Prime Minister's site.

## Kosovo coverage check

The plan is national. There are no records to count; the useful check is
whether its commitments have a successor.

::: {.callout-warning}
## To be completed
No successor biodiversity strategy was found in this check. The European
Commission's Kosovo Report 2025 notes that "the adoption of a strategy for
environmental protection and sustainable development for 2024–2033 remains
pending". Confirm with the Ministry whether biodiversity measures after 2020
sit in that pending strategy or in a separate plan.
:::

## What it adds for Natura 2000 in Kosovo

It is the state's own statement of where Natura 2000 preparation stood and
who is responsible, which a proposal must build on and a reviewer will read.
It also confirms from a primary source what the desk review expected about
the Emerald Network (see [Emerald Network](emerald-network.qmd)). It serves
the [national tier](../natura2000.qmd#q1) as context only.

## Caveats and sensitivity

- It covers 2016–2020; nothing here says what happened after.
- A plan records intentions, not results.
- No sensitive content.

## Integration into the pipeline

None; a document to cite. `pipeline.R` is not changed.

## To be completed

```{r}
#| label: todo
#| output: asis
todo_list(row)
```

## Citation

Ministry of Environment and Spatial Planning. *Action Plan for Biodiversity
2016–2020, Strategy and Action Plan for Biodiversity 2011–2020.* Department
of Environment Protection, Prishtina. No publication date was seen on the
cover; the PDF was produced in June 2018.
