---
title: "Emerald Network sufficiency in neighbouring countries"
subtitle: "How far Serbia, North Macedonia, Montenegro and Albania have got in proving their Emerald lists sufficient, and which of their sites border Kosovo."
actions:
  - kind: check
    what: "Find the primary Bern Convention document with each country's sufficiency index (the Standing Committee's decisions and the conclusions of the biogeographical seminars); coe.int returned HTTP 403 to every request from this network"
    contact: "https://www.coe.int/en/web/bern-convention/conclusions-of-the-biogeographical-seminars, from another connection"
  - kind: check
    what: "Whether Serbia's sufficiency figures count the candidate sites that lie inside Kosovo (see the Emerald Network page)"
    contact: "Bern Convention Secretariat"
todo:
  url: "Primary page for the sufficiency figures on coe.int; blocked from this network"
  doi: "No DOI expected for Standing Committee documents; confirm once the primary document is found"
  licence: "Terms of Council of Europe documents; read on coe.int"
  access: "Open according to the secondary source; confirm on coe.int"
  format: "Likely PDF meeting documents; confirm on coe.int"
  access_method: "Confirm on coe.int"
  spatial_resolution: "Not spatial: the index is a share of features judged sufficient"
  temporal_coverage: "Date of each country's latest biogeographical evaluation; read from the primary document"
  latest_version: "The 44th Standing Committee meeting (December 2024) is the latest cited by the secondary source; check for later decisions"
  kosovo_codes_seen: "Record whether any sufficiency statistic includes Kosovo territory under RS"
  tier: "Not a spatial dataset; the tier applies to the site layer, which is on the Emerald Network page"
  effort: "Low once the primary document is found: a four-row table"
  sensitivity: "None expected"
---

```{r}
#| label: setup
source("../R/sources_helpers.R")
suppressPackageStartupMessages(library(sf))
row <- source_row("neighbour-emerald-sufficiency")
cache <- file.path(cache_dir("neighbour-emerald-sufficiency"), "neighbour_sites_50km.csv")
service <- "https://bio.discomap.eea.europa.eu/arcgis/rest/services/ProtectedSites/EmeraldSites/MapServer/3"
```

::: {.source-status}
`r status_line(row)`
:::

## At a glance

```{r}
#| label: glance
metadata_table(row)
```

## What it is

Under the Bern Convention each Party proposes Emerald sites, and regional
biogeographical seminars then judge, species by species and habitat by
habitat, whether a country's list is sufficient. The share of features found
sufficient is reported as a sufficiency index. For Kosovo's neighbours the
index says how complete their networks are along the shared mountains, and
therefore how much of a transboundary network already exists on the other
side of the border.

The figures below come from a secondary source only. CEE Bankwatch Network
reported on 9 December 2024, after the 44th meeting of the Convention's
Standing Committee, that all Western Balkan countries must reach a 35 per
cent sufficiency target by 2030, and gave each country's current index. The
primary Council of Europe documents could not be opened.

| Country | Sufficiency index (Bankwatch, 9 December 2024) |
|---|---|
| Serbia | 13.5 % |
| North Macedonia | 16 % |
| Montenegro | 16.3 % |
| Albania | 28.3 % |

## Access

```{r}
#| label: access
#| echo: true
#| code-fold: show
access_table(list(
  try_access("https://bankwatch.org/press_release/bern-convention-western-balkan-countries-need-to-propose-new-protected-natural-areas"),
  try_access("https://www.coe.int/en/web/bern-convention/-/44th-standing-committee"),
  try_access("https://rm.coe.int/16807466b6"),
  try_access(paste0(service, "?f=json"))
))
```

The Council of Europe's web pages returned HTTP 403 throughout. Its document
server, `rm.coe.int`, did answer, so individual documents can be fetched once
their addresses are known; the calendar for the implementation of the
Emerald Network is one. The site layer itself is read from the EEA's copy of
the Emerald data.

## Kosovo coverage check

The sufficiency index is a national figure and says nothing about Kosovo
directly. What can be checked is what each neighbour has put on the ground
near Kosovo.

```{r}
#| label: build
#| echo: true
#| code-fold: true
#| code-summary: "Show how the neighbours' sites near Kosovo are counted"
build_error <- NULL
if (!file.exists(cache)) {
  build_error <- tryCatch({
    source("../R/functions.R", local = TRUE)
    ko   <- st_transform(st_union(kosovo_outline()), 3035)
    zone <- st_buffer(ko, 50000)
    s <- arcgis_query_sf(service, st_bbox(st_transform(zone, 4326)),
                         fields = "SITECODE,SITENAME,COUNTRY_CODE,SITESTATUS",
                         verbose = FALSE) |> st_make_valid()
    s <- s[lengths(st_intersects(s, zone)) > 0, ]
    ring <- st_difference(zone, ko)
    s$ha_in_ring <- vapply(seq_len(nrow(s)), function(i)
      sum(as.numeric(st_area(st_intersection(st_geometry(s)[i], ring)))), 0) / 1e4
    write.csv(st_drop_geometry(s), cache, row.names = FALSE)
    NULL
  }, error = function(e) conditionMessage(e))
}
have_cache <- file.exists(cache)
```

```{r}
#| label: check
#| eval: !expr have_cache
s <- read.csv(cache)
tab <- do.call(rbind, lapply(split(s, s$COUNTRY_CODE), function(x) data.frame(
  `Country code` = x$COUNTRY_CODE[1], Sites = nrow(x),
  `Status` = paste(sort(unique(x$SITESTATUS)), collapse = ", "),
  `Area in the 50 km ring outside Kosovo (km²)` = fmt_int(round(sum(x$ha_in_ring) / 100)),
  check.names = FALSE)))
knitr::kable(tab, align = "lrlr", row.names = FALSE)
```

```{r}
#| label: no-cache
#| eval: !expr "!have_cache"
#| output: asis
cat("::: {.callout-warning}\n## To be completed\nThe EEA map service did not answer at render time (",
    build_error %||% "no message", "). Re-render when it does.\n:::\n", sep = "")
```

```{r}
#| label: check-text
#| eval: !expr have_cache
#| output: asis
st <- table(s$SITESTATUS)
cat(sprintf("Status of the %d sites within 50 km of Kosovo: %s.", nrow(s),
            paste(sprintf("%s %d", names(st), as.integer(st)), collapse = ", ")))
if (all(s$SITESTATUS == "Candidate")) {
  cat(" None has been adopted. Read with the sufficiency figures, the",
      "transboundary network Kosovo would join is itself unfinished on every side.")
}
```

## What it adds for Natura 2000 in Kosovo

It sets expectations. If the neighbours have proved between 13 and 28 per
cent of their features sufficient, as the secondary source reports, a Kosovo
network will be judged in a region where few boundaries are settled. It also
flags a practical risk: the Serbian candidate sites inside Kosovo (see
[Emerald Network](emerald-network.qmd)) may enter Serbia's statistics, and a
Kosovo sufficiency ledger ([question 4](../natura2000.qmd#q4)) should not
count the same ground twice.

## Caveats and sensitivity

- The four percentages are secondary. Until the Council of Europe document
  is read, treat them as reported, not verified.
- Sufficiency is judged per biogeographical region; a single national
  figure hides regional differences.
- The index is a share of features, not of territory.

## Integration into the pipeline

No pipeline step is needed for the index itself: it belongs in the report as
a small table with its source and date. The site layer is covered by the
proposed `emerald_sites_near()` on the Emerald Network page. `pipeline.R` is
not changed.

## To be completed

```{r}
#| label: todo
#| output: asis
todo_list(row)
```

## Citation

::: {.callout-warning}
## To be completed
Cite the Bern Convention document that gives the indices once it is found.
Until then the source is: CEE Bankwatch Network (2024). *Bern Convention:
Western Balkan countries need to propose new protected natural areas.* Press
release, 9 December 2024 (secondary source).
:::
