---
title: "National park spatial plans"
subtitle: "The spatial plan for Bjeshkët e Nemuna National Park and the draft strategic environmental assessment of Sharri's plan, both published as PDFs."
actions:
  - kind: request
    what: "The zoning of both national parks (protection zones, special zones, planned development) as GIS layers"
    contact: "Institute for Spatial Planning, Ministry of Environment, Spatial Planning and Infrastructure"
  - kind: check
    what: "Whether the Sharri spatial plan itself (not only its SEA) is published, and whether the SEA has been finalised"
    contact: "MMPHI and KEPA websites"
  - kind: check
    what: "If the zoning exists only as map figures in the PDFs, record their scale; digitising them is an open action for the Institute's agreement, not something this page does"
    contact: "The two PDFs"
todo:
  doi: "No DOI; government plans"
  licence: "No terms stated in the plans; the Sharri SEA carries a UN-Habitat copyright notice allowing reproduction of excerpts with the source cited"
  spatial_resolution: "Scale of the zoning maps; read in the plans"
  temporal_coverage: "Planning horizon of each plan; read in the documents"
  effort: "Low with GIS layers; high if the zoning must be digitised from figures"
  sensitivity: "None expected: zoning and planned development"
---

```{r}
#| label: setup
source("../R/sources_helpers.R")
row <- source_row("np-spatial-plans")
bj <- "https://mmphi.rks-gov.net/mmphifolder/OtherDocuments/Plani%20Hap%C3%ABsinor%20PK%20Bjeshk%C3%ABt%20e%20Nemuna.pdf"
sh <- "https://www.ammk-rks.net/assets/cms/uploads/files/Draft%20Raporti%20-%20Vler%C3%ABsimi%20Strategjik%20Mjedisor%20p%C3%ABr%20Parkun%20Komb%C3%ABtar%20Sharri_.pdf"
```

::: {.source-status}
`r status_line(row)`
:::

## At a glance

```{r}
#| label: glance
metadata_table(row)
```

## What it is

Kosovo's two national parks, Bjeshkët e Nemuna in the west and Sharri in the
south, are the core of any future Natura 2000 network. Each is governed by a
spatial plan that zones the park and says where development may go.

The *Plani Hapësinor për Parkun Kombëtar "Bjeshkët e Nemuna"* is published by
the Ministry of Environment, Spatial Planning and Infrastructure and drafted by
its Institute for Spatial Planning; the PDF (89 pages, in Albanian) dates from
June 2023. It notes that the park law adopted by the Assembly on 13 December
2012 gave an area of 62,488 ha, revised to 63,028 ha after detailed
measurement, and it defines an impact zone of 50 m around the park boundary.

For Sharri the document found is a draft *Strategic Environmental Assessment*
of the park's spatial plan (161 pages, PDF dated March 2025), prepared with
UN-Habitat. It sets environmental objectives for the plan, among them
avoiding the fragmentation of habitats and impacts on the integrity of
"protected areas and Natura 2000 areas".

## Access

```{r}
#| label: access
#| echo: true
#| code-fold: show
access_table(list(
  try_access(bj),
  try_access(sh),
  try_access(sh, ssl_verify = FALSE)
))
```

The Bjeshkët e Nemuna plan is on the ministry's server (14.6 MB). The Sharri
SEA is on KEPA's server, whose certificate had expired; the third request
skips verification. Neither document links to GIS data.

## Kosovo coverage check

::: {.callout-warning}
## To be completed
The plans cover the two parks, which are in NatDA with mapped boundaries. The
check that matters, overlaying the parks' zoning with the report's habitat
screen and GBIF evidence, needs the zoning as GIS layers. Until then: record
the scale of the zoning maps in each PDF, and compare the Bjeshkët e Nemuna
area (63,028 ha) with the area NatDA gives the park.
:::

## What it adds for Natura 2000 in Kosovo

A Natura 2000 site inside a national park has to fit the park's zoning:
strict zones are candidate cores, and zones planned for tourism or
infrastructure are where an appropriate assessment will be needed. The plans
also show what the parks' managers already consider their key values. With
GIS layers, the zoning would serve the [site tier](../natura2000.qmd#q1)
directly.

## Caveats and sensitivity

- The Sharri document is a draft SEA, not the adopted plan.
- Plans record intentions and rules, not the state of the habitats.

## Integration into the pipeline

Belongs with the site-report tool rather than `pipeline.R`: a proposed
`read_park_zoning(path)` in `R/site_report.R` would add a zoning layer to the
two pilot site reports (see [question 8](../natura2000.qmd#q8)). `pipeline.R`
is not changed.

## To be completed

```{r}
#| label: todo
#| output: asis
todo_list(row)
```

## Citation

::: {.callout-warning}
## To be completed
Neither PDF gives a citation form in the pages read. Cite each by title,
issuing body and date: MMPHI, Instituti për Planifikim Hapësinor (2023),
*Plani Hapësinor për Parkun Kombëtar "Bjeshkët e Nemuna"*; MMPHI with
UN-Habitat (2025), *Raporti i Vlerësimit Strategjik Mjedisor për Planin
Hapësinor të Parkut Kombëtar Sharri* (draft).
:::
