---
title: "Protected and strictly protected wild species (AI 18/2012 and its successors)"
subtitle: "Kosovo's legal lists of protected wild species; the instruction of 2012 has since been followed by instructions of 2017 and 2020."
actions:
  - kind: check
    what: "Confirm the chain of repeal: whether AI 12/2017 repealed AI 18/2012, and whether AI 12/2020 is the instruction in force today"
    contact: "Official Gazette, https://gzk.rks-gov.net/ (search 'llojeve të egra')"
  - kind: check
    what: "Extract the species lists from the annexes of AI 12/2020 into a table (name, protected or strictly protected), and match them to the GBIF backbone"
    contact: "https://gzk.rks-gov.net/ActDocumentDetail.aspx?ActID=35969"
todo:
  doi: "Legal acts carry no DOI; cite by number and Gazette date"
  licence: "Official acts are public; confirm the Gazette's terms for reuse"
  access_method: "Read how the annexes are published on the Gazette (HTML or PDF)"
  spatial_resolution: "Not spatial: a legal list of species"
  temporal_coverage: "In force from its entry into force; read the date in the act"
  effort: "Low once the annex is a table; medium if it must be parsed from a PDF"
  sensitivity: "None: a public legal list"
---

```{r}
#| label: setup
source("../R/sources_helpers.R")
row <- source_row("protected-species-ai-18-2012")
```

::: {.source-status}
`r status_line(row)`
:::

## At a glance

```{r}
#| label: glance
metadata_table(row)
```

## What it is

Under the Law on Nature Protection (No. 03/L-233), the ministry responsible
for the environment declares by administrative instruction which wild
species are protected and which strictly protected. The desk review named
Administrative Instruction No. 18/2012, which the Red Book of Vascular Flora
(2013) also cites as one basis for its selection. It is not the current act.
The Official Gazette holds Administrative Instruction MEA No. 12/2020 "for the
declaration of protected and strictly protected wild species", published on
29 December 2020, which supersedes Administrative Instruction No. 12/2017 on
strictly protected wild species. Whether 12/2017 in turn repealed 18/2012 was
not confirmed.

## Access

```{r}
#| label: access
#| echo: true
#| code-fold: show
access_table(list(
  try_access("https://gzk.rks-gov.net/ActDetail.aspx?ActID=35969"),
  try_access("https://gzk.rks-gov.net/ActDocumentDetail.aspx?ActID=35969")
))
```

The Gazette pages answer and the act is public. The exact URL of 18/2012 was
not found: a search of the Gazette for it returned the 2017 and 2020
instructions and a related instruction on trade in protected species
(No. 26/2012), but not 18/2012 itself.

## Kosovo coverage check

::: {.callout-warning}
## To be completed
The act is Kosovo's own and applies to its whole territory. The check that
matters is a comparison of lists: extract the species in AI 12/2020's annexes,
resolve them against the GBIF backbone, and set them against Annexes II, IV
and V of the Habitats Directive and Annex I of the Birds Directive, which the
pipeline already holds in `data/eu_directives_species.csv`. Species strictly
protected nationally but absent from the annexes, and annex species with no
national protection, are both findings.
:::

## What it adds for Natura 2000 in Kosovo

Transposition of the Nature Directives includes making annex species legally
protected in national law. A table of the national lists, matched to the same
backbone as the annexes, shows how far that has gone and lets the report flag
records of nationally protected species beside the directive columns it
already carries. It serves the [national tier](../natura2000.qmd#q1).

## Caveats and sensitivity

- The act named by the desk review is not the one in force; the current one
  is 12/2020, subject to the repeal check above.
- Legal names may use older taxonomy; matching needs the same care as the
  annex lists (see `R/build_directive_list.R`).
- No sensitive content.

## Integration into the pipeline

Low to medium effort. A proposed `R/build_national_protection_list.R`, on the
pattern of `R/build_directive_list.R`, would parse the annex, resolve the
names and write `data/kosovo_protected_species.csv`; `pipeline.R` would add a
`nationalProtection` column. `pipeline.R` is not changed here.

## To be completed

```{r}
#| label: todo
#| output: asis
todo_list(row)
```

## Citation

Cite the act in force by its number and Gazette date: *Udhëzim Administrativ
MEA – Nr. 12/2020 për shpalljen e llojeve të egra të mbrojtura dhe strikt të
mbrojtura.* Gazeta Zyrtare e Republikës së Kosovës, 29 December 2020.
