---
title: "R packages for reaching these sources"
subtitle: "The R packages the pages in this section use, their current CRAN versions, what each reaches, and how each handles Kosovo."
actions:
  - kind: check
    what: "Install rredlist, eurostat and ows4R on the machine that renders the site (they are on CRAN but not installed there) and run their Kosovo examples"
    contact: "install.packages(c('rredlist', 'eurostat', 'ows4R'))"
  - kind: check
    what: "How rredlist (IUCN API) and eurostat code Kosovo in their results"
    contact: "Package documentation; the IUCN API needs a token"
todo:
  doi: "CRAN packages carry DOIs of the form 10.32614/CRAN.package.<name>; record them per package if cited"
  licence: "Each package's licence is on its CRAN page"
  spatial_resolution: "Not applicable: software"
  temporal_coverage: "Not applicable: software"
  latest_version: "Per package, in the table below (CRAN, 29 September 2026)"
  tier: "Not applicable: software"
  n2k_question: "Not applicable: this page supports every other page"
  sensitivity: "None"
---

```{r}
#| label: setup
source("../R/sources_helpers.R")
row <- source_row("r-access-toolkit")
cache <- file.path(cache_dir("r-access-toolkit"), "cran_versions.csv")
```

::: {.source-status}
`r status_line(row)`
:::

## At a glance

```{r}
#| label: glance
metadata_table(row)
```

## What it is

Every page in this section reaches its data from R, with a handful of
packages. This page lists them with the version CRAN offered on the day of
the check and the version installed where the site was rendered, what each
reaches, and, because it decides whether a Kosovo query returns anything, how
each identifies Kosovo. `rgbif` and `CoordinateCleaner` are already used by
`pipeline.R` and are listed for completeness.

## Access

```{r}
#| label: access
#| echo: true
#| code-fold: show
access_table(list(try_access("https://cloud.r-project.org/src/contrib/PACKAGES")))
```

```{r}
#| label: build
#| echo: true
#| code-fold: true
#| code-summary: "Show how the versions are read"
pk <- c("rgbif", "CoordinateCleaner", "sf", "terra", "geodata", "giscoR",
        "eurostat", "rredlist", "ows4R", "leaflet", "curl", "pdftools",
        "eunis.habitats", "rnaturalearth")
if (!file.exists(cache)) {
  ap <- tryCatch(available.packages(repos = "https://cloud.r-project.org"),
                 error = function(e) NULL)
  if (!is.null(ap)) {
    ip <- installed.packages()
    write.csv(data.frame(
      package = pk, cran = ap[pk, "Version"],
      installed = ifelse(pk %in% rownames(ip), ip[pmatch(pk, rownames(ip)), "Version"], "not installed"),
      checked = format(Sys.Date())), cache, row.names = FALSE)
  }
}
have_cache <- file.exists(cache)
```

```{r}
#| label: table
#| eval: !expr have_cache
v <- read.csv(cache)
reach <- c(
  rgbif = "GBIF occurrence downloads and APIs (the report's data)",
  CoordinateCleaner = "Automated coordinate tests (the pipeline's cleaning)",
  sf = "Vector data; GDAL, including /vsicurl/ remote reads",
  terra = "Raster data; remote crops of COGs and tiled GeoTIFFs",
  geodata = "Per-country climate, elevation and boundary downloads",
  giscoR = "Eurostat GISCO boundaries",
  eurostat = "Eurostat statistics",
  rredlist = "IUCN Red List API (token needed)",
  ows4R = "OGC web services (WMS, WFS, CSW) from R",
  leaflet = "The interactive maps on these pages",
  curl = "HTTP requests behind try_access()",
  pdftools = "Text from PDF reports and tables",
  eunis.habitats = "EUNIS codes and crosswalks, including to Annex I",
  rnaturalearth = "Natural Earth boundaries (the pipeline's last fallback)")
kosovo <- c(
  rgbif = "Country code XK; GADM GID XKO for geometry filters",
  CoordinateCleaner = "No usable Kosovo code in its default reference; the pipeline supplies its own polygon",
  sf = "None: geometry only",
  terra = "None: geometry only",
  geodata = "XKO (ISO3) and XK (ISO2), as GADM",
  giscoR = "No Kosovo feature in the 2024 countries layer: Kosovo lies inside Serbia (RS/SRB)",
  eurostat = "Not checked",
  rredlist = "Not checked",
  ows4R = "None: depends on the service",
  leaflet = "None",
  curl = "None",
  pdftools = "None",
  eunis.habitats = "None: habitat codes only",
  rnaturalearth = "adm0_a3 'KOS'; iso_a3 '-99' (no ISO code), per R/functions.R")
knitr::kable(data.frame(Package = v$package, CRAN = v$cran, Installed = v$installed,
                        Reaches = reach[v$package], `How it identifies Kosovo` = kosovo[v$package],
                        check.names = FALSE), align = "lllll", row.names = FALSE,
             caption = sprintf("CRAN versions read on %s", v$checked[1]))
```

## Kosovo coverage check

The question for a toolkit is how each package names Kosovo, and three of the
answers were tested in this session rather than taken from documentation.

```{r}
#| label: check-geodata
#| echo: true
#| code-fold: show
cc <- geodata::country_codes()
cc[cc$NAME == "Kosovo", c("NAME", "ISO3", "ISO2")]
```

```{r}
#| label: check-gisco
#| echo: true
#| code-fold: show
g <- tryCatch(giscoR::gisco_get_countries(resolution = "20", year = "2024"),
              error = function(e) NULL)
if (!is.null(g)) {
  print(sf::st_drop_geometry(g[g$CNTR_ID %in% c("XK", "RS"), c("CNTR_ID", "ISO3_CODE", "NAME_ENGL")]))
  print(sf::st_bbox(g[g$CNTR_ID == "RS", ]))
}
```

GISCO's 2024 country layer has no feature for Kosovo, and Serbia's polygon
reaches south to about 41.86° N: Kosovo is inside it. The pipeline already
knows this (`R/functions.R` compares its border against GISCO's
international boundary only), but anyone who clips data to "Serbia" from
GISCO is clipping Kosovo in with it.

Across this section, Kosovo turned up under at least six different
identifiers: **XK** (GBIF, EU country code), **XKO** (GADM, geodata), **XKX**
(the pipeline's own ISO3-style code), **KS** (EFFIS), **RS** (Emerald
candidate sites, BirdLife's inventory and GISCO, where it is merged into
Serbia), and **Ko** (Euro+Med). A script that filters any of these sources
by one code will silently miss what the others file under another.

## What it adds for Natura 2000 in Kosovo

Reproducibility. Every figure on these pages can be recomputed with free
software from the sources' own servers, and each check is shown as code a
participant can run and change. Tier: not applicable.

## Caveats and sensitivity

- `rredlist`, `eurostat` and `ows4R` are on CRAN but were not installed where
  the site was rendered, so their Kosovo behaviour is not tested here.
- Versions move; the table records the day of the check.

## Integration into the pipeline

`pipeline.R` already loads `rgbif`, `sf`, `CoordinateCleaner` and friends.
The pages here would add `terra` (rasters) and `eunis.habitats` (crosswalk)
if their proposed functions are adopted. `pipeline.R` is not changed.

## To be completed

```{r}
#| label: todo
#| output: asis
todo_list(row)
```

## Citation

Cite each package with `citation("<package>")`; for example
`citation("eunis.habitats")` gives Magno, Monteiro & Bentes (2023),
doi:10.32614/CRAN.package.eunis.habitats.
