---
title: "Red Book of Fauna of the Republic of Kosovo"
subtitle: "The national red book of animals (Ibrahimi et al. 2018), published by the Ministry with KEPA and the Kosovo Institute for Nature Protection."
actions:
  - kind: request
    what: "The species list with national IUCN categories as a table, and permission to use the localities behind the maps at no finer than 10 km for threatened species"
    contact: "Kosovo Institute for Nature Protection (KEPA/AMMK), and the Ministry of Environment, Spatial Planning and Infrastructure, which holds the rights"
  - kind: check
    what: "Count the taxa per national category and per group from the PDF, and match the names to the GBIF backbone with rgbif, without redistributing the book"
    contact: "The PDF on ammk-rks.net"
  - kind: check
    what: "Report the expired TLS certificate on ammk-rks.net to KEPA"
    contact: "KEPA web administration"
todo:
  doi: "The colophon's ISBN field is empty and no DOI was found; ask the Ministry"
  temporal_coverage: "Years of the field data behind the assessments; read in the methods chapter"
---

```{r}
#| label: setup
source("../R/sources_helpers.R")
row <- source_row("red-book-fauna")
url <- "https://www.ammk-rks.net/assets/cms/uploads/files/Publikime-raporte/Libri_i_kuq.pdf"
```

::: {.source-status}
`r status_line(row)`
:::

## At a glance

```{r}
#| label: glance
metadata_table(row)
```

## What it is

*Libri i Kuq i Faunës së Republikës së Kosovës* is Kosovo's red book of
animals. It was published in Prishtinë in 2018 by the Ministry of Environment
and Spatial Planning, with the Kosovo Environmental Protection Agency (KEPA)
and the Kosovo Institute for Nature Protection, edited by Halil Ibrahimi with
nine co-authors from Kosovo and the region. The desk review cited it as
Ibrahimi et al. 2019, which matches the PDF's date of July 2019 rather than
the imprint year of 2018. It runs to 203 pages in Albanian, with chapters by
group (molluscs, oligochaetes, insects, fish, reptiles, birds and others),
national IUCN categories, and a species account for each listed animal. Those
accounts name the localities where each species is known in Kosovo and carry
distribution maps.

## Access

The book was found, not at the address the desk review expected, but in
KEPA's listing of reports and publications. The server's certificate had
expired, so the first request below fails; the second repeats it without
certificate verification, which is how the file was read.

```{r}
#| label: access
#| echo: true
#| code-fold: show
access_table(list(
  try_access("https://www.ammk-rks.net/al/drejtorite/38/raporte-studime-publikime"),
  try_access(url),
  try_access(url, ssl_verify = FALSE)
))
```

The colophon reserves all rights: reproduction and republication without
the Ministry's permission are prohibited. The PDF is therefore not copied
into this repository, and nothing is extracted from it here beyond what this
page describes.

## Kosovo coverage check

The book is national by definition. The check that matters is whether its
locality data exist as data. They do not in public: localities appear as
place names in the text and as printed maps, and no table of records or
coordinates is released. The desk review's note that "locality data are not
published" holds, with that nuance.

::: {.callout-warning}
## To be completed
Count the species per national category and group from the PDF, and match
the names to the GBIF backbone, to see how many red-book animals the GBIF
extract already records in Kosovo. Keep the result as counts and names, not
as a copy of the book's content.
:::

## What it adds for Natura 2000 in Kosovo

A national red list says which animals are at risk in Kosovo itself, which
neither the EU annexes nor the global IUCN list (both already in the report)
can. Species that are nationally threatened and also on Annex II are obvious
priorities for site selection; species threatened nationally but absent from
the annexes point to where national protection must do what Natura 2000
will not. It serves the [national tier](../natura2000.qmd#q1).

## Caveats and sensitivity

- All rights reserved; the book can be cited and read, not republished.
- Maps and named localities of threatened species are sensitive. Any use in
  this site must stay at 10 km or coarser for persecuted or collected species
  and state what was generalised.
- Categories follow IUCN regional guidelines as applied by the authors; they
  are not comparable one-to-one with global categories.

## Integration into the pipeline

Medium effort. A proposed `data/kosovo_red_list_fauna.csv`, holding species,
national category and GBIF key only, built by an `R/build_red_list.R` like the
existing `R/build_directive_list.R`, would let `pipeline.R` add a
`nationalRedList` column to every record. That needs either the Ministry's
agreement or a table it publishes. `pipeline.R` is not changed here.

## To be completed

```{r}
#| label: todo
#| output: asis
todo_list(row)
```

## Citation

As the title page gives it: Ibrahimi, H., Gashi, A., Rexhepaj, D., Zhushi
Etemi, F., Grapci Kotori, L., Fehér, Z., Bino, T., Šerić Jelaska, L.,
Mesaroš, G. & Théou, P. (2018). *Libri i Kuq i Faunës së Republikës së
Kosovës.* Ministria e Mjedisit dhe Planifikimit Hapësinor, Prishtinë.
