---
title: "Red Book of Vascular Flora of the Republic of Kosovo"
subtitle: "The 2013 national red book of plants: 237 taxa assessed against IUCN criteria, 61 of them Critically Endangered."
actions:
  - kind: request
    what: "The full species list with national categories as a table (the English summary has none), and the field coordinates recorded for each assessed taxon, at no finer than 10 km for collected species"
    contact: "Ministry of Environment, Spatial Planning and Infrastructure; the book's editors (the summary says the herbarium samples are kept at the University of Prishtina, Faculty of Natural and Mathematical Sciences)"
  - kind: check
    what: "Find the full Albanian edition online, parse the species list and match it to the GBIF backbone with rgbif"
    contact: "KEPA publications page, https://www.ammk-rks.net/al/drejtorite/38/raporte-studime-publikime"
  - kind: check
    what: "Whether the 'Atlas of the Plants of Kosovo' listed on KEPA's publications page carries distribution data for the same taxa"
    contact: "KEPA publications page"
todo:
  doi: "No DOI or ISBN seen in the summary; ask the Ministry"
  licence: "No terms stated in the summary; ask the Ministry before redistributing any part"
  temporal_coverage: "Years of the fieldwork behind the assessments; not stated in the summary"
---

```{r}
#| label: setup
source("../R/sources_helpers.R")
row <- source_row("red-book-flora")
url <- "https://www.ammk-rks.net/assets/cms/uploads/files/Publikime-raporte/The_Red_Book_of_Vascular_Flora_of_Republic_of_Kosovo-_English_(Summary).pdf"
```

::: {.source-status}
`r status_line(row)`
:::

## At a glance

```{r}
#| label: glance
metadata_table(row)
```

## What it is

*The Red Book of Vascular Flora of the Republic of Kosovo* was prepared by a
team led by Fadil Millaku (editor-in-chief) with Ferat Rexhepi, Elez
Krasniqi, Qazim Pajazitaj, Xhavit Mala and Naim Berisha, published by the
Ministry of Environment and Spatial Planning in 2013 with support from German
development cooperation (GIZ). Its English summary states that 237 taxa were
assessed against the IUCN criteria: 1 Extinct, 1 Extinct in the Wild, 61
Critically Endangered, 86 Endangered, 19 Vulnerable, 34 Near Threatened and
35 Least Concern. The selection drew on the literature, on Administrative
Instruction No. 18/2012 on protected species, and on the state of each
species in the field.

The methods are worth noting for what they imply about data. For every
species assessed, the team took herbarium samples, photographs and GPS
coordinates, counted mature individuals where possible, and measured area of
occupancy and extent of occurrence on the UTM grid. The locality data
therefore exist, but they are not in the summary.

## Access

The summary sits on KEPA's server, whose TLS certificate had expired; the
second request repeats the first without verifying it.

```{r}
#| label: access
#| echo: true
#| code-fold: show
access_table(list(try_access(url), try_access(url, ssl_verify = FALSE)))
```

The 17-page English summary contains the preface, the methodology, the
category counts and the IUCN criteria. It does not contain the species list
or any species account, so the desk review's suggested step, parsing the list and
matching it to the GBIF backbone, is not feasible from this file.

## Kosovo coverage check

::: {.callout-warning}
## To be completed
The red book covers Kosovo alone, so coverage is not in question. What cannot
yet be checked is overlap with the GBIF extract: which of the 237 taxa GBIF
records in Kosovo, and where. That needs the species list from the full book.
Once it is available, match the names with
`rgbif::name_backbone_checklist()` and count records per taxon in
`data_exports/kosovo_overall_biodiversity.csv`, keeping any locality output at
10 km for collected species.
:::

## What it adds for Natura 2000 in Kosovo

The report already flags species on the EU annexes and on the global Red
List. Neither captures what is threatened in Kosovo itself. A national list of
148 Critically Endangered or Endangered plants would let the report mark
records of nationally threatened species, and let site selection favour
candidate sites that hold them. The field coordinates behind the assessments
would, if shared, be among the most precise plant data for the country. The
list serves the [national tier](../natura2000.qmd#q1).

## Caveats and sensitivity

- The summary alone gives counts, not species.
- National categories apply regional IUCN guidelines and cannot be compared
  one-to-one with global ones.
- Several of the Critically Endangered taxa are small populations of
  endemics; their coordinates must never be published precisely.

## Integration into the pipeline

Medium effort once a table exists: a proposed `R/build_red_list.R`, modelled
on `R/build_directive_list.R`, would resolve the names and write
`data/kosovo_red_list_flora.csv`; `pipeline.R` would then add a national
category column to every plant record. `pipeline.R` is not changed here.

## To be completed

```{r}
#| label: todo
#| output: asis
todo_list(row)
```

## Citation

From the title page: Millaku, F. (ed.), Rexhepi, F., Krasniqi, E., Pajazitaj,
Q., Mala, X. & Berisha, N. (2013). *The Red Book of Vascular Flora of the
Republic of Kosovo.* Ministry of Environment and Spatial Planning, Prishtinë.
