---
title: "River basin management work programme 2025–2027"
actions:
  - kind: check
    what: "Find the current location of 'Programi i Punës dhe Kalendari i Aktiviteteve (2025–2027)'; the starting URL returned HTTP 404 on 29 September 2026 although search engines still index it"
    contact: "Ministry of Environment, Spatial Planning and Infrastructure, https://mmphi.rks-gov.net/ (water department pages)"
  - kind: check
    what: "Whether the river basin management plans it prepares (for the 2028–2033 period, according to a search-engine summary that was not verified) will include protected-area registers under the Water Framework Directive"
    contact: "MMPHI water department"
todo:
  custodian: "The ministry's water department, if the document is found at MMPHI"
  url: "The document's current address; the starting URL is dead"
  doi: "No DOI expected"
  licence: "Terms of the document"
  access: "Not established while the file is missing"
  format: "PDF, per the dead link's file name"
  access_method: "Unknown until found"
  spatial_resolution: "Not spatial: a work programme"
  temporal_coverage: "2025–2027 per its title (not verified)"
  latest_version: "Unknown until found"
  kosovo_coverage: "National by its title; not verified"
  kosovo_codes_seen: "Not applicable"
  tier: "Not applicable to a work programme; relevant for the river basin plans it prepares"
  n2k_question: "When river basin plans will map water-dependent protected areas, which Natura 2000 freshwater sites would join"
  effort: "Low: a document to cite"
  sensitivity: "None"
---

```{r}
#| label: setup
source("../R/sources_helpers.R")
row <- source_row("river-basin-programme")
dead <- "https://mmphi.rks-gov.net/MMPHIFolder/OtherDocuments/1.%20Programi%20i%20Punes%20per%20PMPL.pdf"
```

::: {.source-status}
`r status_line(row)`
:::

## At a glance

```{r}
#| label: glance
metadata_table(row)
```

## What it is

::: {.callout-warning}
## To be completed
The document could not be opened, so nothing about its content is verified.
The desk review describes it as the ministry's work programme for river basin
management, 2025–2027. A search engine's summary of the dead link describes
a work programme and timetable for preparing river basin management plans
for the "White Drin, Ibër, Morava, Binça and Lepen" rivers for 2028–2033;
that summary is secondary and unverified.
:::

River basin management plans matter for Natura 2000 because the Water
Framework Directive requires a register of protected areas that includes
water-dependent Natura 2000 sites, and because the plans' programmes of
measures decide what happens to the rivers that freshwater habitats and fish
depend on.

## Access

```{r}
#| label: access
#| echo: true
#| code-fold: show
access_table(list(
  try_access(dead),
  try_access("https://mmphi.rks-gov.net/")
))
```

The starting URL returned HTTP 404 with both capitalisations of its folder
name; the ministry's home page answers. The document is listed as moved or
removed; its new location is an open action.

A dead link is a small thing with a large consequence for evidence. When a
ministry's document moves without a redirect, every report that cited its
address stops being checkable, and a search engine that still indexes the old
address makes the document look available when it is not. That is why this
page records the attempt and its date, and why the section as a whole keeps
an evidence row for every value: a reader can see not only what was found but
what could no longer be found.

## Kosovo coverage check

::: {.callout-warning}
## To be completed
Not applicable until the document is found. Once found: record which river
basins it covers, whether it plans a protected-area register, and the
timetable for the first plans.
:::

## What it adds for Natura 2000 in Kosovo

Timing and coordination. If the first river basin plans are being prepared
now, the Natura 2000 work on freshwater habitats (3260, 91E0, 92A0) and
Annex II fish should feed their protected-area register, rather than follow
it by six years.

## Caveats and sensitivity

- Lead, not verified: the link is dead and the content unknown.

## Integration into the pipeline

None; a document to cite. `pipeline.R` is not changed.

## To be completed

```{r}
#| label: todo
#| output: asis
todo_list(row)
```

## Citation

::: {.callout-warning}
## To be completed
Cite once the document is found.
:::
