---
title: "EU for Natura 2000 in Serbia"
subtitle: "An EU-funded project that identified Serbia's first list of potential Natura 2000 sites; the sites along Kosovo's border are the transboundary half of any Kosovo proposal."
actions:
  - kind: request
    what: "Draft boundaries of the 277 pSCIs and 85 SPAs, at least those within 20 km of Kosovo, with their Standard Data Forms"
    contact: "Ministry of Environmental Protection of Serbia, through the Natura 2000 portal https://natura2000.gov.rs/en/; or the Institute for Nature Conservation of Serbia"
  - kind: check
    what: "Whether the Serbian list places any pSCI or SPA inside Kosovo, as Serbia's candidate Emerald sites do (see the Emerald Network page)"
    contact: "The site list, once obtained"
  - kind: check
    what: "Whether natura2000.gov.rs publishes a map service behind its viewer page; the root page returned HTTP 500 on 29 September 2026"
    contact: "https://natura2000.gov.rs/en/natura-2000-viewer/"
todo:
  doi: "No DOI; the results are announced on web pages, not published as a dataset"
  licence: "No terms stated for the site list; ask the Ministry when requesting the boundaries"
  access_method: "Unknown until the Ministry says how boundaries would be supplied"
  spatial_resolution: "Scale of the draft boundaries; state it once they are obtained"
  temporal_coverage: "Fieldwork years; the project summary PDF on eptisasee.com was not opened in this check"
  kosovo_coverage: "Unknown: the pages opened do not say whether any proposed site lies in Kosovo"
  kosovo_codes_seen: "Record which code the Serbian list uses for sites in or near Kosovo once the list is obtained"
  tier: "Depends on the boundary scale; a pSCI boundary would normally serve the site tier"
  effort: "Estimate once the delivery format is known"
  sensitivity: "Ask whether the SDFs carry species locations that Serbia treats as sensitive"
---

```{r}
#| label: setup
source("../R/sources_helpers.R")
row <- source_row("serbia-eu-for-natura2000")
```

::: {.source-status}
`r status_line(row)`
:::

## At a glance

```{r}
#| label: glance
metadata_table(row)
```

## What it is

"EU for Natura 2000 in Serbia" was an EU-funded technical-assistance project
that ended on 30 November 2021. Serbia's Natura 2000 portal, run for the
Ministry of Environmental Protection, reports that in about two and a half
years a team of national and international experts identified 277 potential
Sites of Community Interest (pSCIs) and 85 Special Protection Areas (SPAs)
under the Habitats and Birds Directives. The Ministry, the Institute for Nature
Conservation of Serbia, the Institute for Nature Conservation of Vojvodina
Province and the Serbian Environmental Protection Agency took part.

The project's own site describes its purpose as establishing the first list of
potential Natura 2000 sites, "together with the development of an information
system, database and GIS". The list itself, its boundaries and its forms are
not published on either site.

## Access

```{r}
#| label: access
#| echo: true
#| code-fold: show
access_table(list(
  try_access("https://natura2000.gov.rs/"),
  try_access("https://natura2000.gov.rs/en/natura-2000-in-serbia-achievements-and-next-steps/"),
  try_access("https://natura2000.gov.rs/en/natura-2000-viewer/"),
  try_access("https://natura-2000.euzatebe.rs/en/downloads")
))
```

On 29 September 2026 the portal's root address returned HTTP 500 while its
English pages answered. The project's downloads page offers two newsletters
and a paper of legislative recommendations; no site list, map or GIS file. The
portal's "Natura 2000 viewer" page links to the EEA's viewer of the EU
network, which holds no Serbian sites, since Serbia is not a Member State.
The desk review named `natura2000.gov.rs` without an exact page; the
pages above are the ones found there.

## Kosovo coverage check

::: {.callout-warning}
## To be completed
The draft site boundaries are not published, so no overlay with Kosovo's
border could be run. The pages opened say nothing about sites in or next to
Kosovo. Once the boundaries are obtained: list every pSCI and SPA within 20 km
of the border, measure how much of each lies inside Kosovo, and record the
country code on each site. Serbia's candidate *Emerald* sites already extend
into Kosovo under code RS (see [Emerald Network](emerald-network.qmd)), so the
same must be checked for the Natura 2000 list.
:::

## What it adds for Natura 2000 in Kosovo

Sites on Šar, Prokletije and Kopaonik stop at a border the species do not
recognise. A Kosovo proposal that ends where a Serbian or Macedonian one
begins, with the same habitat on both sides, is easier to defend in the
biogeographical process than one drawn without looking across. The Serbian
list would also show which features a neighbour with a comparable fauna and
flora judged to need sites, which is a check on Kosovo's own
[sufficiency ledger](../natura2000.qmd#q4). The tier depends on the boundary
scale, which is not yet known.

## Caveats and sensitivity

- A potential site list is a technical proposal, not a designation, and may
  change before Serbia submits it.
- Only the project's summary numbers were verified; nothing about individual
  sites was.
- Any site list that includes Kosovo's territory under Serbian codes needs to
  be read as a data question, not a statement on status.

## Integration into the pipeline

Low effort once the boundaries exist as a GeoPackage or shapefile: a proposed
`neighbour_candidate_sites(boundary, path, buffer_km = 20)` in
`R/functions.R`, read beside `kosovo_protected_areas()`, would give the report
a transboundary layer. Until then there is nothing to integrate. `pipeline.R`
is not changed.

## To be completed

```{r}
#| label: todo
#| output: asis
todo_list(row)
```

## Citation

::: {.callout-warning}
## To be completed
No citation guidance was found on either site. Cite the Ministry's list in the
form it gives when the boundaries are supplied.
:::
