---
title: "Soil and geology layers (SoilGrids and others)"
subtitle: "Soil properties from ISRIC's SoilGrids, cut to Kosovo; no open geological map was confirmed."
actions:
  - kind: check
    what: "Find an open geological map covering Kosovo (a European 1:1M or 1:5M compilation, or the national geological map), with its licence"
    contact: "European Geological Data Infrastructure (EGDI); ICMM, whose data shop sells Kosovo maps at 1:50,000–1:200,000"
  - kind: check
    what: "Map serpentine (ultramafic) outcrops, which carry several of Kosovo's endemic plants; neither SoilGrids nor land cover shows them"
    contact: "A geological map, once found"
  - kind: check
    what: "Read SoilGrids' licence and citation on isric.org"
    contact: "https://www.isric.org/explore/soilgrids"
todo:
  custodian: "ISRIC – World Soil Information, per the host name files.isric.org; confirm on isric.org"
  doi: "SoilGrids 2.0 data paper DOI; read on isric.org"
  licence: "SoilGrids licence; read on isric.org (not read in this check)"
  temporal_coverage: "Not applicable: modelled soil properties"
  latest_version: "Version behind the 'latest' folder on files.isric.org"
  effort: "Low for SoilGrids; unknown for geology until a source is found"
  sensitivity: "None"
---

```{r}
#| label: setup
source("../R/sources_helpers.R")
suppressPackageStartupMessages({
  library(sf)
  library(terra)
  library(leaflet)
})
row <- source_row("soils-geology")
cache <- file.path(cache_dir("soils-geology"), "soilgrids_soc_0-5cm_kosovo.tif")
vrt <- "https://files.isric.org/soilgrids/latest/data/soc/soc_0-5cm_mean.vrt"
```

::: {.source-status}
`r status_line(row)`
:::

## At a glance

```{r}
#| label: glance
metadata_table(row)
```

## What it is

The desk review grouped soils and geology under one lead. The soil half has
an obvious open source: SoilGrids, ISRIC's global maps of soil properties at
250 m, served as cloud-optimised GeoTIFFs behind VRT files at
`files.isric.org`. The layer used here is soil organic carbon in the top 5 cm.
The geology half has none confirmed. Geology matters in Kosovo more than
elsewhere, because several endemic plants grow only on serpentine, limestone
or silicate (see [endemic plants](endemic-plants-hacquetia.qmd)).

## Access

```{r}
#| label: access
#| echo: true
#| code-fold: show
access_table(list(try_access("https://files.isric.org/soilgrids/latest/data/"),
                  try_access(vrt)))
```

```{r}
#| label: build
#| echo: true
#| code-fold: true
#| code-summary: "Show the remote crop"
build_error <- NULL
if (!file.exists(cache)) {
  build_error <- tryCatch({
    Sys.setenv(GDAL_DISABLE_READDIR_ON_OPEN = "EMPTY_DIR")
    r  <- rast(paste0("/vsicurl/", vrt))          # Interrupted Goode Homolosine
    ko <- project(vect(kosovo_outline()), crs(r))
    k  <- mask(crop(r, ko, snap = "out"), ko)
    writeRaster(k, cache, overwrite = TRUE, gdal = "COMPRESS=DEFLATE")
    NULL
  }, error = function(e) conditionMessage(e))
}
have_cache <- file.exists(cache)
```

```{r}
#| label: build-failed
#| eval: !expr "!have_cache"
#| output: asis
cat("::: {.callout-warning}\n## To be completed\nThe remote crop failed at render time (",
    build_error %||% "no message", "). Re-render when files.isric.org answers.\n:::\n", sep = "")
```

## Kosovo coverage check

```{r}
#| label: check
#| eval: !expr have_cache
#| output: asis
k <- rast(cache)
inside <- !is.na(values(rasterize(project(vect(kosovo_outline()), crs(k)), k), mat = FALSE))
v <- values(k, mat = FALSE)[inside]
cat(sprintf(paste(
  "Inside Kosovo: %s cells of 250 m, %d of them no-data (%.1f per cent). Soil",
  "organic carbon in the top 5 cm ranges from %s to %s in the layer's stored",
  "units (dg/kg per the SoilGrids convention, to be confirmed on isric.org).",
  "No country attribute, so no country code applies."),
  fmt_int(length(v)), sum(is.na(v)), 100 * mean(is.na(v)),
  fmt_int(min(v, na.rm = TRUE)), fmt_int(max(v, na.rm = TRUE))))
```

```{r}
#| label: map
#| eval: !expr have_cache
source("../R/functions.R", local = TRUE)
k4 <- project(k, "EPSG:3857")
pal <- colorNumeric(grDevices::hcl.colors(9, "YlOrBr", rev = TRUE), values(k4), na.color = NA)
leaflet() |>
  add_basemaps() |>
  addRasterImage(k4, colors = pal, opacity = 0.75, project = FALSE) |>
  add_boundary_outline(kosovo_outline()) |>
  addLegend(pal = pal, values = values(k4), title = "SOC 0–5 cm<br>(stored units)") |>
  addLayersControl(baseGroups = basemap_groups,
                   options = layersControlOptions(collapsed = TRUE))
```

SoilGrids is stored in the Interrupted Goode Homolosine projection, an
equal-area projection for the whole globe, unlike every other layer in this
section. The build step therefore projects the Kosovo outline into the
raster's projection and crops there, rather than reprojecting the raster to
fit the outline: reprojecting a raster resamples every cell, and doing it
before cropping would mean resampling far more of the globe than is needed.
Only the map on this page reprojects, to Web Mercator for display.

::: {.callout-warning}
## To be completed
Geology: no open layer was checked. Find one, crop it to Kosovo, and report
the area of serpentine and limestone, which are the substrates Kosovo's
endemic plants depend on.
:::

## What it adds for Natura 2000 in Kosovo

Soil and substrate are covariates for habitat models, and substrate in
particular separates Annex I types that land cover cannot: calcareous from
siliceous grassland and rock, serpentine grassland from both. SoilGrids gives
consistent soil properties; a geological map would give the substrate.
National screening only ([tier](../natura2000.qmd#q1)). Of the two halves,
the missing geology matters more for Kosovo: a serpentine outcrop is a
candidate site for its endemic flora whatever its soil carbon, and no layer
checked so far can show one.

## Caveats and sensitivity

- SoilGrids is a global model; local accuracy in Kosovo's mountains is
  unknown.
- The geology half of this page is unverified.
- No sensitive content.

## Integration into the pipeline

Low effort for SoilGrids: a proposed `kosovo_soil(boundary, property = "soc")`
as above. `pipeline.R` is not changed.

## To be completed

```{r}
#| label: todo
#| output: asis
todo_list(row)
```

## Citation

::: {.callout-warning}
## To be completed
Use the citation ISRIC requests for SoilGrids, not read in this check.
:::
