---
title: "WorldClim"
subtitle: "Global climate layers, reached for Kosovo in one line of R with the geodata package, which files Kosovo under XKO."
actions:
  - kind: check
    what: "Read WorldClim's licence and requested citation on worldclim.org, and record the custodian"
    contact: "https://www.worldclim.org/"
todo:
  custodian: "Maintainers of WorldClim; read on worldclim.org"
  doi: "WorldClim 2.1 data paper DOI; read on worldclim.org"
  licence: "WorldClim's terms of use; read on worldclim.org"
  temporal_coverage: "Reference period of the WorldClim 2.1 climatology; read on worldclim.org"
---

```{r}
#| label: setup
source("../R/sources_helpers.R")
suppressPackageStartupMessages({
  library(sf)
  library(terra)
  library(leaflet)
})
row <- source_row("worldclim")
dir <- cache_dir("worldclim")
cache <- file.path(dir, "climate", "wc2.1_country", "XKO_wc2.1_30s_tavg.tif")
```

::: {.source-status}
`r status_line(row)`
:::

## At a glance

```{r}
#| label: glance
metadata_table(row)
```

## What it is

WorldClim provides global gridded climate, such as monthly mean
temperature, at resolutions down to 30 arc-seconds (about 1 km). This page uses it mainly to show how the R package
`geodata` reaches such layers: `geodata` serves per-country tiles of
WorldClim 2.1 from `geodata.ucdavis.edu`, and it identifies Kosovo by the
ISO3-style code **XKO** (ISO2 **XK**), the same code as GADM, which the GBIF
download also filters on.

## Access

```{r}
#| label: access
#| echo: true
#| code-fold: show
access_table(list(
  try_access("https://www.worldclim.org/"),
  try_access("https://geodata.ucdavis.edu/climate/worldclim/2_1/tiles/iso/XKO_wc2.1_30s_tavg.tif")
))
```

```{r}
#| label: build
#| echo: true
#| code-fold: show
# One line of R: monthly mean temperature for Kosovo at 30 arc-seconds.
build_error <- NULL
if (!file.exists(cache)) {
  build_error <- tryCatch({
    geodata::worldclim_country("XKO", var = "tavg", res = 0.5, path = dir)
    NULL
  }, error = function(e) conditionMessage(e))
}
have_cache <- file.exists(cache)
```

```{r}
#| label: build-failed
#| eval: !expr "!have_cache"
#| output: asis
cat("::: {.callout-warning}\n## To be completed\nThe geodata download failed at render time (",
    build_error %||% "no message", "). Re-render when geodata.ucdavis.edu answers.\n:::\n", sep = "")
```

## Kosovo coverage check

```{r}
#| label: codes
cc <- geodata::country_codes()
knitr::kable(cc[cc$NAME == "Kosovo", c("NAME", "ISO3", "ISO2", "NAME_LOCAL")],
             row.names = FALSE, caption = "How geodata names Kosovo")
```

```{r}
#| label: check
#| eval: !expr have_cache
#| output: asis
w <- rast(cache)
ko <- vect(kosovo_outline())
annual <- mask(crop(mean(w), ko, snap = "out"), ko)
inside <- !is.na(values(rasterize(ko, annual), mat = FALSE))
v <- values(annual, mat = FALSE)[inside]
cat(sprintf(paste0("The country tile covers %.1f–%.1f° E and %.1f–%.1f° N, ",
                   "wider than Kosovo. Inside the outline: %s cells, %d no-data; ",
                   "annual mean of the monthly means %.1f to %.1f °C (mean %.1f °C)."),
            xmin(w), xmax(w), ymin(w), ymax(w), fmt_int(length(v)), sum(is.na(v)),
            min(v, na.rm = TRUE), max(v, na.rm = TRUE), mean(v, na.rm = TRUE)))
```

The tile is cut to a rectangle around Kosovo, not to its border: `geodata`
leaves the clipping to the user, which is why the outline from the pipeline
is applied here.

The code matters more than it looks. `geodata` finds Kosovo because it uses
GADM's list of countries, where Kosovo is XKO. Both wrong guesses were
tried while writing this page. Asking for "XKX", the code the pipeline stamps
on its own records, stops with "unknown country: XKX". Asking for "SRB", in
the hope of catching Kosovo inside Serbia, returns Serbia's tile, a rectangle
from 42° to 46.5° N that takes in most of Kosovo by accident and stops short
of its southern mountains. The same lookup, `geodata::country_codes()`, is worth running before
any per-country download from this package, and its answer recorded, as the
table above does. `geodata` also caches what it downloads under the `path`
it is given, so a second render reads the local file rather than the server;
here that cache sits under `data/sources/worldclim/`.

```{r}
#| label: map
#| eval: !expr have_cache
source("../R/functions.R", local = TRUE)
pal <- colorNumeric(grDevices::hcl.colors(9, "RdYlBu", rev = TRUE),
                    values(annual), na.color = NA)
leaflet() |>
  add_basemaps() |>
  addRasterImage(annual, colors = pal, opacity = 0.75) |>
  add_boundary_outline(kosovo_outline()) |>
  addLegend(pal = pal, values = values(annual), title = "Mean temperature (°C)") |>
  addLayersControl(baseGroups = basemap_groups,
                   options = layersControlOptions(collapsed = TRUE))
```

## What it adds for Natura 2000 in Kosovo

The same kind of covariates as [CHELSA](chelsa.qmd), with the advantage that
`geodata` makes them one line of code, which suits teaching.
Two climatologies from different methods also give a cheap check on each
other in Kosovo's mountains. National screening only
([tier](../natura2000.qmd#q1)).

## Caveats and sensitivity

- Interpolated climate; mountain gradients are smoothed at 1 km.
- The country tile is a rectangle; clip it before computing national figures.
- No sensitive content.

## Integration into the pipeline

Low effort; the package does the work. A proposed
`kosovo_climate(boundary, source = "worldclim")` beside the CHELSA variant.
`pipeline.R` is not changed.

## To be completed

```{r}
#| label: todo
#| output: asis
todo_list(row)
```

## Citation

::: {.callout-warning}
## To be completed
Use the citation WorldClim requests on worldclim.org, and cite `geodata`
with `citation("geodata")`.
:::
